# Mangrove forest ecology

## Zonation & disturbance {#zonation}

### Preface

Establishment of monospecific stands of mangrove trees in zones parallel to shorelines (zonation) has been among the most well-studied phenomenon in mangroves. Several hypotheses attempting to explain zonation in mangroves have existed throughout time, each of which has been supported to varying degrees within the literature:

*Zonation as mangrove succession due to land building - Davis, 1940:*

  - From his work in Florida, Davis believed that mangroves promoted shoreline progradation, which was followed by establishment of *Avicennia germinans* species in the most seaward portion of the intertidal zone. Eventually *Rhizophora mangle* with a few *Laguncularia racemosa* individuals would outcompete *A. germinans* as the shore (and thus habitat for *A. germinans*) continued to prograde. Davis believed that the zones of species were a successional process, eventually climaxing in a terrestrial forest dominated by non-mangrove species. [Not supported]
  
*Zonation as response to geomorphic condition - Thom, 1967:*

  - Thom did his field work in a composite lagoon and riverine system in Tabasco, Mexico, and provided one of the first most comprehensive sets of evidence against Davis's hypothesis of zonation as mangrove succession. In looking at the different habitats in which mangrove tree species existed, Thom concluded that zonation of species was in response to the hydrological and substratum conditions of a site, which were ultimately variables controlled by geomorphic processes. As a result, he concluded that spatial patterns of mangrove species were in response to geomorphic changes in the land rather than mangrove species inducing patterns in geomorphology (and thus species zonation was relatively "stable"). [Supported]
  
*Zonation as adaptation to physico-chemical gradients - Watson, 1928; Davis, 1940; Macnae, 1968; Ball 1988:*

  - One of the more commonly examined hypotheses for examining species zonation and investigated by a wide range of researchers. Watson's "inundation classes" and discussion of salinity may be one of the earliest, well-cited references to environmental gradients. The hypothesis holds that mangrove species have adapted to establish or compete in microenvironments that exist along environmental gradients. Given the dynamic nature of mangrove ecosystems, it is unlikely that mangrove species specialize to unique microenvironments, but more likely exhibit tolerance to a wide range of conditions. However, some may preferentially outcompete others in environments towards the ends of gradient spectrums, and thus ecophysiology may play an important role in competition and survival or growth *following* establishment. [Supported, for particular considerations]

*Zonation as result of interspecific competition - Clarke and Hannon, 1970:*

  - 

*Zonation due to tidal sorting of propagules - Rabinowitz, 1978:*

  - Upon investigating no significant difference in survival and growth rates of *Avicennia, Rhizophora, Laguncularia,* and *Pelliciera* tree individuals planted in monospecific stands in Panama, Rabinowitz concluded that mangrove zonation is not primarily controlled by physiological variables. In response, she hypothesized that the zonation of species may be due to differences in propagule size and weight between species. The hypothesis posits that differences in size and weight have enabled species to establish in areas with different tidal regimes (smaller propagules in shoreward zones, larger propagules in landward zones). Rabinowitz provides a discussion of the hypothesis and identifies several other studies that may support the hypothesis. [Partially supported - directionality NOT supported whereas tidal depth supported]

*Zonation as result of differential predation on propagules - Smith, 1987:*

  - [@smith1987seed] was among the first studies to experimentally test the existence of differential rates of propagule predation by grapsid crabs for different mangrove tree species. His results showed that differential rates of mangrove predation did exist for the distinct species, and were correlated to the nutritional content (e.g., presence of tannins, simple sugars, etc.) of the species. Furthermore, he found a significant negative relationship between species dominance and rate of propagule predation, which led him to conclude that differential rates of propagule predation influence the zonation of species in mangroves. Perhaps most clearly, he found that *A. marina* was consumed heavily within the mid-intertidal region of the forest, in which no parent trees of *A. marina* exist. Smith notes that no single hypothesis is likely to explain all of zonation, but that the results of his study are significant and parallel those of other studies in terrestrial forests. [Partially supported]


Several comprehensive reviews of zonation in mangroves are provided in:

  - Lugo and Snedaker, 1974 (Reviewed in Section \@ref(lugosnedaker1974))
  - Smith, 1992 (Reviewed in Section \@ref(smith1992))

### Thom, 1967 {#thom1967}

*"Mangrove ecology and deltaic geomorphology: Tabasco, Mexico"* [@thom1967mangrove]

**Key contribution:** Thom's work in the lagoons and riverine settings ("composite river & wave-dominated") of Tabasco, Mexico was one of the most well-cited examples against Thom's hypothesis of mangrove succession. In looking at the different hydrological regimes and substratum of various mangrove "zones," he concluded that mangrove zonation in species was in response to geomorphic condition and was relatively stable in time (as opposed to altering landscapes and facilitating succession).

Thom generally concluded that mangroves were *opportunistic* in their establishment, but particular species do have the capabilities to affect the *rates* at which geomorphic processes occur such as prop roots of *Rhizophora* spps. choking tidal channels.

The following chart details the presence of different stand types within the geomorphic settings of Tabasco, Mexico.

```{r, echo=FALSE}
knitr::include_graphics(rep("images/thom1967.bmp"))
```

**Key findings:** Thom's findings provided evidence that:

  - Mangroves establish in areas of naturally accreting soil, but facilitate accretion of soil upon establishment (e.g., Rhizophora choking creeks & channels).
  - Mangroves are opportunistic in their establishment, largely responding to geomorphic conditions
  - Thom went on to provide a series of geomorphic clastic and carbonate settings in which mangroves "generally" exist

**Further reading:**

  1. Thom 1982 - Identification of five geomorphic clastic settings
  2. Thom 1984 - Identification of three geomorphic carbonate settings

### Lugo & Snedaker, 1974 {#lugosnedaker1974}

*"The ecology of mangroves"* [@lugo1974ecology]

**Key contribution:** Lugo and Snedaker's seminal work on the ecology of mangroves (in Florida) is most commonly cited for their definition of six mangrove forest types:

  - **Fringe forest** - Mangrove assemblages that are found on the fringes of sheltered coastlines. Zonation in species typically follows Davis's patterns.

  - **Riverine forest** - Tall forests found in floodplains and along riverine systems of deltas and estuaries. These may occur in conjunction with "fringe forests," are tidally flushed, yet have more complex hydrological influences from upland, freshwater sources.

  - **Overwash forest** - Forest types often consisting of *Rhizophora* species over low islands or small projections in bays and estuaries that are only inundated with high tides.

  - **Basin forest** - Mangrove assemblages that occur in the "basins" in between tidal channels and distributaries, or behind the levees of rivers and channels. Often among the most extensive sections of mangrove forests and may exhibit zonation with changes in elevation as the landward zone of the forest is reached.

  - **Hammock forest** - A variant of the basin forest occuring in fringe forests in areas of raised topography.

  - **Dwarf forest** - Mangrove forests that occur along flat coastal fringes and are stunted in growth (height) due to nutrient limitations.

In addition, Lugo and Snedaker provide a review of matter cycling in mangroves; however, more recent reviews of the topic have been written and should be referred to.

Lugo and Snedaker provide a conceptual model of energy flows in mangroves that should be reviewed:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/lugo1974.bmp"))
```

### Rabinowitz, 1978a

*"Dispersal properties of mangrove propagules"* [@rabinowitz1978dispersal]

**Key contribution:** This paper provides a review of propagule physiology and establishment for six species from Panama (*L. racemosa, A. germinans, A. bicolor, R. mangle, R. harisonii*, and *P. rhizophorae*). Rabinowitz reviews the different anatomy and physiology of the four genera, and then provides results from experiments on dispersal characteristics such as time of floatation, time to rotting and time to establishment.

**Key knowledge:** Mangroves are viviparous in that propagules go from parent tree to embryonic stage to establishment continuously without a "dormant" period. Propagules often sprout roots during the dispersal stage, which aids in their establishment.

**Key finding:** After her experiments, Rabinowitz found that smaller propagules from *Avicennia* and *Laguncularia* were less capable of sinking or establishing underwater, and thus are more likely to be limited to higher areas with periodic absences of tidal inundation (where they are characteristically found). The larger propagules of *Rhizophora* and *Pelliciera* are capable of sinking and establishing underwater, and thus are not limited in areas of establishment.

### Rabinowitz, 1978b

*"Early growth of mangrove seedlings in Panama, and an hypothesis concerning the relationship of dispersal and zonation"* [@rabinowitz1978early]

**Study design:** Examined growth and survival of *R. harisonii. A. sps, L. racemosa,* and *P. rhizophorae* in each of four monospecific parent swamps in different locations of Panama (i.e., four parent stands at each of four sites). Gardens of each of the four species were planted and examined for growth and survival to see if seedlings fared better in stands of their parent species.

**Key findings:** Rabinowitz examined establishment of four different genera (*Avicennia, Laguncularia, Pelliciera,* and *Rhizphora*) in stands of both their parent genus as well as others. Her findings were primarily:

  1. seedlings generally grew better in habitats of other mangroves than in own parents habitat
  2. growth of seedling largely equal across parents swamps
  3. superior growth (when present) typically seen in foreign swamp
  4. no habitat with greater growth across all mangroves
  
Thus, Rabinowitz found little evidence of the hypothesis that physiological conditions along a tidal gradient control mangrove zonation. The graph below shows survival of seedings of the four genera in the four different parent habitats.

```{r echo=FALSE}
knitr::include_graphics(rep("images/rabinowitz1978a.bmp"))
```

**Key contribution:** In response to her findings, Rabinowitz provides the "tidal sorting" hypothesis, which posits that "mangrove zonation may be controlled by tidal sorting of propagules according to size and by differential ability of propagules to establish in deep water."

She provides a discussion of the hypothesis as well as links it to several studies in the academic literature which support the notion.

**Key limitations:** The key limitations of the study is the length of the work (just one year), as well as consideration of only seedlings rather than saplings and seedlings. Ecophysiological controls on survival rates may not occur until later in time (in which competitive influences are stronger), and thus the conclusion that Rabinowitz gives may have been premature (see [@mckee1995seedling]).

**Further reading:**

  1. Rabinowitz, 1978 - "Dispersal properties of mangrove propagules"
  2. Sousa et al, 2007
  
### Ball, 1980 {#ball1980}

*"Patterns of secondary succession in a mangrove forests of Southern Florida"* [@ball1980patterns]

**Key contribution:** This study examines secondary succession in Biscayne Bay, FL and attributes differences in species composition within the two dominant forest stands (below and above mean tidal reach) as a result of competition between *R. mangle* and *L. racemosa*.

**Study design:** The authors coupled study of aerial photography of the region with field based surveys of the relative importance of seedlings, saplings, and live and dead trees within:

  1. A "historical"" forest stand
  2. Three induced forest stands
  
**Key results:** Their result found that *R. mangle* was dominant in stands below mean tidal range, whereas *L. racemosa* was dominant (though sparse and stunted) in areas upland of the mean tidal range. Ball concludes that the spatial patterning is a result of competitive dominance of *R. mangle* in the low intertidal, given that seedlings and saplings of both *R. mangle* and *L. racemosa* are both present in the stand.

Ball hypothesizes that both species likely established at an early time, and *L. racemosa* was able to establish due to shade intolerance and faster growth rates. As presence of *R. mangle* has increased and overtopped *L. racemosa*, a shift in community composition is occurring.
  
### Putz & Chan, 1986

*"Tree growth dynamics and productivity in a mature mangrove forest in Malaysia"* [@putz1986tree]

**Key contribution:** This study describes mangrove development and growth dynamics in a protected forest in Matang, Malaysia over the course of 60 years. The research represents one of the few studies done on mangroves over many decades, as disturbance regimes often prevent long-term succession as well as a lack of research interest.

**Key findings:** Several key findings in long-term mangrove development were identified:

  - *R. apiculata* slowly decreased in forest dominance relative to *B. gymnorrhiza* and *B. parviflora*. *Bruguiera* species are more shade-tolerant and may represent a more traditionally "climax" class of mangrove species.
  - Gap openings were largely caused by termites and fungus killing trees, which would fall against neighboring trees, damage roots and bark and allow termites to transfer trees. Putz and Chan describe the disturbance dynamics as "domino-like," but also note it is unlikley to be "normal" disturbance dynamics.
  - Based on growth-rates and age-class distributions, Putz and Chan conclude that self-thinning did occur within the forest, particularly for *R. apiculata*.
  - Total biomass was relatively high, with "stable" biomass estimated at 350-400 Mg/Ha.

### Smith, 1992 {#smith1992}

*"Forest structure" (In Robertson and Alongi, 2002)* [@smith1992forest]

**Overview:** This chapter (from Robertson and Alongi, 2002) gives a broad overview of mangrove forest structure with a particular emphasis on zonation in species. Smith describes factors that influence species composition, the patterns of species assemblages across the intertidal zone, and physical forest structure attributes such as height or stem density.

**Key knowledge:**

  - Mangroves lack understory vegetation commonly found within terrestrial tropical forests
  - Species zonation patterns differ between "Old World" and "New World" tropics
  - @ix primary zonation hypotheses:
    1. land building & succession
    2. geomorphology
    3. response to physico-chemical gradients
    4. tidal sorting hypothesis
    5. differential predation
    6. interspecific competition
  - Many mangroves have even-aged size-class structure
  - Stand height, density and biomass accumulation seem to correlate with climatic factors, particularly rainfall
  - Mangroves appear to have more pioneer-phase traits (e.g., r-strategy dispersal of propagules) relative to mature-phase traits - which may support Thom, 1967's hypothesis of "opportunistic" establishment.

**Status of zonation hypotheses as of 1992:**

```{r echo=FALSE}
knitr::include_graphics(rep("images/smith1992tab3.bmp"))
```

**Further reading:**

  1. Chapman 1976 - Extensive review of early literature surrounding mangrove zonation

### Bunt, 1996 {#bunt1996}

*"Mangrove zonation: An examination of data from seventeen riverine estuaries in tropical Australia"* [@bunt1996mangrove]

**Key significance:** This study compiles species occurrence data from 17 rivers in northern Australia to test hypotheses of zonation. This study is a precursor to Ellison 2000 (see Section \@ref(ellison2000)), and employs an attempt at quantitatively assessing zonation in mangroves.

**Study design:** They use a methodology developed by Williams *et al* 1991 that scores species distributions between 0 and 1, with 0 indicating presence at the water edge and 1 indicating presence in the landward limit of the mangrove. In addition, a standard deviation is calculated which is a measure of how widely it is distributed across the transect with respect to its most probably location.

Transects were laid normal to the water at three different locations along each river (downstream, midstream and upstream) and were extended until the landward limit of the mangrove forest.

**Key results:** The study results show that species-specific patterns of zonation across the various rivers do not exist. Although some zonation trends appear across transects within rivers, the dataset as a whole is highly variable.

In particular, the centers of distribution are not consistent from river to river, and furthermore orderings are not consistent even within transects along a river.

Bunt attributes the variation in distribution of species to abiotic and biotic controls that vary largely at fine scales. For future investigations of spatial distributions of mangroves, they recommend a site-specific approach as regional trends are unlikely to exist.

### Chen & Twilley, 1998 {#chentwilley1998}

*"A gap dynamic model of mangrove forest development along gradients of soil salinity and nutrient resources"* [@chen1998gap]

**Key contribution:** This study employs a FORMAN model to examine forest development and succession in South Florida across nutrient gradients. The study is novel in that it employs an individual-tree based model to examine different effects of recruitment, nutrient availability, and response to disasters (hurricanes) for *R. mangle, A. germinans* and *L. racemosa*.

**Key findings:** The study primarily examines i) species dominance (as measured by BA) in response to salinity and nutrient response surfaces, ii) regeneration following disturbance, and iii) long-term mangrove succession (500 yrs). 

Key findings from each of the three model explorations are detailed below:

  1. *nutrient gradients* - BA didn't increase linearly with increasing soil fertility, indicating that light or competition influence forest development; patterns of forest development varied depending on time scales (35 vs. 100 vs. 300 years). At 100-year time frame, patterns in species partitioning along resource gradients was observed.
  2. *disturbance* - equal sapling recruitment scenarios for all three species overestimated total basal area, which is indicative that recruitment rates are not equal amongst species;
  3. *succession* - long-term succession was found to be sensitive to recruitment rates for each species. The degree of shade tolerance/intolerance of the different species was concluded to be significant in long-term succession dynamics.
  
**Research relevance:** Can the same models be employed to look at shifts in forest structure due to sea-level rise?

### Ellison, 2000 {#ellison2000}

*"Testing patterns of zonation in mangroves: Scale dependence and environmental correlates in the Sundarbans of Bangladesh"* [@ellison2000testing]

**Key contribution:** This is one of the first statistically rigid accounts of testing zonation in mangroves. The authors employ a variety of statistical techniques to test the hypothesis that mangrove species correlate with surface elevation (i.e., tidal zone) and edaphic characteristics in mangroves.

**Study design:** Eleven blocks with three transects 200 m in length (20 adjacent 10 x 10 m plots) running from the shoreward to landward sections of the forest were established in the Sundarbans. Density, frequency and basal area were measured for all species within each plot, as well as edaphic characteristics such as salinity, field capacity (moisture), CEC, soil texture, and nutrient availability.

The data were tested across different scales (within transect, within block, across blocks) to test for mangrove zonation at different scales.

**Key results:** The results did not find any evidence of statistically robust zonation at any of the scales tested. Canonical correspondence analysis did show low amounts of variation in species composition explained by edaphic conditions (24%), but relationships were unclear for the majority of the species that dominate the forest.

The authors conclude that patterns of mangroves are likely more complex than is generally thought in the literature, and may be due to complex interactions between edaphic conditions across space as well as species-specific adaptation to those specific conditions. Additionally, they suggest that interspecific competition or propagule dispersal may be important in determining size and abundance of co-occuring species.

The results of the study match those of others that have attempted to explain zonation patterns through statistical tests (see Section \@ref(bunt1996). 

### Sherman *et al*, 2000 {#sherman2000}

*"Small-scale disturbance and regeneration dynamics in a neotropical mangrove forest"* [@sherman2000small]

**Key contribution:** Sherman *et al* describe chracteristics of gap creation and closure through time and gap regeneration dynamics in the Dominican Republic. The primary intermediate disturbance at their site is lightning-induced gaps, with the forest regenerating from a large tidal wave several decades prior. Their evidence does not support the hypothesis that zonation in mangrove species is reinforced by gap regeneration, with *R. mangle* dominating gap regeneration in all three zone types.

**Key findings:** The following key findings are provided by the study:

  - Hypothesis that species zonation is reinforced by gap dynamics was not supported by evidence
  - *R. mangle* dominated gap regeneration, not due to increased density of seedlings (found to be highly variable with patterns emerging depending on dominant species zone), but due to higher survivorship in standing water; seedling & sapling growth found to be significantly greater in gaps than forest
  - Gaps were characterized by a collapse of the peat-mat and thus regions of standing water
  - Found that seedling and sapling density generally followed mature tree (by basal area) compositions for both transects
  - Current spatial distribution of species zones is still unexplained for the site
  
**NOTE** Need to go back and figure out how long this study was performed for!

### Sousa *et al*, 2007

*"Supply-side ecology in mangroves: Do propagule dispersal and seedling establishment explain forest structure?"* [@sousa2007supply]

**Supply-side ecology:** "hypothesis of community structure being influenced more heavily by supply of new individuals to a site rather than post-recruitment biotic interactions (Roughgarden by Lewin [1986])"

*Tidal sorting hypothesis:* TSH has multiple components:

  1. Dispersal of mangrove propagules is controlled primarily by their shape and size
  2. Establishment of mangrove propagules is differentiated by size and depth of tidal waters
  
**Key contribution:** Sousa *et al* provides one of the few empirical tests of the TSH. Their results by and large do not strictly support all components of the TSH:

  1. Direction (NOT SUPPORTED) - all propagules dispersed seawards, regardless of size
  2. Distance (SUPPORTED) - *Laguncularia* propagules travelled much further than *Rhizophora* or *Avicennia*
    - trapping of seedlings behind logs or other barriers contributes to patchy establishment
  3. Establishment (SUPPORTED) - only *Rhizophora* established in the pen at the lowest tidal elevation, which is indicative that larger heavier propagules are better able to establish in deeper water (as TSH predicts)

**Key findings:**

  1. Directional flow of rainfall runoff overwhelms tidal flows and causes all propagules to disperse in a seawards direction
  2. *Rhizophora* seedlings moved very short distances, indicating that their commonly-believed widespread dispersal via marine systems may be misrepresentative of average dispersal mechanisms, particularly within mature forests.
  3. The TSH does not sufficiently explain zonation at the Punta de Galeta. Sousa *et al* provide two alternative hypotheses for zonation
    - TSH may function strictly during periods of tidal storm surge, when smaller propagules are carried much further inland
    - Zonation may be a result of historical patterns of biogeography as the three species appeared in the region at different times
    
In addition, the authors provide three key criticisms of Rabinowitz's TSH:

  1. Her early studies of species survival and growth in stands of different parent organisms only lasted for a year, and thus physico-chemical factors may not have had time to influence survival given large nutrient reserves. Other studies, however, have since supported Rabinowitz's findings and provide evidence against the physico-chemical hypothesis of zonation
  2. The patterns of zonation predicted by the TSH do not always occur in mangroves; their are examples of forests where the species patterning is different or even opposite from what is predicted.
  3. The pattern of disperal predicted by TSH does not hold in all cases; Sousa *et al* found that dispersal direction was actually opposite of what was predicted by TSH.
  
See also:

  1. Dispersal patterns of Mangrove Propagules [@rabinowitz1978dispersal]
  2. Early growth of mangrove seedlings in Panama... [@rabinowitz1978early]

## Ecophysiology & establishment

### Chapman, 1976, Ch. 8 {#chapman1976ch8}

*"Ch. 8 - Ecological factors"* [@chapman1976ecological]

**Key contribution: ** This is a somewhat dated review of ecological factors that are influential on mangrove vegetation. There are more recent reviews that should be referred to, but this chapter provides a nice overview of some of the earlier work related to mangrove vegetation and ecological factors.

**Key notes: **

*Inundation* - The effects of salinity versus frequency and duration of inundation likely influences seedlings versus mature trees differentially. Inundation "classes" were originally proposed by Watson (and others), but likely have little practical value now. Number of days of *lack of* inundation is understood (by Chapman) to be the most important component of "inundation" more broadly. There is a seasonal component that is important to consider (i.e., areas that have no tidal inundation and no rainfall may be more hypersaline and hostile to establishment).

*Salinity* - Chapman notes that there are correlations between salinity and presence of vegetation, but the relationships may be overstated.

*Soil composition* - The texture of soils is extremely important as it relates to drainage and also aeration of the soil. The presence of sand within the soil can be extremely important; often have complete lack of oxygen within 5 cm soil depth (along channels as well). Presence of peat or other organic layers can greatly influence gas and water infiltration. The dark color of soils is related to iron sulfide production by anaerobic bacteria.

*Water table and drainage* - The mean level of the water table has been found at some sites to increase as you move towards the shoreward section of the forest. There is very little movement with seasonal changes in tide (i.e., spring vs. neap tides), which may be related to the texture of the soil present. Some carbonate settings have extremely coarse substrates (majority of sand > 2 mm).

*non NaCl soil chemistry* - CEC is generally higher in the surface levels of the soil.

*Biota:* Chapman provides 6 habitats in which biota exist, as an extension of MacNae. They most significant habitat for my particular research purposes are the ecosystem engineers, which would consist of the crustaceans that inhabit the canopies, the soil surface, and the soil itself.

**Further reading: ** Need to look up sulfur reduction processes.

### Chapman, 1976, Ch. 9 {#chapman1976ch9}

*"Ch. 9 - Physiology of mangroves"* [@chapman1976physiology]

*Physiology* - Functioning

**Key contribution:** This chapter provides a nice overview of the physiology of mangroves, primarily in relation to:

  1. Physiology of viviparous seedlings
  2. Physiology of the pneumatophores
  3. Physiology of salt control
  4. Osmotic pressures
  5. Transpiration rates
  6. Carbon assimilation rates
  
*Viviparous seedlings* - The salinity of seedlings has been shown to be much lower than that of parent trees. This might increase buoyancy of the propagule and also protect it against very hostile environments. Seedlings quickly take up salts upon establishment, rapidly reaching to osmotic potentials of parent trees.

Additionally, seedlings are found to be able to respire anaerobically. Respiration may be inhibited by several factors, such as presence of a fruit wall (*A. marina*) or the uptake of large concentrations of salts (*R. mucronata*).

*Pneumatophores* - A number of studies have shown that pneumatophores are largely plant systems responsible for exchanges of gases. The discussion of pneumatophores here focused on three primary points:

  1. *Gaseous exchange* - Much of this is conducted through the lenticels, which may also occur on stilt roots. The individual pneumatophores are seen as being responsible for aerating the particular section of cable root it exists upon, as well as any stabilizing roots underneath.
  2. *Respiratory function* - Respiratory activity of pneumatophores likely makes up a significant (1/8 to 1/2) of total respiration.
  3. *Other purposes* - Pneumatophores may server other purposes besides gas exchange or respiration. For example:
    - *Sonneratia* and *Avicennia* pneumatophores produce very fine roots near the surface of the soil which aid in absorbing nutrients. As sediment accretes, the lower rootlets cease functioning and new sets closer to the surface are produced. Older pneumatophores may exist primarily as support organs for the absorbing rootlets
    - This is interesting in reference to carbon, soil accretion and biological contribution; it would be good to look into morphological differences of roots in mangroves. A review article?
    
*Salt control* - Different strategies exist to deal with maintaining osmotic potentials and handling salts within mangroves.

  1. *Salt exuding* - *Avicennia* and *Aegiceras* spps. are able to exude salt from glands in leaves, which is an energy intensive process. These same species do not appear to exclude salt at the roots to the same degree that other species without salt secretion glands do.
  2. *Exclusion at roots* - Species without salt secretion glands maintain high osmotic potentials within their roots, which prevents much of the salt in the root system from entering the tree.
  3. *Increasing leaf succulence* - Differential rates of storage of water and salts in vacuoles and leaf tissues vary amongst species, as well as with tree age and also environmental salinity levels. The coupling of root exclusion and increasing leaf succulence may be primary mechanisms of handling salinity for non-secreting species.

*Osmotic pressures* - Osmotic pressure is the tendency for water to enter a solution through osmosis. High OPs indicate a higher ability to take in water via osmosis. Interestingly, the OP in excess of sea water (i.e., effective OP) in mangroves is comparable to OP values in terrestrial species. When considering OP in mangroves, need to think of three compartments:

  1. OP in roots - uptake of water
  2. OP in leaves - transpiration and thus loss of water
  3. OP in vascular stem - flow rates within the plant
  
For non salt-secreting species, the flow of water through the system is largely salt-free. OPs can vary within species depending on different salinity conditions, which might be reflected in succulence or other traits measurable by remotely sensed data (hyperspectral imagery).

*Transpiration rates* - The transpiration rates of mangroves are reduced in saline conditions, and are affected by other environmental factors such as soil texture (e.g., more rapid transpiration from plants in sand than mud), but also depends on species. Evidence has been shown that transpiration rates are higher for *Avicennia* in presence of salt.

Some researchers have considered mangroves as xerophytic systems (i.e., arid systems with lack of water), but others have attacked this in that mangroves receive continuous supply of water (though limited) and thus are not living in a water limited system. For non-mangrove species living in a mangrove ecosystem, xerophytic conditions are likely present.

*Carbon assimilation* - The information provided here is cursory and simply points to a lack of information. Marilyn Ball has done much more work on this topic, particularly in relation to PWUE, and her articles should be referred to here.

### Chapman, 1976, Ch. 10 {#chapman1976ch10}

### Ball, 1988 {#ball1988}

*"Ecophysiology of mangroves"* [@ball1988ecophysiology]

**Key significance: ** This is an excellent review of the ecophysiology (how the phyisology of the trees responds to their particular environment) of mangrove trees. This paper should be reviewed more than once.

One of the **key concepts** to emerge from this study is:

  - Mangroves have largely evolved to have large tolerances to salinity and inundation given the very dynamic environments in which they exist. Rather than evolving to exist in narrow niches, competition may drive zonation and thus the existence of multiple species in moderate stress environments.

**Key notes: **

*Waterlogging* - Key fact is that oxygen diffuses through air 10,000 times more rapidly than through water, thus root systems of mangroves are adapted such that they avoid extensive inundation through areal roots. 

  - degree of soil saturation effects form and availability of different nutrients -- how specifically does this happen?
  - may have different amounts of availability of P and N along the tidal gradient
    - N primarily exists as ammonia in frequently tidal regions, which binds easily to organic compounds and thus persists within the soil system
    - N exists primarily as nitrate in less frequently flooded zones, which is more prone to leaching. Typically find that these zones are P limited rather than N limited, however, which is easily absorped onto clay or precipitated with Ca, Al, or Fe
    
*Salinity* - Important to remember that tolerance of salinity is largely dependent upon the developmental stage of the plant! The presence of cotyledonary reserves allow seedlings to persist in conditions that are otherwise inhospitable to them, but may not establish successfully following depletion of these reserves.

Mangroves frequently classified according to three different mechanisms of tolerance:

  1. salt exclusion - all mangroves exclude salt at the root tips
  2. salt accumulation - all mangroves accumulate ions (though they show preferential differences for different ions) to maintain osmoregulation
  3. salt excretion - selective species have ability to excrete salts
  
The particular salinity tolerance of a species may be particularly difficult to determine due to the species deriving water from sources other than surface water (i.e., freshwater underground seepage). Compartmentalization of ions is extremely important for maintaining biochemical processes, for example photosynthesis, especially for those species that are incapable of secreting salts.

The ratio of Na+ / K+ is important in chloroplasts (cell of the leaf responsible for photosynthesis). Deficiencies in K+ concentration have been shown to result in loss of photosynthetic capability, particularly for Avicennia species. 

The major mechanism of ion transport into the transpiration system of the plant is through symplastic pathways (inside of plasma membrane where water and low-molecular-weight solutes can diffuse freely).

Species of some genera (*Acanthus, Aegialitis, Aegiceras* and *Avicennia*) are capable of secreting salt from leaves.

*Photosynthesis* - Most evidence points towards C~3~ photosynthetic pathways within mangrove species, however they exhibit very conservative water-use efficiencies relative to most C~3~ plants. Mangroves exhibit a variety of traits/mechanisms to manage water use efficiency while maintaining carbon assimilation.

Three major leaf characteristics for maintaining leaf temperatures that vary among species are:

  1. leaf angle - affects radiant heat loading on the leaf and can be adjusted to reduce solar incidence
  2. leaf size - heat transfer is a function of the leaf geometry and wind speed; decreases in leaf size allows the leaf to maintain temperatures close to ambient temperatures, with sacrificing solar incidence
  3. heat capacity per unit area - a function of dry weight and water content per unit area; higher heat capacity buffers against warming of leaves

A classic, hypothetical schematic of interspecific tolerances for salinity is shown below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/ball1988Trees.bmp"))
```

*Major tradeoffs* - The two major tradeoffs that are seen in mangroves in relation to photosynthesis and salinity tolerances are:

  1. Salt exclusion during water uptake may induce significant carbon costs to the plant; often seen through more root biomass coupled with decrease in rate of water uptake
  2. High water use efficiency restricts rates of water efflux but also restricts rates of CO~2~ influx

**Relevant for research: **

  - differential amounts of root biomass may be seen along salinity gradients due to differing degrees of water use efficiency and maintenance of photosynthetic production - what are the implications for soil C?

### McKee, 1993 {#mckee1993}

*"Soil physicochemical patterns and mangrove species distribution - reciprocal effects?"* [@mckee1993soil]

**Key contribution: ** 

*Methods: * The study employed field and greenhouse analyses to examine the impacts of reducing soil conditions and sulphide presence on seedling growth. Field work was performed in Florida, with common zonation of *R. mangle* in the fringe forest and mixes of *R. mangle* and *A. germinans* in the basin forest that filters out to monospecific stands of *A. germinans* in the most landward basin.

Soil samples were collected along transects directly next to *Rhizophora* prop roots, *Avicennia* pneumatophores, a mix of the two roots, as well as no roots for comparison of reducing soil conditions and sulphides. A factorial treatment arrangement for four flooding conditions on two species was designed in a greenhouse experimental setting to examine the effects of reducing conditions and sulfide on seedling growth, and vice versa.

**Key results: **

*Field studies* - Distinct zonation between both species composition and reducing soil conditions and sulphide were seen. Changes in Eh and sulphide concentration strongly correlated with shifts in species compositions. 

  - Higher Eh was found for mixed basin and fringe forest, whereas lower Eh was found with monospecific-basin *Avicennia*
  - Higher sulphide concentrations were found in monospecific basin *A. germinans* region (contrasts with results of seedlings study).

*Greenhouse studies* - Greatest difference in biomass growth between flooding treatments was seen belowground. *A. germinans* exhibited much lower growth in the flooded treatments relative to the drained treatment, whereas flooding stimulated belowground growth in *R. mangle*.

Presence of sulphide had little effect on growth (AG or BG) of *R. mangle*, but was found to reduce AG and BG growth in *A. germinans*.

Additionally, the presence of both species induced higher redox potential (Eh) relative to soils without any mangrove seedilings. Addition of sulphide induced a reduction in Eh.
  - This is likely due to O~2~ leaking from mangrove roots into the rhizosphere and thus inducing higher redox potentials
  - May also reduce buildup of sulphide through oxidation of sulphide in rhizosphere or inhibiting activity of sulphate-reducing bacteria
  
**Key conclusions:** The authors make several key conclusions given their results:

  - The results of the field and greenhouse experiments do not match, in that *A. germinans* was found in the portion of the forest with the highest sulphide conditions, but seedlings were found to be highly sensitive to low redox and high sulphide concentrations.
    - Hypotheses are given that i) other controls on presence of *R. mangle* in this region are stronger than sulphide - *A. germinans* interactions, ii) may be a result of other factors (propagule dispersion, predation, competition), or iii) adult trees may exhibit different tolerances than seedlings
  - The two-way effect of redox and sulphide on roots (and roots on redox) make it difficult to infer mechanistic processes from correlative studies

**Other notes: **

  - Seawater is the primary source of sulphate for reduction by dissimilatory sulphate reducing bacteria. H~2~S may be present in degrees to be toxic to mangroves, particularly seedlings.

### McKee, 1995 {#mckee1995}

*"Seedling recruitment patterns in a Belizean mangrove forest"* [@mckee1995seedling]

**Key significance:** This study examines the relative influence of edaphic and dispersal related characteristics on species patterning of mangroves in the Twin Cays (Belize). The study is unique in that it does well to look at the effects of multiple hypotheses (ecophysiology vs. dispersal properties) on zonation.

**Research design:** Transects were laid perpendicular to the shoreline and extended through different "zones" of mangrove species. Plots were stablished along the transects and edaphic chracteristics as well as seedling densities and forest structure were collected. Additionally, tests of seedling growth were performed by transplanting individuals to forest plots of different mangrove parent species (but similar edaphic conditions). Propagules were collected and measured to examine dispersal properties.

**Key results:** McKee concludes that different factors were more important for different species. In particular:

  - *R. mangle* - Majority of variation in seedling density was explained by distance to parent trees, which would indicate dispersal properties were dominant controls
  - *A. germinans* - Availability of resources, specifically light and NH~4~ explained majority of seedling density, indicating that nutrient limitations were primary controls.
  - *L. racemosa* - A mix of dispersal and ecophysiological controls were found to explain seedling densities; however, seedlings were rare throughout the forest and thus the site conditions are likely too stressful for *L. racemosa* to exhibit canopy dominance.
  
In relation to seedling survival, the following is concluded:

  1. dispersal and establishment properties (related to predation, dessication or inability to root) were primary causes of seedling mortality (i.e., during the establishment phase)
  2. following establishment, physico-chemical factors such as salinity, flooding or nutrient availability were found to largely determine survival of seedlings
  
There are a number of nice figures that relate species zone to different criteria, but a nice visualization of quantifying relative density of species zonation is shown below (may be worth recreating for Krabi and Pak Panang):

```{r, echo=FALSE}
knitr::include_graphics(rep("images/mckee1995Oeco.bmp"))
```

**Other notes:** McKee collects a number of edaphic variables (e.g., salinity, sulphide, redox potential, etc.) and then uses PCA to extract three ecologically relevant "factors" that she uses to associate with seedling establishment and survival: i) salinity, ii) inundation, and iii) nutrients/resources. This appears to be a clever way of dealing with emergent properties and controls on mangroves and may be of interest for further examining interactive factors.


### Sherman, 1998 {#sherman1998}

*"Soil-plant interactions in a neotropical mangrove forest: Iron, phosphorus and sulfur dynamics"* [@sherman1998soil]

**Key contribution:** This study examines the interactions of species distribution and select nutrients along a 3 km transect in the Dominican Republic. Species distributions, nutrients and abiotic factors were found to covary along the transect, with different processes dominating pathways in the lower versus upper intertidal regions.

**Research design:** A 3 kilometer transect was laid out through the forest and 14 permanent, 30 m radius plots were established to measure species composition and surface and soil water biogeochemistry from 6 different dates over a two year period.

**Key results:**

*Salinity* - The salinity levels along the transect decreased as the tidal influence decreased and the influence of fresh water from rivers increased. Additionally, salinity was found to be higher at soil depths of 50 cm than the surface water.

*Sulfate* - Sulfate levels reflected changes in salinity. Sulfate was constant in concentration until 1700 m into the forest (the approximate reach of tidal inundation), at which poin it steadily declined. Sulfate concentrations were deeper in deep soil water than surface waters.

*Species presence* - Presence of *Rhizophora* was significantly correlated with salinity and sulfate concentrations of deep soilwater; *A. germinans* was significantly negatively correlated with deep soilwater salinity and sulfate. *L. racemosa* was not correlated with salinity, but was significantly correlated with surface water total phosphorus. Many of the other surface water soil parameters were not correlated with species distributions.

**Key conclusions:**

  1. The correlations between species distribution and soil chemistry parameters are not necessarily uni-directional. Others have shown that mangrove trees may have significant influence on the chemistry of the soil (e.g., McKee 1993 - oxidiation of rhizosphere which modifies redox potential and sulfide concentrations)
  2. The patterns across phosphorus, sulfate, alkalinity and pyrite (Fe) concentrations across the transect were found to be largely controlled by sulfate reduction and its associated chemical reactions.
  3. The nutrient dynamics of fringe vs basin forests vary, and are reflected in physiological strategies by different species. For example, *R. mangle* has higher nutrient resorption rates than *A. germinans*.
  4. Strong interactions appear to exist between hydrology, species distributions and soil dynamics. In the lower intertidal basin, physical processes appeared to dominate biogeochemical cycling because of high export rates of material. Chemical and biological processes in the upper intertidal region are more important given the lower amount of nutrient export.

### Krauss, 2008 {#krauss2008}

*"Environmental drivers in mangrove establishment and early development: A review"* [@krauss2008environmental]

**Key contribution:** This is likely the most comprehensive and up to date review of ecophysiology and mangrove development in mangroves. This should be reviewed several times.

**Five central paradigms** - Krauss *et al* provide five central paradigms that shape a lot of the work thinking and work related to mangrove eco-physiology:

  1. *Lack of a functional understory in most mangrove forests* - often have seedlings and saplings, but no vines, herbaceous plants or shrubs. Several studies have concluded that terrestrial understory species have just not adapted shade or salinity tolerance (or both)
  2. *Inundation classifications* - studies have shown that species are found in distinctive zones, but also that many species of mangroves are capable of colonizing a wide range of inundation frequencies; one of the earliest definitions of inundation classes was given by Watson, 1928
  3. *Salinity tolerance, zonation and biomass allocation* - Much of this work has been forwarded by M. Ball, e.g., see Ball 1988, Section \@ref(ball1988). The interaction of inundation, salinity, soil saturation and fertility has been found to explain zonation in many instances, though not necessarily all. Mangrove zonation is a hypothesis-rich field.
  4. *Interrelatedness of nutrient enrichment, growth and herbivory* - Tradeoffs exist between slow growth, defense against herbivory and nutrient limitation. Field experiments may address nutrient limitation to understand growth rates and relative survival/establishment strategies
  5. *Geomorphological classification* - A number of geomorphological classifications have been proposed.

**Temperature** - Temperature largely important for two considerations:

  1. Controls the geographical extent of mangroves globally (cold limits); different species have been shown to have different tolerances (*Avicennia* spp. typically seen in coldest conditions)
  2. Excessive heat can impact photosynthetic functioning and damage plants through direct cellular damage
    - mechanistic understanding of how temperatures affect mangrove individuals (extreme heat or cold stress) is lacking and needs further research
    
**Atmospheric CO~2~** - For a given amount of nutrients, higher CO~2~ concentrations will often increases relative growth rates, however, the particular growth seen will likely vary significantly across species.

**Salinity** - Salinity is one of the major environmental conditions that mangroves have to have evolved to be able to handle a wide range of conditions exist upon which mangroves have to be able to tolerate

Given high salinity concentrations in the external environments, mangroves have to maintain osmotic potentials below these levels (~ -2.5 to -6.0 MPa). 

Inorganic molecules are stored in the vacuoles, whereas organic salt molecules are maintained in the non-vacuolar spaces to help regulate salinity and protect functioning. A variety of mechanisms exist to maintain photosynthesis under saline conditions, but the two are largely decoupled in that salt flux to leaves does not increase with higher transpiration.

**Light** - Interactive effects between levels of light and salinity exist, thus need to consider the two in concert. The combination of the saline environment and high levels of solar incidence create a high risk for photoinhibition (in which solar incidence is greater than can be used and may damage functioning).

  - Species of mangroves differ in their strategies to avoid photoinhibition, e.g., vertical leaf angles of *Rhizophora* spps whereas *Bruguiera* spps have larger xanthophyll pigment pools
  - Biomass growth may change depending upon salinity vs. light levels, and may be different across species
  
Presence of light gaps may be particularly important for certain species (e.g., *Rhizophora*). There are a number of studies beyond [@sherman2000small] that have investigated gap dynamics and are worth looking at.
  
**Nutrients** - Mangrove ecosystems are largely nutrient limited ecosystems due to infertility of upland soils and often a lack of terrigenous input. Response of mangrove trees to nutrient additions are common to that of other forests, often with increased growth rates and reduced nutrient use efficiencies.

Several traits of mangroves exist to deal with nutrient limitations:

  1. allocation of biomass to roots (particularly fine roots) relative to shoots
  2. large maternal reserves for propagules
  3. high amounts of nutrient resorption - resorption of nutrients from leaves prior to senescence
  4. high nutrient use efficiency for photosynthesis
  5. schlerophylly and tissue quality -- reduces both decomposition and herbivory
  6. mutualistic associations, e.g., with vesicular arbuscular mycorrhizae or encrusting epibionts (?)
  
In addition, different strategies exist for high nutrient environments:

  1. investment in leaf development and rapid growth rather than defense related traits
  2. high PWUE - high extraction of water which salinizes the soil and may exclude competitors

The availability of nutrients vs other environmental conditions often induces tradeoffs among different species. 

**Flooding and SLR** - Flooding of mangroves greatly changes the soil conditions within very short time frames (less than a day). Mangrove roots may metabolize anaerobically for short periods of time, but more effective to avoid root hypoxia through internal aeration.

Mangroves may aerate the rhizosphere through loss of O~2~ from roots, which oxidizes the soil and protects against harmful sulfide. The oxidized rhizosphere may also reduce flows of O~2~ from the plant to the soil during periods of inundation.

However, mangroves are facultative halophytes and thus grow best under species-specific optimal levels of salinity and inundation, but growth levels generally are reduced under conditions beyond their particular optimum.

Prolonged inundation, for example through sea level rise, reduces growth rates which may negate the effects CO~2~ fertilization. Over more short time periods, growth was reduced quickly but normal metabolic functioning quickly rebounded (after a period of a few weeks).

**Biota** - Interacting effects of microconditions within the forest and biota, particularly insects or crabs that are predators of mangroves trees may be significant in establishement of seedlings.

Additionally they can play important roles as ecosystem engineers, aerating soil, changing microtopography, or altering the texture of the soil.


### Dangremond, 2015 {#dangremond2015}

*"Environmental tolerances of rare and common mangroves along light and salinity gradients"* [@dangremond2015environmental]



## Trophic interactions

### Smith, 1987 {#smith1987}

*"Seed predation in relation to tree dominance and distribution in mangrove forests"* [@smith1987seed]

**Key contribution: ** This is one of the earliest examinations of the effects of predation (by grapsid crabs) on propagule survival for different species. They found significant relationships between propagule predation and nutritional content, as well as location within the intertidal floor. The results led them to conclude that propagule predation is a significant factor in defining zonation in mangrove forests.

**Experimental design: ** Propagules were tethered to twine strings and monitored over 18 days in 5 replicate plots in each of three forest types (high, intermediate and low conspecific dominance) for each of the five species. Additionally, propagules of each of the five species were analyzed for nutritional content and related to predation patterns using PCA.

**Key results: ** The findings show that differential predation for certain species does occur. In particular, predation on *A. marina* in the mid intertidal was found to explain the relative absence of parent *A. marina* trees in the mid-intertidal section of the forest. Rates of predation of *A. marina* were found to be higher than in the low or high intertidal regions.

With the exception of *Ceriops tagal*, amount of predation was significantly negatively correlated with conspecific dominance (i.e., low dominance by parent trees led to high predation rates) across the other four species.

### Smith *et al*, 1989 {#smith1989}

*"Comparisons of seed predation in tropical, tidal forests from three continents"* [@smith1989comparisons]

**Key contribution:** This study extends [@smith1987seed] to examine patterns of propagule predation in four distinct localities: Florida, Panama, Australia and Malaysia. Smith *et al* conclude that propagule predation has a significant influence on spatial patterning of mangrove species, though to varying degrees depending on propagule species and also species of predator.

Smith *et al* term the hypothesis that spatial assemblages are a result of predation as the "dominance-predation" model.

**Study design:** Following the methods of [@smith1987seed], replicate plots with secured propagules were established in stands with dominance of propagule parent trees and absence of propagule parent trees. The seedlings were monitored for 4 days to identify predation such that they are incapable of establishing (i.e., 50% mass consumed, taken down hole, or radicle and hypocotyl consumed).

**Key results:** The study confirms the prior work in that species of *Avicennia* were completely consumed in stands in which their parent trees were rare or absent, but not consumed in stands where they were abundant.

The findings for *Rhizophora* species were a bit more mixed, with the "dominance-predation" model holding true in Malaysia and Australia, but failing in the North American sites. In Florida, no *Rhizophora* propagules were consumed. In Panama, more propagules were consumed in the *Rhizophora*-dominated site relative to the *Avicennia*-dominated site.

More *Bruguiera* propagules were consumed where it was rare than it was common in Malaysia, and the reverse held true for Australia.

Differences in predator distribution may explain some of the variability. In particular, grapsid crabs did not exist at the site in Florida, which may explain lack of consumption of *Rhizophora*. In forests with only snail consumers, only *Avicennia* were consumed.

This may be a result of the feeding mechanisms of crabs versus snail, and the physical structure of *Rhizophora* vs. *Avicennia* propagules (i.e., crabs are able to rip through tough hypocotyl of *Rhizophora* propagules).

### Cannicci *et al*, 2008 {#cannicci2008}

*"Faunal impact on vegetation structure and ecosystem function in mangrove forests: A review"* [@cannicci2008faunal]

**Key contribution:** This is a well-written review on faunal impacts on forest structure and ecosystem function within mangroves. It does not necessarily relate to purely herbivory (which many of the studies related to mangrove predation do), but also considers some of the community ecology aspects important for ecosytem functioning.

**Key notes: **

*Insects* - Most common forms of insect herbivory are:

  1. Leaf-feeding - can be very significant and produce massive leaf-loss events
  2. Wood-boring - very limited evidence (Feller in Belize), but has found to produce leaf-loss as great or greater than direct leaf herbivory. Insects bore wood on branches and leaves along the branch are subsequently lost, which can create significant canopy gaps
  3. Flower/fruit/seed-feeding - may be a result of reduced biomass allocation to reproductive tissues following leaf herbivory or other damage; or may be a result of direct feeding on propagules
  
The costs of leaf herbivory and stress on the plant are a function of both the direct impacts, as well as the additional biomass and resources needed to be put into regenerating leaves or building defensive mechanisms. Toxins (*E. allagocha*; Euphorbiaceae), tannins or thicker leaves may require resource inputs that could otherwise be allocated towards growth.

Ants are the most dominant insects in mangroves, both numerically and energetically, despite the difficult floor conditions in which many require. Ant communities consist of both terrestrial species as well as those endemic to mangroves.

Although data is limited, studies have shown that ants in mangroves can significantly reduce the presence of herbivorous pests (note that this is decoupled from improving plant "performance" *per se*).

*Crabs* - Feeding patterns of crabs vary by species within mangrove. Some predominantly feed on leaf litter, whereas others have adopted arborial life style and feed largely on green leaves. They scrape the leaves which likely has similar impacts on tree fitness as other herbivores.

Crabs understood as main agents of high leaf litter turnover in mangroves. In particular, sesarmid crabs in Southeast Asia are main consumers of leaf litter, though magnitude of influence is highly variable.

In Neotropics, ocypodidae crabs are found to be most influential on leaf litter dynamics.

In general, however, feeding behavior of crabs is poorly understood within mangroves.

Biomass cycling of crabs:

  1. Low assimilation rates of leaf litter (<50%)
  2. Additionally, approximately 60% of dry weight is defecated
    - Crab fecal matter may be of higher nutritional content than leaf litter, acting as food source for benthic organisms
    - Shredding of leaves occurs, and bacterial diversity on crab fecal matter found to be 70 times greater than that of raw leaf litter
    
"Reduced competition from crabs" - Several mechanisms have been proposed for reductions in seedling/sapling competition due to propagule predation:

  1. *Predation-dependence model* - inverse relationship between rate of predation and dominance of particular species
    - Smith has found support for it, whereas a number of other studies have found a lack of evidence in support of the hypothesis and thus have proposed alternative mechanisms
  2. *Canopy-gap mediated model* - hypothesizes that predation is more intense under closed canopies than in light gaps
    - Has led some to propose a "mutualism" between sesarmid crabs and mangrove trees in that crabs reduce competition and allow established saplings to compete while mangroves provide food
  3. *Flooding regime model* - time for terrestrial foragers is limited due to variation in exposure of the forest floor
    - Sousa found otherwise
  
"Bioturbation of soil" - Impacts of crabs as ecosystem engineers may have strong implications for soil quality within mangroves:

  - Sesarmid crabs were found to significantly reduce aluminum and sulfur contents of soil
  - Fiddler crabs may increase the depth of iron reduction
  - Burrowing create preferential flow pathways which may flush out accumulated salts, particularly within rhizospheres
  
*Molluscs* - Most well represented taxon along with decapod crustaceans, however their ecological role is poorly understood.

  - Snails can be significant feeders of leaf-litter, feeding during both low and high tides in which they chemically sense leaves underwater
  - Sponges may play a role in inducing growth of adventitious rootlets that absorb ammonium and other nitrogenous compounds
  - Oysters and barnacles may have negative effects on root growth, and may also result in breaking of roots due to harvesting by humans

### Sousa, 2011 {#sousa2011}

*"Trophic interactions in coastal and estuarine mangrove forest ecosystems"* [@sousa2011trophic]

**Key contribution:** This is an extensive review on trophic interactions in mangroves. It extends beyond conventional consideration of herbivory and the effects of trophic interactions on forest structure by including discussion of faunal interactions.

**Productivity** - Mangroves have been found to have higher aboveground net primary productivity rates (mean ~14 Mg DW / ha) than terrestrial tropical forests (mean 5 Mg DW / ha).

Methods of measuring NPP are varied and inaccurate, and thus the estimates provided within the literature are likely to be quite coarse. Little research has examined the productivity of phytoplanton in tidal waters or cyanobacteria on aboveground root systems, though some studies have shown this to be half or more of ecosystem productivity.

**Herbivory** - Primary component of energy flow through ecosystem, yet relatively unstudied within mangrove ecosystems.

*Insect herbivory* - General consensus is that insect herbivory is not a significant consideration, although this is likely based on little evidence. Methodological issues constrain our current estimates of the degree of leaf herbivory by insects.

Of the studies that exist, rates of leaf removal are relatively low at 7.6% for *Avicennia* spps and 5% for *Rhizophora* spps. Longitudinal studies that have followed leaf herbivory through the duration of its development (from bud to the end of the study period) have found herbivory rates to be much higher (10-30%).

Additionally, occasional mass herbivory events due to emergence of lepidopterans can defoliate massive expanses of trees. Caterpillars emerge and within three months trees have lost > 75% of foliage.

*Crab herbivory* - Majority of crabs forage on forest floor, but some exhibit (almost) purely arborial habits Some are found in both terrestrial and low tree canopies, whereas four species of sesarmid crabs are purely arboreal (come to floor rarely if ever).

  1. *Aratus pisonii* - Atlantic/Caribbean/Eastern Pacific; feeds on young leaves largely from *R mangle*
  2. *Parasesarma leptosoma* - Indo-West Pacific; feeds on young leaves
  3. *Armases elegans* - West Africa, feeds on young leaves
  4. *Selatium brockii* - Feeds on algae growing on tree trunks
  
*Stem boring* - wood-eating beetles and moths can have significant impacts on forest structure, particularly through the opening of small gaps within the forest via death of branches.

*Coccotrypes rhizophorae* - *Rhizophora* specialist beetle that bores roots, wood and propagules; can change morphology of root systems (Atlantic/Caribbean/Eastern Pacific)

*Coccotrypes fallax* - the congener to *C rhizophorae* in the Indo-West Pacific

*Propagule predation* - Propagule predation may occur before or after dispersal:

  - "Predispersal predation" occurs when propagules are still on parent trees. Often have boring insects (e.g., *C. rhizophorae*) which may ultimately cause premature dropping of propagules, their death, or reduced growth rates of seedlings following establishment. *C. rhizophorae* lay eggs in *Rhizophora* propagules and can greatly damage the propagules.
  - "Postdispersal predation" - a relative boom in research following Smith's tethered propagules studies. Although findings corroborate with his conclusions per *Avicennia* propagules in some cases, other studies have shown the predation-dependence model to not hold true.

**Detrivory and decomposition:**

*Outwelling hypothesis* - idea that substantial amounts of organic matter are exported to adjacent estuaries, which supports dense populations of detritivores and secondary and tertiary consumers

Odum and Heald's work in Florida exemplified the process by which outwelling of organic matter in mangroves may occur:

  1. Mangrove leaf litter drops and undergoes decomposition by bacteria, fungi and protozoa
  2. Crabs and other detritivores further consume leaf litter, breaking it up into finer pieces
  3. Organic matter in fine, processed form is exported via tidal flows
  4. Exported OM fuels marine and detritus-based food chains (postulated)

The importance of this particular pathway has since been questioned, as supporting evidence has been largely been circumstantial. The circumstantial evidence of this pathway may indicate a large contribution of DOC.

Outwelling of OM likely occurs in most tidally inundated and riverine forests, though the degree to which it happens depends on site, season and years.

Different patterns of outwelling exist for basin forests separated by berms, in which only highest tides flush the system, versus riverine or fringe forests that undergo more periodic flushing.

*Retention and recycling of carbon and nitrogen* - export of organic matter and nutrients is also greatly controlled by recycling of nutrients within the system.

*Detritus-feeding crustaceans* - A review of studies finds that, on average, 57% of detritus is either consumed or buried by crabs.

Studies have found a strong preference by crabs for older, more decomposed leaf litter in which C:N ratios are much lower, and tannin content and toughness of the leaf are much reduced. Nevertheless, the C:N ratio is still quite high and thus crabs likely need to look for additional N sources. Studies have suggested that this may come from grazing on algae or predation of animal tissue (crustaceans, mollusks, annelids, insects and fish).

*Detritus-feeding mollusks* - consumption of detritus by mollusks can be significant in addition to that of crustaceans.

*Deposit-feeding crabs* - largely comprise of fiddler crabs; feed on bacteria, microalgae and detrital material (often bacteria, smaller diatoms and protozoa); assimilation efficiency is much higher for bacteria than microalgae

*Litter decomposition* - the nutritional content of leaves from different species varies, but in general decomposition improves the nutritional quality over time. Tannins are high but are leached from the leaf quite rapidly; following leaching of soluble compounds, lignocellulose is primary component; bacteria digest lignocellulose at a much faster rate than other decomposters, and animals cannot digest lignocellulose

*Wood and root decomposition* - a higher proportion of lignocellulose exists in wood and roots, thus decomposition is much slower than for leaves.

Little is known about the functional groups and diversity of fungi in mangroves, but they may play a significant role in decomposition of wood and roots. Additionally, wood-boring insects with mutualistic relationships with bacteria that can process lignocellulose may be important for decomposition of woody biomass (e.g., termites).

*Predation* - In addition, the review provides a discussion of predator-prey interactions but I have not included notes here as it is beyond the direct scope of the qualifying exam.



## Sea level rise & biogeography

### Ellison *et al*, 1999 {#ellison1999}

*"Origins of mangrove ecosystems and the mangrove biodiversity anomaly"* [@ellison1999origins]

**Key significance:** This study examines two separate hypotheses for the distribution of mangrove diversity globally. In essence, they seek to understand the stark decreases in species diversity in the ACEP region relative to IWP.

**Key notes: **

Two primary hypotheses for the global distribution of mangrove species diversity exist:

  1. "Center of origin" hypothesis - holds that mangrove taxa originated in the IWP and then dispersed to other parts of the world
  2. "Vicariance hypothesis" - taxa evolved around the Tethys Sea and regional species diversity has results from in situ diversification after continential drift
  
**Research design:** The authors consider primarily five lines of evidence to parse whether the Center of Origin versus Vicariance Hypothesis is more likely to explain current patterns of species diversity. The five lines of evidence they use are:

  1. Fossil records - distribution of fossils all around Tethys would support vicariance, whereas fossils in Eastern Tethys (IWP) would support Center of Origin.
  2. Available area:
    - Center of origin - species richness negatively correlated with distance from center of origin rather than available area
    - Vicarance - species richness should be correlated with availble area
  3. Distributions of fauna - Would expect to see similarities in diversity anomaly between mangrove tree species and the fauna that inhabit the forests
  4. Geographical nestedness - geographical nestedness may occur through colonization of far-reaching areas by select species, or by extinction of taxa following vicariance.
  5. Nestedness of entire communities - nestedness of mangroves and all gastropods as described above

**Key results:** Ellison *et al* conclude that the lines of evidence provides support for the vicariance hypothesis. 

In particular, strong species-area relationships were found, evidence in the fossil record, and a unimodality in the distribution of gastropod genera were found to be convinving evidence in support of the vicariance hypothesis.

### Mckee & Faulkner, 2000 {#mckee2000}

*"Mangrove peat analysis and reconstruction of vegetation history at the Pelican Cays, Belize"* 

**Key contribution:** This study examines the historical development of peat formation in the Pelican Cays in Belize. This work is a precursor to [@mckee2007caribbean] which studied the formation of peat deposits in the Belize cays in depth.

**Study design:** A transect across Pelican Cays was laid out that spanned *Rhizophora* fringe, *Avicennia* basin/woodlands, a treeless area with standing water, and the same pattern to the other coast. Peat cores to 0.5 m and 2 m were taken and soil and peat characteristics were described.

**Key results:** Upon identifying roots, shoots and leaf fragments within the peat to species type, the authors were able to infer the successional process of the forest. In stands that are currently *Avicennia* dominated, *Rhizophora* roots make up the majority of the deep peat, indicating they once dominated that stand. In addition, root biomass was found to make up the majority (~80%) of the peat biomass.

### Valiela *et al*, 2001 {#valiela2001}

*"Mangrove forests: One of the world's threatened major tropical environments"* [@valiela2001mangrove]

**Key contribution:** Following widespread recognition of the loss of other tropical ecosystems (rain forests and coral reefs), this paper makes the argument that mangroves are under-appreciated but perhaps even more threatened than these other ecosystems.

**Study design:** They compile data on the global status of mangroves from the literature, and assess to what degree human activities have contributed to their loss.

They use the Global Mangrove Atlas coupled with estimates of status and extent of mangroves across different countries to look at rates of loss for different continents and globally.

They assess the human-based impacts on mangrove extent by correlating lost mangrove extent with GDP and also coastal population density.

**Key results:** The study estimates that approximately 35% of mangroves have been lost from their "original" extent (i.e., the earliest year in which data could be found; 1920s for some data from Africa and Asia).

Global mean rates of loss were approximately 2% per year, with continent specific rates ranging from 1 to 3.6% (3.6% was for the Americas).

They attribute approximately two-thirds of the loss of mangroves directly to human related activities. The correlation between GDP and loss of mangrove extent is significant, whereas little correlation between loss of mangroves and coastal population density is found.

Furthermore, Valiela *et al* delve into the importance of mariculture for mangrove habitat losses. Conversion of mangroves to shrimp ponds is the dominant driver of deforestation, and has additional indirect impacts on the environment such as:

  - Reductions in offshore shrimp stocks due to need for juveniles
  - Impacts on offshore fisheries that are used as feed within the shrimp ponds

### Krauss *et al*, 2014 {#krauss2014}

*"How mangrove forests adjust to rising sea level"* [@krauss2014mangrove]

**Key contribution:** This review differs from others in that it focuses on the role that vegetation plays in maintaining soil surface elevations in mangroves.

**Key notes: **

As of 2010, current rate of mean sea level rise is 3.2 +/- 0.4 mm per yr (based on satellite altimetry), though these rates likely vary locally. IPCC projections (AR5) estimate SLR rates of 3 - 5 mm / yr over the next 100 years.

"Relative sea level rise" refers to the rate of sea level rise in conjunction with subsidence or vertical accretion of coastal areas (i.e., consideration of both eustatic SLR as well as changes in the mangrove surface elevation).

Surface elevation changes are a result of both biotic and abiotic processes. Abiotic processes include:

  1. Inorganic sedimentation
  2. Groundwater fluxes (anthropogenic depletion of groundwater reservoirs may be significant)
  3. Deep land movements
  
Whereas biological processes include:

  1. Plant litter and woody debris deposition
  2. Root accumulation
  3. Sediment trapping by vegetative structures
  4. Algal mat development on soil surface

Other processes include acute deposition from storm events, or feedback processes between vegetation systems and degree of inundation.

*Aerial roots and sedimentation* - Different functional root types will influence sedimentation rates differently, and through different processes. Pneumatophores are believed to retain deposited soil better than other types, whereas stilt roots may result in greater deposition of sediment.

*Litter and woody debris accumulation* - In absence of tidal flushing as well as crabs and other organisms responsible for turnover of organic matter, accumulation of OM may be significant.

*Benthic mat formation* - Intact algal mats may bind sediments and resist erosion. 

*Subsurface root accumulation* - May be the most important biological contributor to surface elevation change in some settings (carbonate settings or other settings with limited mineral sediment input). Can be significant enough to affect the direction and rate of change in surface elevation.

  - Refractory nature of root tissues and existence within anaerobic environment are prone to slow rates of decomposition
  - Rates of SEC were correlated with both fine and coarse root production
  - "Specific root length" is a measure of the length of root per unit mass; mangroves have more dense roots than tropical species, which is likely due to anoxic conditions

Influence of roots on surface elevation is bidirectional; death of roots can result in collapse of roots under weight of soil

*Factors affecting root contributions to soil accretion*: 

In general, the influence or roots on SEC is a relative function of production and decomposition of roots.

  - Salinity - response of root growth to salinity is species dependent, but root growth only slightly reduced with increasing salinity; increases in root to shoot ratio may actually be a result of increased aboveground biomass rather than increased root biomass (for example through nutrient release)
  - Nutrients - in general do not see an increase in root production, but rather increases in biomass production aboveground; in some settings may even see decrease in elevation due to release of nutrient-limited microbial communities and thus decomposition of belowground organic matter
  - Flooding - effect of flooding on root-production is hard to identify, and may occur in both directions depending on setting (increasing root-production in basin type settings, decreasing root production in others)
  - Soil texture - high bulk density decreases root growth due to more difficult conditions of growing
  - Disturbance - may be biotic or abiotic: 
    - Largely hypothesized that organisms may influence root production, though the mechanisms and degree to which this occurs is unknown
    - Acute deposition of sediments from storm events may kill trees, or may provide much needed substrate for subsiding forests; root zone expansion following deposition of sediment may be significant and may further increase surface elevations
    - Lightning may kill trees, causing root collapse and alter microhabitat hydrology (Sherman, 2000)

*Climatic and environmental feedbacks*:

  1. Precipitation -- seasonal or storm based variability in rainfall can cause swelling or subsidence of peat layers or water table
  2. Elevated CO~2~ - likely that increases in CO~2~ concentrations will improve biomass growth, though most of this is likely to occur in the aboveground systems
  3. Feedbacks related to SLR - growth rates may increase sharply for a short period of time, but eventually fall off under prolonged inundation

### Gilman, 2008 {#gilman2008}

*"Threats to mangroves from climate change and adaptation options: A review"* [@gilman2008threats]

**Key significance:** The paper reviews the dominant threats to mangroves from climate change related impacts. In addition, they provide several suggestions for measures that can be taken to potentially bolster mangrove resilience to climate change related pressures.

**Key notes: ** The key threats to mangroves from climate change as identified by the paper are:

*Sea level rise* - relative sea level rise is a key threat, and is relatively difficult to measure as tidal gauges may not accurately reflect sea level conditions within the actual forest of interest. 

Mangroves are vulnerable when they are unable to increase surface elevation at a rate greater than sea level rise, which is due to both surface accretion as well as belowground processes.

Considerable feedbacks exist in which increasing surface elevation may decrease hydroperiod and thus sediment accretion, or alternatively subsidence may promote hydroperiod and increased sediment accretion.

Three general response of mangroves to changing sea-levels are:

  1. Stable site-specific relative sea-level - sea-level not changing relative to mangrove and thus mangrove is "stable"
  2. Site-specific relative sea subsidence - mangrove margins migrate seawards
  3. Site-specific relative sea rising - mangroves retreat landward; may expand into landward areas
  
Other factors beyond sea level rise are also important for determining resilience of mangroves to SLR. Four main factors are identified:

  1. Rate of SLR relative to changes in mangrove surface elevation
  2. Species composition - different species have different rates of change in sediment elevation
  3. Physiographic setting - slope of land upslope and ability of mangroves to migrate landwards
  4. Cumulative effects of stressors - effect ability to persist and also colonize new areas
  
*Extreme high water events* - The impacts of high water events is relatively poorly understood at present, but the impacts may be significant or may induce extensive stress in mangroves.

*Storms* - Winds, precipitation and storm surge heights are expected to increase with climate change. More intense storms may damage or defoliate mangroves, but may also cause erosion, peat collapse, or otherwise adversely impact edaphic conditions.

*Precipitation* - Projections are for more precipitation, but in a spatially heterogeneous manner. Changes in precipitation (total as well as seasonal amounts) may impact vegetation growth, soil conditions, and also ability of the mangrove to migrate landwards or seawards.

*Temperature* - The main effects of temperature on mangroves are:

  1. Changing species composition
  2. Changing phenological patterns
  3. Increasing mangrove productivity where temperature doesn't exceed an upper threshold
  4. Expanding mangrove ranges to northern latitudes
  
*Atmospheric CO~2~* - Likely to improve productivity of trees in low-salinity environments, but may not have substantial impact on those in high-salinity environments.

*Recommendations for improving resilience* - The author provide a list of recommendations for improving mangrove resilience to the impacts of sea level rise:

  1. "No regret" reductions of stress -- Simply better management techniques that reduce anthropogenic stresses on existing mangrove forests
  2. Manage activites that directly influence sediment elevation -- Not over-depleting groundwater reservoirs or cutting off landward supplies sediment to the forest will improve its ability to respond to SLR
  3. Managed retreat -- Manage landscapes such that mangroves have a region in which they can retreat towards 
  4. System of protected area networks -- Ensuring representation of mangrove types within protected area networks or replication of identical communities to improve resilience to risk (i.e., "portfolios" of communities)
  5. Mangrove restoration -- Restoring impacted areas through reforestation or afforestation
  6. Regional monitoring network -- Using standardized techniques to monitor and examine key criteria (e.g., SLR and surface elevation change *a la* Sasmito 2016)
  7. Education and outrech

### Sasmito *et al* 2016 {#sasmito2016}

*"Can mangroves keep pace with contemporary sea level rise? A global data review"* [@sasmito2016can]

**Key contribution:** This study systematically reviews data on surface elevation change (SEC) and surface accretion rate (SAR) in mangroves to examine their ability to keep pace with projections of sea level rise globally.

Of critical importance is:

  1. Surface accretion rate (SAR) -- process of sediment accumulation on the surface of the forest floor
  2. Surface elevation change (SEC) -- a function of biotic and abiotic processes, such as growth of roots or compaction or subsidence due to OM decomposition
  3. Sub-surface change (SSC) -- the difference between SAR and SEC
  
The data review examines studies that use Rod Surface Elevation Table and Marker Horizon methods to examine elevation change.

**Key results** Only 19 studies with data from some 74 sites were found through the literature review. Running different models of the influence of geomorphological setting and management type (restored, plantation, natural, etc.) on SAR, SEC and SSC found them to be influential.

In particular, basin mangroves may be better able to adapt to SLR than fringe mangroves or those on islands with shallow tidal ranges (and thus little sedimentation).

The study nods to the importance of belowground processes in conjunction with geomorphological setting and management practices for SEC.

Extrapolation of SAR and SEC rates with projected SLR shows that mangroves may be able to keep pace with low (RCP 2.6) emissions trajectories, but are unlikely to cope with high emissions pathways (RCP 8.5) regardless of geomorphological setting or management regime.

Additionally, there is more systematic monitoring of SEC using RSET-MH in Indonesia, the Philippines, Singapore, SW Thailand and Vietnam underway. This data will better inform the ability of mangroves to respond to sea level rise.

  - Additional thought -- how can this be worked into my research trajectory?

**Relevance for research:** Interesting in that they call for standardized global monitoring of surface elevation data. Could potentially work into remote sensing of geomorphology in mangroves.


## Carbon

### Kristensen et al, 2008 {#kristensen2008}

*"Organic carbon dynamics in mangrove ecosystems: A review"* [@kristensen2008organic]

**Key significance: ** 

*Origin of mangrove organic C*

Stable isotopes can be effective in determining sourcing of organic C in sediments.

Mangrove sediments are generally relatively rich in organic C, with a median value of 2.2% and 28% of values between 2 and 5%.

*Chemical composition of mangrove litter*

Rhizophora leaves in particular have undergone extensive work on their chemical construction. Rhizophora leaves are comprised of: carbohydrates, amino acids, lignin-derived phenols, tannins, fatty acids, triterpenoids, and n-aklanes.

Mangrove litter thus provide a particular signature that is preserved during transformations and may be helpful in sourcing of organic C in mangroves.

Carbohydrates form the bulk of mangrove tissue with rates of ~10-30% in leaves, and up to 65% in woody tissues.

Amino acids from the second most

Tannins form the third most. Mangrove have relatively high contents of tannins, which affects the feeding patterns of crabs. Leaching of tannins from leaf litter is important step in decomposition of leaf litter.

Phenols derived from lignins form a smaller portion, and are more refractory than other molecular compounds.

Fatty acids are useful in identifying sources of OM as they are uniquitous in living organisms and have biological specificity.

*Foodwebs and role of fauna in OM processing*

Mangrove derived OM is key source, but additionally have OM sources from epiphytic algae, benthic microalgae, bacteria and fungi, and macroalgae and mixed OM imported by tidal currents.

Fauna play important roles in OM dynamics, for example processing of litter into smaller pieces, as well as transport of litter into belowground pools. Important in nutrient dynamics, oxidation of soil, and oxidation of reduced compounds.

*Organic matter decomposition and mangrove sediment biogeochemistry*

Leaf letter that is not exported enters sediment and is processed by microorganisms. Stepwise process of decay is:

  - Significant leaching of soluble organic substances - carbohydrates, tannins and phenolic compounds
  - Extracellular enzymatic hydrolysis - mediated by aerobic and anaerobic prokaryotes and marine mycelial decomposers
    - fungi particularly adapted for capture of cellulose-rich litter
    
Various controls on the speed at which leaf litter is decomposed exist, for example how wet it is (controls access to microbial decomposers), or the C:N ratio of the litter (higher N = faster decomp rate).

Sedimentary decomposition pathways occur largely under anaerobic conditions:

  - Large OM molecules split to small moieties by fermenting prokaryotes
  - These molecules then oxidized completely to CO~2~ by wide variety of anaerobic microorganisms via electron acceptors in following sequence: Mn^4+^, NO~3~^-^, Fe^3+^, and SO~4~^2-^
  
Aerobic and sulfate reduction are most important processes, and account for approximately 40-50% of reduction each. Thus, many mangroves have reduced inorganic S in form of FeS~2~ and S^0^.

Denitrification may be significant in areas impacted by sewage.

Sulfur reduction is hampered by presence of more ready electron acceptors (O~2~ and Fe^3+^), and thus in environments in which Fe^3+^ is high, sulfur reduction may be less significant than Fe respiration.

Methanogenesis would occur if all electron acceptors are depleted; however ocean water provides steady supply of S and thus largely prevents system from reaching a point in which methanogenesis occurs.

The type of respiration that occurs is influenced by the forest conditions, in particular age: younger forests with potentially sandy substratum may exhibit more aerobic and Fe reduction whereas older forests with strong organic C input may exhibit sulfate reduction.

*Burial and permanent storage in sediments*

C burial depends on environmental conditions. For example, older stands have higher productivity and consequently higher efficiency of C burial in sediments. There is also proportionally greater C burial in low intertidal.

*Outwelling hypothesis*

Clear evidence that significant amounts (~half) of mangrove litter is exported to near coastal ecosystem, the use of this OM in coastal food webs is inconsistent.

The influence of OM exported from mangroves is likely limited to within a km or 2 from the forest. The lack of a major outwelling phenomena may be explained by:

  1. distribution of exported mangrove litter depends on hydrogeomorphology, which may result in direct sedimentation in immediate vicinity of the forest
  2. over the time-scale at which outwelling occurs, significant fraction of litter is lost as DOC
  
Export of carbon to nearshore ecosystem is significant in mangroves and thus loss of mangrove globally may have had significant implications for DOC to the ocean.

### Alongi, 2012 {#alongi2012}

*"Carbon sequestration in mangrove forests"* [@alongi2012carbon]

**Key significance:** This article is Daniel Alongi's own mechanistic review of carbon sequestration in mangroves. There is considerable overlap with some of the other reviews that are listed here.

**Key notes: **

*Carbon allocation and ecosystem storage*

In understanding C sequestration in mangroves, it is critical to understand the various pools from C sequestration by mangrove trees:

  - Approximately 50% is respired by trees, though this is rough estimate given limited data, particularly for root systems
    - Root respiration and production is particularly difficult to measure
    - Recent analyses suggest greater proportions of C may be allocated to roots in mangroves than other trees

The relative dominance of the belowground pool in ecosystem C appears to be a global pattern.

Most living roots are within top 40 cm of soil, but the sloughing off of fine roots can be considerable and be a large addition of C to soil.

Large dead root pool may be a nutrient conserving mechanism, in that new roots can enter dead root channels and reassimilate released nutrients from decomposing biomass. Extensive root growth may also aid in uptake of water, other nutrients, as well as stabilize against tide and winds.

*Rates of soil accretion & C sequestration*

Over short-term, C is stored in biomass whereas over long-term, C is stored primarily belowground.

Sediment accretion averages 5 mm per yr globally, but some highly-impacted sites (e.g., in China) show rates of up to 44 mm per yr.

Frequency of tidal inundation is the primary factor in determining the rate of accretion.

Deforestation of mangroves results in collapse of peat soils and may result in significant emissions of CO~2~, which may be up to 2-10% of emissions from tropical deforestation.

*Global significance*

Mangroves account for approximately 3% of C sequestration though they occupy <1% of total area of tropical forests.

### Alongi, 2014 {#alongi2014}

*"Carbon cycling and storage in mangrove forests"* [@alongi2014carbon]

**Key significance:** This ARMS review paper covers carbon dynamics in mangroves with a focus on both mechanistic processes as well as mean ecosystem and global budgets. The paper is an excellent overview of the state of knowledge on C dynamics in mangroves.

**Key notes: **

The primary sources of carbon within mangroves are:

  1. Trees that fix atmospheric carbon via photosynthesis
  2. Macroalgae colonizing aboveground roots and the forest floor
  3. Transport and deposition of materials from adjacent ecosystems
  
Although the relative contributions depend upon site location and geomorphological conditions, most C is produced in sit and stored belowground.

*NPP*

Mangroves have high NPP, typically higher than that of other coastal systems, but relatively equivalent to coral reefs. Mangrove NPP rates are approximately equivalent to both humid tropical forests as well as peat swamp forests.

Although inconclusive, there is evidence that disturbance may prolong maturation of forests and thus have long-term production and sequestration of C in mangroves (photosynthetic rate increases rapidly in first 20 years, but levels off without declining for up to 80 years).

*Sedimentation*

Mass balance approaches in mangroves determined to be highly inaccurate.

Frequency of tidal inundation is primarily control on sedimentation rates, i.e., forests further from the sea experience less sedimentation.

Size, shape and distribution patterns of trees are critical in sedimentation in mangroves.

Burial rates of carbon in soils are estimated by difference between carbon losses (export, consumption and mineralization) and annual accumulation rates (rest is presumed to be preserved).

Global estimate for soil C burial rate in mangroves is approximately 24 Tg C per yr. Relatively small on global scale, but is significant given their occupation of a relatively small amount of global coastal ocean area ~0.5%.

Some estimates place deforestation of mangroves at comprising approximately 2-10% of global CO~2~ emissions from deforestation.

*Losing carbon*

Key losses of carbon are due to respiration and tidal export. Majority of respiration is from trees.

Tidal export is a key process that is relatively poorly understood. May occur as:

  1. Particulate organic carbon - export of particulate organic matter (leaves, bark, branches, etc.); accounts for approximately 10-11% of terrestrial carbon input to ocean
  2. Dissolved organic carbon - organic carbon that is able to pass through a filter
  3. Dissolved inorganic carbon - mangrove contribution of DIC to ocean is disproportionate to areas of waterways and creeks; tidal pumping from forest floor is likely significant
    - microbial metabolism in deep soil results in CO~2~ production, which dissolves in interstitial waters and is exported to coastal ocean (aerobic respiration and sulfate reduction are usually major pathways, though Fe and Mn reduction may also be important)
    
*Global carbon budget*

A schematic of the global carbon budget in mangroves is displayed below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/alongi2014.bmp"))
```

Of key significance is that approximately 24 Tg C per yr is buried globally, and approximately 60% of mangrove NPP is estimates to be exported as dissolved and particular carbon to coastal ocean (~129 Tg C per yr).

