# Soils & geomorphology

## Background

### Jenny, 1994 {#jenny1994}

*"Factors of soil formation: A quantitative system of pedology"* [@jenny1994factors]

**Key contribution:** This is Hans Jenny's seminal work on the factors of soil formation. The most significant contribution of the work is his conceptual model for soil formation, which is shown below:

  $$soils = f(climate,\ organisms,\ relief,\ parent\ material,\ time,\ humans)$$
The model allows for conceptual investigation of the various factors that influence the formation of a soil "system." Jenny defined soils as "systems" that were largely influenced by initial states (relief, parent material, time) as well as their surroundings (climate, organisms).

### Amundson, 2003 {#amundson2003}

*"Soil formation"* [@amundson2003soil]

**Key contribution:** An excellent review of key studies and thinking around soil formation, that is extended by [@amundson2014soil]. It is worth reviewing both of these articles.

**Key notes: **

*Factors of soil formation*

Climate and organisms are the "surrounding environment," whereas topography, parent material and time represent the "initial state of the system".

One of best examples of use of factors of soil formation model to quantify soil development is the relationship between soil organic carbon and nitrogen storage to climate (mean annual temperature and precipitation).

More specifically, soil organic carbon generally increases with decreasing temperature and increasing precipitation. Fairly robust relationship that has emerged through 7-8 decades of work.

*Soil morphology*

Soil formation is a depth dependent process.

Present US nomenclature for naming soil horizons consists of two key features:

  1. Upper case "master horizon" indicator (O, A, E, B, C, R, W)
  2. Lower case modifier that provides more information on horizon or processes that formed it (a, b, c, d, e, etc.)
  
Although most soil formation processes are continuous with depth, some occur at abrupt intervals as evidenced by banding within soil profiles. A good example of this is a prominent E horizon present in spodosols, which results from leaching of silicate clay, iron, and aluminum leaving only sand and silt.

The "master horizon" indicators are provided below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/amundson2003master.bmp"))
```

In addition, the modifier characteristics are shown below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/amundson2003modifiers.bmp"))
```

*Mass balance models of soil formation*

In general, the gain or loss of a given chemical element may be described as:

$$ m_{j, flux} = m_{j,s} - m_{j,p} $$
Where m~j~ is the element j, in the soil (s) or parent material (p). Thus, fluxes are determined as the content of a mineral relative to the parent material of that soil.

Volumetric collapse may occur through weathering (leaching of elements, compaction) whereas expansion may occur through biological activity or physical processes.

*Mass balance equation of biogeochemistry of soil formation*

Principal processes of chemical differentiation relevant for the formation of soil:

  1. Chemical evolution of universe/solar system
  2. Chemical differentiation of Earth from solar system components
  3. Biogeochemical effects of soil formation on crustal chemistry
  
Chemical composition of crustal rock on earth is a result of both:

  1. Formation and accretion of the earth itself
  2. Differentiation in element composition of the core, mantle and crust
    - Crust is generally depleted in noble gases, H, C and N, but enriched in the other elements
    - Parent material is dominated by O, Si, Al, Fe, Ca, Na, Mg, and K

The "enrichment factor" is the ratio of the mass concentration of elements in soils by those in crustal rocks. Values less than 1 indicate losses from that soil system, whereas values greater than 1 indicate gains.

In general, soils are:

  - Very depleted in Mg, Na, Ca, K, and Be (Beryllium) due to aqueous weathering (loss of cations)
  - Somewhat depleted in Si, Fe, and Al
  - Enriched in C, N, and S
    - Vast majority of enrichment of C and N occurs through biological activity
    
*Mass balance of soil formation with respect to time*

For nonarid, temperate environments, soil formation typically is marked by soil expansion as a result of physical and biological processes, followed by collapse as a result of chemical weathering of the soil.

For cool tropical climates, the influx of nutrients via wind-blown atmospheric sources may be important. Initially, all major nutrients for plants are available with exception of C and N. Following their incorporation into the soil by early OM production and cycling, a peak productivity is reached before other nutrients are depleted and OM production begins to decline. At this point, atmospheric inputs of key nutrients may be important.

The importance of atmospheric inputs of chemical nutrients is particularly important for landscapes in which geological rejuvenation does not occur (i.e., no tectonic action).

*Mass balance of soil formation with respect to climate*

In examining hyper-arid soils, the lack of chemical weathering and atmospheric accretion of chemical constituents results in volumetric expansion of soil and accumulation of essential plant nutrients. This is indicative that at a critical threshold of precipitation, accumulation of soils may occur. This is also indicative of the importance of water in chemical weathering processes.

*Matter and energy transfer processes in soils*

Movement of soil constituents can broadly be described as diffusive or advective processes. Study of these processes is largely wrapped up within the domain of "soil physics".

  - "Advection" - transport with mean fluid flow
  - "Diffusion" - transport of compounds through action of random motions; often occurs against mean flow
  
Diffusive transport is predominant form of movement in soils.

The soil carbon cycle is one of the few chemical constituent cycles that has been described in detail.

*Mechanistic modeling of organic and inorganic carbon in soils*

Extremely important to consider variation in soil carbon with depth. Different soil carbon dynamics and concentrations across depth profiles may emerge. 

Graphs of SOC concentrations across depth for mollisols are reminiscent of mangroves, aridosols are relatively low and flat but with some variation, spodosols may have a sharp spike or dent in the graph as a result of the extensive leaching in the E horizon.

Modeling carbon movement into soils:

  - Depends on the soil type, but for grasslands, the general process can be modeled as:
  
  $$ \frac{dC}{dt} = -v(\frac{dC}{dz})- kC + \frac{F}{L}e^{-z/L} $$
Where:

  - $-v$ is advection rate (cm/yr)
  - $z$ is soil depth (cm)
  - $F$ is total plant carbon inputs (g /cm^2 / yr)
  - $L$ is e-folding depth (cm) ("e-folding" is time interval in which an exponentially growing quantity increases by a factor of e)
  
The first term describes downward advective transport, the second term describes decomposition and the third term describes plant inputs that are distributed exponentially. Conceptually:

  $$ \frac{dC}{dt} = down\ advection - decomposition + plant\ inputs  $$

Modeling carbon movement out of soils:

  - Primary pathway is through production of CO~2~ from decomposition
  - Various forms of Fick's law (law for diffusion) have been applied to describe transport of CO~2~ from soils

Important to consider the work of Cerling (1984) in which he described the carbon-isotope composition of soil CO~2~, which he recognized ultimately controls the isotope composition of pedogenic carbonate.

CO~2~ concentrations in soil typically increase with depth.

Steady-state soil production/diffusion model for CO~2~ can be described as:

  $$ \epsilon\frac{dCO_{2}}{dt} = 0 = D\frac{d^{2}CO_{2}}{dz^{2}} + \phi $$

Where:

  - $\epsilon$ is free air porosity in soil
  - $D$ is the effective diffusion constant
  - $z$ is the soil depth
  - $\phi$ is CO~2~ production
  
Or conceptually:

  $$ porosity * change\ in\ CO_{2} = 0 = net\ diffusion + biological\ production $$

Cerling was interested in identifying processes controlling carbon-isotope composition of soil CO~2~.

Carbon isotopes:

  - Ultimate source is atmospheric CO~2~, which has a C^13^/C^12^ ratio of approximately -7 ppt.

Carbon isotopes are expressed as:

$$ \delta^{13}C(ppt) = (\frac{R_{s}}{R_{std}} - 1)*1,000 $$

Where:

  - R~s~ is the carbon isotope ratio of the sample
  - R~std~ is the carbon isotope ratio of the international standard
    - The standard comes from the Pee Dee Belemnite (PDB), which is a Cretaceous marine fossil from S. Carolina. It has an anomalously high ^13^C:^12^C ratio, and thus most $\delta^{13}C$ values of natural materials are negative.
  
Three photosynthetic pathways have evolved in plants:

  - C~3~ where $\delta^{13}C$ is approximately -27 ppt.
  - C~4~ where $\delta^{13}C$ is approximately -12 ppt (tropical grasses that evolved in response to decreasing CO~2~ levels or strong seasonality/water stress)
  - CAM as an isotopic intermediary between C~3~ and C~4~.
  
Root respiration is believed to account for approximately 50% of CO~2~ flux out of soils.

Soil organic matter commonly shows enrichment of C^13^, as C^12^ typically lost as CO~2~ but C^13^ remains in humic substances.

Quade *et al* (1989) found there were systematic trends in soil CO~2~ concentrations and $\delta^{13}C$ values due to changes in CO~2~ production rates with elevation and soil depth.

Formation of carbonate in soils may be described as:

$$ CO_{2} + H_{2}O +Ca^{2+} = CaCO_{3} + 2H^{+}$$

The use of the pedogenic carbonate and carbon-isotope model has allowed for:

  - Description of the observed increase in $\delta^{13}C$ for soil CO~2~ and CaCO~3~ with depth
  - Mechanistic understanding of why soil CO~2~ is enriched in ^13^C relative to plants (steady-state diffusional enrichment of ^13^C)
  - Can be used in paleoecology, informing on past vegetation complexes
  
*Lateral transport by erosion*

A combination of physical and biological processes, aided by gravity, can drive downslope movement of soil material.

Bioturbation coupled with soil erosion has been shown to be very important in soil formation in some ecosystems/landscapes. Soils that have some plant matter at surface often have high organism activity in the "biomantle" (upper surface), with highly mixed and sorted zones.

Mixing by bioturbation has important physical and chemical implications:

  1. Mixing of soil and loss of stratification at early stages
  2. Quick incorporation of organic matter
  3. Periodic cycling of soil structure, which helps prevent development of horizons within the biomantle
  4. Strongly non-random movements of soils that result in unique landscapes or structures
  
Mass of soil present is a function of soil transport in, soil transport out, and *in situ* soil production. The model to describe this is:

$$ \rho_{s}\frac{\delta h}{\delta t} = 0 = (\rho_{r} \phi) + (K \rho_{s}\frac{\delta^2z}{\delta x^2})  $$

Where:

  - $h$ is the soil thickness (cm)
  - $\rho_s$ and $\rho_r$ are the soil and rock density (g/cm^3^)
  - $\phi$ is the soil production constant (cm/yr)
    - soil production is a function of depth, parent material and environmental conditions; rate of production decreases with increasing soil depth

"Soil residence time" is the idea of how long a soil may exist on a landscape.

In depositional landforms (e.g., active floodplains of major rivers), sediment residence time can be no more than a few hundred years. Soil profiles exhibit stratification, weakly developed horizons that are buried, and little if any chemical weathering.

### Bethke, 2011 {#bethke2011}

*"The thermodynamic ladder in geomicrobiology"* [@bethke2011thermodynamic]

**Key contribution: ** This article gives an overview of the "thermodynamic ladder" in aerobic respiration in soils. They specifically seek to calculate the portion of available energy in soils that different functional groups of microbes can take advantage of to drive their metabolisms.

Ultimately, the usable energy available to various groups changes with water composition and affects microbial reactions.

**Key notes: **

"Thermodynamic ladder" - this is the concept that posits microbial activity in aquifers is organized according to an energetic hierarchy of electron accepting processes. Microbes on "higher rungs" of the ladder are able to maintain an advantage of groups over lower rungs that compete for same electron donor.

In respiration, the species that is initially reduced is the electron donor, whereas the oxidized species is the acceptor. In the process, the electron donor is "oxidized" whereas the acceptor is "reduced", which determines the redox state of the soil system.

Ladder: O~2~ > NO~3~^-^ > Fe > Mn > SO~4~^2-^

*Microbial catalysts*

This particular study constrains its examinations to pristine aquifers, and is most focused on the least energetically favored functional groups: iron reducers, sulfate reducers, and methanogens.

Acetotrophic sulfate reducers - respiring microbes that catalyze electron transfer from acetate to sulfate ions, producing carbonate and sulfide species.

$$ CH_{3}COO^{-} + SO_{4}^{2-} ->  2HCO_{3}^{-} + HS^{-}$$

Iron reducing bacteria will use a variety of other minerals, including ferric oxides and oxyhydroxides and Fe(III) bearing clays as oxidants.

For sulfate reduction and methanogenesis, the amount of H+ ions that are consumed and produced are approximately the same, and thus net reactions for sulfate reductions depend weakly, if at all, on pH

Conversely, reduction of Mn and Fe consumes a large amount of H+ ions, and thus are strongly acid consuming and depend on the pH of the soil.

Most mangrove soils are moderately acidic, which is maintained by frequent inundation and moist character of soil column. Upon drying the soils, they can become extremetly acidic. Thus, Fe reduction is likely to occur in mangrove soils, but will also depend upon the species of Fe present.

*Available energy*

The "more negative the free energy of an electron accepting half-reaction, the more energy is available to microorganisms."

Energy available to Mn and Fe reducers depends on the mineralogic form the metal takes.

Iron reduction is favored in the nominal environment of the paper over sulfate reduction in the form of soil ferrihydrite, but goethite (less fresh?) may be prevalent and thus less available for iron reducers. This ultimately might lead to loss of advantage of iron reducers over sulfate reducers or methanogens.

*Usable energy*

Microbes can only run metabolism where energy available outside the cell exceeds its internal store.

Iron reducers and sulfate reducers capture more energy per mol of electron donor consumed than methanogens. However, in methanogens capturing less energy than iron and sulfate reducers, they can exist in environments in which less energy is available.

*Bioexperiments*

The authors detail a number of bioexperiments, which I have skipped over here.

*Conclusions*

The study fails to find a fixed thermodynamic hierarchy amongst methanogens, sulfate reducers and iron reducers. 

In particular, the amount of usable energy that was seen by the various groups was approximately the same in conditions in which pH was near neutral. Iron reducers exhibit a small advantage where groundwater is acidic.

The experiment found that ecological factors such as population viability and mutualism rather than thermodynamics alone may be important in determining microbial activity.

### Amundson, 2014 {#amundson2014}

*"Soil formation"* [@amundson2014soil]

**Key contribution:** This is an extensive review on the formation of soil that extends Amundson's earlier work in the first volume of the *Treatise on Geochemistry*.

**Structure:** The chapter is structured as follows:

  1. Overview of what a "soil" is
  2. Conceptual partitioning of the earth surface
    1. Stable landforms
    2. Erosional landscapes
  3. Human dimensions of soil formation
  4. Soil chemistry in deserts
  5. Soil formation on Mars

**Key notes: **

*Conceptual issue of what constitutes a "soil"* - different views of soil thickness exist among different scientific communities. For example:

  1. Earth sciences - primarily concerned with the mobile layer of weathering profile in which material no longer reflects parent rock (often restricts to A horizon as B horizon still has parent material)
  2. Geochemical perspectices - interested in the whole weathering process and thus interest extends to more considerable depths

Important to consider the depth at which I am interested for mangroves? This is likely to be the top meter or so of soil, but especially the soil layer in which vegetation health is most dependent upon (more in line with "earth sciences").

*Soil data access* - There are several platforms for which rich data on soils in the US can be collected:

  1. Web Soil Survey - portal for soil maps and lab data
  2. Soil-Web - accesses same database as Web Soil Survey, but uses a different architecture to interface with key technologies (Google Earth, Maps, & smartphones)
  3. Digital soil map of the world -- currently under development but needs more effort (see Sanchez, 2009, Science)
  
*Soil formation* - Can very abstractly be thought of in three classes:

  1. Stable landforms, where derivative of slope = 0
  2. Erosional landforms, where derivative of slope is negative
  3. Depositional landforms, where derivative of slope is positive

**Stable landforms** - Depositional processes can be understood largely through examination of elements relative to that of a parent material. Often this is done by normalizing that element relative to an element that is immobile in the soil weathering environment (Zr, Ti, Nb, etc.).

Often use *chronosequences* to examine changes over time due to inability to perform experiments over extremely long time scales.

Some characteristic steps of soil formation in stable landscapes are outlined:

  - Primary inputs to soil come from organisms; fixing of atmospheric CO~2~ and NO~3~ and NH~4~ bound with amino acids
  - Biotic mechanisms take place and alter the sedimentary structure of the soil (physical mixing)
  - Through chemical weathering, accumulation of secondary clay minerals in the B horizon occurs
  - Clay accumulation and soil thickening occur over long timeframes, volume decreases and bulk density increases resulting in relative "collapse" of soil
  
Whether you consider just the A or B horizon depends on discipline and particular viewpoints.

Within San Joaquin Valley, the "mound and swale" topography is due to the relative increase of clay which creates a higher water table and reduces the A horizon of soil placing pressure on gophers. In response, gophers are hypothesized to undertake directional movement of soil to the tops of mounds.

  - Note, this is the state soil of California (San Joaquin soil series)
  - "Fine, mixed, active, thermic Abruptic Durixeralfs"
  
Important point is that rates of mineral dissolution in almost all soils are transport limited (i.e., it is not the resistance of elements or materials that determine rates, but rather the flow and volume of liquids).

Thus, the rate of weathering is relatively constant with age and weathering fronts may exist at considerable depths.

*Weathering front* - the boundary between the parent rock and either weathered rock or regolith

**Erosional landscapes**

One of the most important conceptual changes in soil formation within the last fifteen years is through modeling of soil transport on hillslopes, particularly through coupling of physical processes with geochemical ones (see Dietrich et al., 2003).

*Denudation* - processes causing wearing away of Earth's surface via moving water, ice, wind or waves; leads to reduction in elevation and relief

The chemical weathering rate can be defined as the difference between the total denudation rate and the physical erosion rate.

  $$ Chemical\ weathering = Total\ denudation - Physical\ erosion$$

The position of the weathering front will depend upon the relative balance between the denudation rate and the propagation rate of the weathering front. If denudation (soil loss) exceeds that of weathering front propagation rate, then the front cant be expected to be shallow, and vice versa.

Denudation is apparently insensitive to rainfall for vegetated landscapes, whereas the weathering front propagation rate is highly sensitive to rainfall levels.

*Human influences*

The scale at which humans are influencing soils is massive. When defining "endangered soils" as those soil series that have lost at least 50% of their extent, considerable portions of soils in the US have been affected (particularly in the mid-West), but also much of the central valley and along the east coast.

The effects of mechanical alteration of soils as well as through the addition of high amounts of water in arid or semi-arid landscapes greatly accelerates chemical weathering processes (e.g., removal of salts, altering of pH, secondary clay formation, etc.). Additional research is needed on the effects of agriculture on undisturbed soils.

*Soil geochemistry in deserts*

Use of isotope research in arid soils reveals that *stable isotopes in soils are rarely conservative tracers of element source*, but rather reflect complex postdepositional processes caused by aqueous or gaseous transport.

An important interpretation of this is that C isotope ratios of soil CO~2~ (and thus any carbonate that forms from it) largely reflects the isotope ratio of plant respiration and the concentration of atmospheric CO~2~. Additionally, C isotope ratios are highly variable along the depth profile.

*Soil formation on Mars*

Investigation of soils in the top 10-15 cm of the surface of mars allows for interesting insight into soil formation and development processes in the absence of water; however, the conclusions of existing work conclude that much of the observation were driven by prior existence of liquids on the surface.

### Soil classification

Classifying soils includes chemical, physical, and biological properties that may be directly observable in the field or quantifiable via lab analyses. The soil taxonomic name that is associated with a particular soil series is built to reflect each of these characteristics.

**Key terms: **

*Pedon* - Smallest three-dimensional sampling unit that displays the full range of characteristic properties of a particular soil

*Polypedon* - A group of closely associated pedons that are commonly found adjacent to another within the field (i.e., may exist as slight variations in one general pedon). A polypedon forms a *soil individual*.

*Soil series* - A class of soil individuals that exhibit the same properties and horizons, though they may be disjunct in space or time

*Diagnostic soil horizons* are often associated with precise quantitative measures and are used to resolutely identify a soil as a particular series or not

Diagnostic soil horizons may occur at the surface or below the surface of the soil. Horizons that occur at the surface of the soil are termed *epipedons*. Seven epipedons exist:

  1. *Mollic* - Dark color rich in organic matter
  2. *Umbric* - Same general characteristics as mollic, but base saturation is much lower
  3. *Ochric* - Mineral horizon that is too thin and light in color to Mollic or Umbric; may be very hard when dry
  4. *Melanic* - Black, organic matter rich horizon characteristic of soils derived from volcanic ash
  5. *Histic* - 20 - 60 cm thick layer of organic soil materials overlaying mineral soil; characteristic of wet areas such as peats
  6. *Anthropic* - Same limits as Mollic epipedon, but has formed through extensive human use (e.g., agricultural soil)
  7. *Plaggen* - Human-made surface layer approximately 50 cm thick made from extensive manuring

The *diagnostic subsurface soil horizons* consist of:

  1. *Argillic* - subsurface accumulation of high-activity silicate clays that have either precipitated downwards or formed in place
  2. *Natric* - silicate clay accumulation but with high exchangeable sodium on the clay complex; common in arid or semi-arid regions
  3. *Kandic* - accumulation of Fe and Al oxides as well as low-activity clays (kaolinite)
  4. *Oxic* - highly weathered subsurface horizon very high in Fe and Al, oxides and kaolinite; found in tropics and subtropics
  5. *Spodic* - illuvial horizon (material displaced by rainwater from one horizon to another) characterized by accumulation of colloidal organic matter and Al oxides; common in highly-leached, sandy cool humid climates (think pine forests)
  6. *Sombric* - illuvial horizon dark in color due to high OM accumulation; found mostly in cool, moist plateaus and mountains of tropics and subtropics
  7. *Albic* - light-colored eluvial (formed in place) low in clay and Fe and Al oxides which have been transported downwards
  8. *Calcic* - accumulation of carbonates (white chalk-like nodules)
  9. *Gypsic* - accumulation of gypsum
  10. *Salic* - accumulation of soluble salts
  
*Soil moisture regimes*:

  1. *Aquic* - soil saturated with water
  2. *Udic* - moisture sufficiently high to meet plant needs
  3. *Ustic* - intermediate moisture between udic and aridic; significant periods of drought may occur
  4. *Aridic* - soil is dry for at least half of the growing season and moist for less than 90 days (torric indicates the same soils that may be hot in summer but not hot in winter)
  5. *Xeric* - moisture regime common to Mediterranean climates; cool, moist winters and warm, dry summers
  
The *soil temperature regimes* primarily refer to:

  1. Frigid (cold)
  2. Mesic (moderate) 
  3. Thermic (hot)

**Soil taxonomy:**

  1. Order - Defined as one of 12 orders
  2. Suborder - Often integrate the moisture regime of the region (e.g., Aquent)
  3. Great group - Often indicate presence of a diagnostic horizon (e.g., Sulphaquent)
  4. Subgroup - Further define diagnostic horizon to add nuance to the classification (e.g., Halic Sulphaquent)
  5. Family - identifies subsets of subgroups that are similar in texture, mineral composition and mean soil temperature at depth of 50 cm
    - e.g., "Typic Argiaquolls, fine, mixed, mesic, active"
    - Typic Argiaquoll that is fine in texture, mixed clay mineral content, mesic in temperature (8 - 15 deg C), and with clays active in cation exchange
  6. Series - named after geographic feature where they were first identified
  
There are 12 *soil orders* globally:

  1. *Alfisols* - argillic, natric or kandic horizon; high to medium base saturation (alf)
  2. *Andisols* - soil formed from volcanic ash (and)
  3. *Aridisols* - dry soils with ochric epipedon (id)
  4. *Entisols* - little profile development, ochric epipedon is common (ent)
  5. *Gelisols* - permafrost; frozen soils (el)
  6. *Histosols* - peatland or waterlogged soils with > 20% OM (ist)
  7. *Inceptisols* - embryonic soil with few diagnostic features; ochric or umbric epipedon (ept)
  8. *Mollisols* - grassland soil rich in organic matter with mollic epipedon (oll)
  9. *Oxisols* - oxic horizon, highly weathered (ox)
  10. *Spodosols* - spodic horizon common with Fe and Al oxides (od)
  11. *Ultisols* - argillic or kandic horizon, low base saturation, very weathered (ult)
  12. *Vertisols* - high in swelling clays, cracks when dry (ert)
  
Mangrove soils are likely to be *Aquents* (frequent saturation with water) or Fluvents (parent material of recent alluvium, more commonly associated with floodplains)

  - likely to be "halic sulphaquents," "sulfic halaquents", or "halic sulphaquepts"
  
*Suborders* commonly integrate the moisture regime of the region into the soil order (e.g., Aquent).

More than 400 *Great Groups* exist, defined primarily by presence of diagnostic horizons, which supply formative elements for the names (suphic diagnostic horizon in Sulphaquents, halic diagnostic horizon in Halaquents)

*Subgroups* - central concepts of a Great Group makes up one subgroup (*Typic*). Others introduce secondary diagnostic characteristics that better situate the soil between different Great Groups. (e.g., Sulfic Halaquents as an Aquent with Halic diagnostic properties but also Sulfuric properties).

*Families* - Approximately 8,000 families have been identified; integrate information on the physical and chemical properties affecting the growth of plant roots.

  - Information on particle size, mineralogy, CEC activity, temperature, and depth of soil

### Soil architecture

**Soil color** 

Soil color is classified using the Munsell charts and is a function of:

  1. Hue - Redness or yellowness
  2. Chroma - Lightness or darkness
  3. Saturation - Intensity of brightness of the color
  
Soil color is indicative of many properties of the soil, including presence of organic matter, moisture content, or presence of Fe or Mn oxides.

*Gleying* refers to reduction of Fe or Fe depletion from the soil and is indicated by a gray soil color.

**Soil texture** 

Soil texture ranges greatly (e.g., from clay to boulders), but often we refer to the fine-earth fraction (sand, silt & clay) in referring to soil texture.

  1. *Sand* - Smaller than 2 mm in size but greater than 0.05 mm; most common material constituent is quartz (SiO~2~), or other silicate materials
  2. *Silt* - Between 0.05 mm and 0.002 mm; Similar in shape and mineral composition to sand but much smaller in size and thus drains water more slowly
  3. *Clay* - Less than 0.002 mm; Extremely large surface areas which results in great propensity to hold water
    - The finest sizes of clays act as colloids (i.e., where particles suspended in a liquid do not readily settle out)
    - More commonly flat and plate-like than round which further inhibits passing of liquids and gasses

*Textural classes* - Soils exist on a continuum of texture, but can be classified generally given their relative percent composition of sand, silt or clay

Central concept of a *loam* is defined as a mixture of sand, silt and clay that exhibits properties of the three in approximately equal proportions.

Adding peat or compost to soils does not technically change the texture, as the texture criteria is only referring to the mineral particles.

**Soil structure** 

Soil particles aggregate together to form *peds* or *aggregates*, which greatly influence the structure and functioning of the soil.

*Peds* may vary in size from 1 mm to 1 m; Commonly used to refer to the whole structure of the soil (as aggregates exist at different scales), but is synonymous with "aggregate". Peds often vary depending on the horizon in which they are found. The principal ped shapes are:

  1. Spheroidal
  2. Platy
  3. Prismlike
  4. Blocklike
  
Soil aggregates consist of macroaggregates that are a combination of microaggregates. Both biological and physical-chemical processes are important in their formation, though physico-chemical processes matter primarily for microaggregates whereas biological processes are more critical for macroaggregates.

*Flocculation* is the mutual attraction between clay and organic molecules, and is key for formation of microaggregates (particularly for sedimentation in mangroves).

  - Form through a covalent bond in which positive charged ions compressed against one another attract negative charges on the opposite's platelet.
  
Organisms (e.g., earthworms) and plant roots are also important in forming of casts through consumption of soil as well as production of sticky substances (glomulin) that are important in forming aggregates.

**Soil density** 

There is a key distinction between particle density and bulk density that should be noted. In particular:

  - Particle density is defined as the mass of the soil particles per unit volume of *soil solids* (i.e., no pore space is considered)
    - Particle densities vary little and are typically between 2.6 - 2.75 Mg/m^3^
  - Bulk density  is the mass of the soil particles per unit volume of soil (including the pore spaces)
    - Bulk density values vary to a much larger degree
    
One of the key reasons for measuring bulk density is for the calculation of *total porosity* (i.e., through comparison of particle density and bulk density). Provides a measure of total pores, but pores are comprised of both *macropores* (> 0.08 mm) as well as *micropores* (< 0.08 mm).

  - *Macropores* readily allow movement of gas and water within the soil, as well as growth of plant roots and movements of soil organisms.
  - "Micropores" are typically filled with water. Even when not filled with water, they are small enough to restrict passage of gasses. Water in these pores is restricted to plants

### Soil hydrology

**Key notes**:

*Properties of water molecules* - Ability of water to influence soil processes is largely due to the structure of water. In particular:

  - Polarity - The H atoms of water line up in a V-shape of 105 degrees which induces a polarity to the molecule.
  - Hydrogen bonding - Hydrogen atom of one water molecule is attracted to the H of another water molecule forming a low-energy bond that results in polymerization of water. Hydrogen bonding thus accounts for the following behaviors of water
    - Cohesion - Adherence of water molecules to one another
    - Adhesion - Adherence of water molecules to the surfaces of solid particles
    - Tension - Results from greater attraction of water molecules to one another than to gases, which results in tight bonding of water molecules at the air/water interface
    
Capillary action is important for the movement of water within soils. Capillary action occurs through the dual action of adhesion and cohesion.

*Soil water energy concepts* - Various forces come into play in determining movements of water in the soil.

Water transfers from states of high energy to low energy, and various forces that determine the potential energy of water are important:

  - Matric forces -- forces caused by adhesion that are responsible for adsorption and capillarity.
  - Osmostic forces -- forces driven by differences in ion concentrations in solutes
  - Gravity -- self explanatory; constant downwards force on water
  
*Water potential* is determined as the difference in potential energy between water within a given soil at a particular location with water at a reference location at standard pressure and temperature with no effects of soil.

Total soil water potential is a function of:

  1. Gravitational potential
  2. Matric potential -- extremely important for supply of water to plant roots
  3. Hydrostatic potential -- pressures from the weight of water in saturated soils
  4. Osmotic potential 
    - Important for mangroves
  5. Other more minor potentials...
  
Matric and hydrostatic potentials are part of the "pressure potential," whereas gravitational and osmotic are considered alone.

*Soil water content and water potential* - An inverse relationship exsists between the water content of soils and how strongly water is held within soils.

In describing water content of soil, need to describe both i) the water content as well as ii) the water potential. Behavior of water in soil is typically determined primarily by water potential rather than water content.

"Water content" is defined as a volume of water per volume soil. Typically measured as difference between wet and dry weights of soil.

"Water potential" can be measured in different ways; Matric potential can be measured via a tensiometer, which is a semi-porous rod filled with water that allows water to flow out of the rod and creates a vacuum within the rod. Tensiometers are useful over a range of 0 to -85 kPa.

*Flow of water* - Flow can be broken into three primary types:

  1. Saturated flows - Saturated flow occurs when soil pores are completely filled. Hydrostatic potentials play a part here and the hydraulic conductivity is largely determined by macropore spaces
    - Flow rates are proportional to the fourth power of the radius, and thus macropore space greatly determine flows during saturated flow.
    - Preferential flow is the faster flow of water through biopores (e.g., pores and channels created by organisms) or large cracks
  2. Unsaturated flows - Large pores are free of water but smaller pores hold water
    - Primary driver of water flow is "matric potential gradient"
    - Soil texture is highly important as clays will moderate flows whereas sandy loams are more "flashy"
  3. Vapor movements - Result of vapor pressure differences in relatively dry soils
  
*Infiltration and percolation* - Texture of soil and in particular presence of macropores are critical here.

*Qualitative descriptions of soil wetness* - Several classes of soil wetness can be described:

  1. Maximum retentive capacity - all pore spaces are filled, matric potential is close to zero and gravitational forces dominate
  2. Field capacity - soil will drain via gravitational forces within 1 - 3 days, and matric forces will dominate (generally a range of -10 to -30 kPa). Field capacity is an important condition given:
    1. Maximum level of water available to plants is present
    2. Soil is at optimal "pliability" in that below field capacity it will begin to crumble and above that it will easily turn to mud
    3. Sufficient pore space is available to allow aeration for microbial activity and plant growth
  3. Plant available water - from 10 - 30 kPa to 1500 kPa - water potentials at which water is largely available to plants
  4. Permanent wilting percentage - point at which plants wilt during both day and night due to inability to extract water from soil (<1500 kPa)
  5. Hygroscopic coefficient - point at which water is only available in vapor form due to extreme aridity

The amount of organic matter within a soil can influence the amount of plant available water greatly. Organic matter has a high water-holding capacity, particularly relative to mineral soils.

"Least limiting water range" - Saturation of soil may produce anoxic conditions which restrict root growth. Often times this is the point at which only 10% of pore space is filled with air. Additionally, soils may become too dry for plant roots to push through them during growth (2000 kPa)

*Relevance for mangroves* - My own interpretation of how this might matter for mangroves is as follows:

For mangroves, matric potentials and osmotic potentials likely drive water flows within mangroves. Given the extremely find sediment of many mangrove forests, there is likely very little pore space for which water and air may enter the soil. Water flushing and aeration may only exist within top 15-20 cm of soil.

### Soil aeration

Aeration is the process of soil ventilation, with gases moving both into and out of the soil. Requires both the supply of oxygen as well as the removal of CO~2~ from the soil.

Three primary processes control the availability of oxygen within soils:

  1. Soil macroporosity
  2. Soil water content
  3. O~2~ consumption

Plants that have adapted to waterlogged soils (i.e., in wetlands) are *hydrophytes*, and have *aerenchyma* which are hollow tissues used for the transport of oxygen to the roots.

Aeration of soil occurs through two primary mechanisms:

  1. Mass flow -- much less significant on total exchange of gases
  2. Diffusion -- most significant process; gases moving in response to their partial pressures
    - partial pressures is the pressure a particular gas in a gas mixture will exert
    - O~2~ will generally move into soil whereas water vapor and CO~2~ will typically move from soil to the atmosphere
    
There are three primary ways of characterizing aeration of soil:

  1. Directly describing the gas constituents in the soil
  2. Air-filled soil porosity (i.e., percentage of soil pores that are filled with air)
  3. Redox potential as a measure of the oxidation and reduction of different elements (i.e., a measure of aerobic metabolism by organisms)
  
Reduction-oxidation (redox) describes the chemical processes by which electrons are transferred. Reduction and oxidation occur together, as electron is transferred from one (the reductant) to the other (the oxidant). Micro-organisms drive majority of redox reactions, often consuming electrons from C or accepting electrons from oxygen. Respiration is largely the consumption of C based electrons from carbohydrates.

Basic example of respiration in which electrons are derived from an organic molecule:

$$CH_{2}O + H_{2}O = CO_{2} + 4e^-$$

The electrons cannot exist in independent form, and thus are run through metabolic cycles in organisms and dumped onto electron acceptors (e.g., O~2~).

Oxygen is important within soils as it rapidly accepts electrons from other elements, which allows for metabolic processes. Aerobic respiration is performed on O~2~ (oxidize organic carbon) and release energy for metabolic activity.

The redox potential of a soil is a measure of how well much oxidation-reduction it is able to perform (i.e., a measure of its ability to accept or donate electrons), and depends upon presence of electron acceptors as well as pH. Redox potential typically decreases as the pH increases.

*Oxidizing soils* - accept electrons easily
*Reducing soils* - donate electrons easily

Standard Eh levels for drained, well-aerated soils is in the 0.4 - 0.7 Volt range, whereas flooded conditions can reduce Eh levels to as low as -0.3 V.

Other electron acceptors besides oxygen are:

  1. Nitrate - accepts 2 electrons when reduced from N(V) to N(III); reduces to N~2~
  2. Manganese (VI) -> reduces to Mn(II)
  3. Iron (III) -> reduces to Fe(II)
  4. Sulfur (from SO~4~^2-^ to S~2~^-^ (reduced))
  5. Carbon (reduction of CO~2~ produces CH~4~)
  
Electron acceptors are typically used up in this order (i.e., sulfur reduction is a process that occurs largely under very anaerobic soils). Anerobic systems result in the build up of reduced species

The form of particular elements is important because it will change availability for uptake by plants (i.e., oxidized vs. reduced states). 

*Factors that affect soil aeration:*

  - Presence of macropores -- most important factor in well-drained soils; function of soil texture, bulk density, aggregate stability, OM content, and biopore formation
  - Microbial activity -- rates of respiration in the soil will greatly influence gas composition, and is dependent upon availability of organic matter
  - Soil heterogeneity -- lower oxygen concentrations in subsoils; often due to slow diffusion, reduced pore space (from compaction), and greater proportion of pore space filled by water
    - plant roots may reduce or increase O~2~ content (diffusion of O~2~ from root aerenchyma)
  - Seasonal differences
    - Changes in precipitation patterns or temperature
    
*Ecological effects of soil aeration:*

  - Nature and rates of microbial activity are determined largely by O~2~ concentrations
  - Redox of elements:
    - Determines form and thus availability of elements, which may be beneficial or toxic
    - Methane production, which occurs at Redox potentials of approximately -0.2
  - Activities of higher plants:
    - Root growth may be curtailed
    - Absorption of water and nutrients may be decreased
    - Formation of certain inorganic compounds may occur
  
*Soil temperature*

Growth rates of plants are more strongly controlled by soil temperature than the air temperature. Biological zero is 5 deg Celcius, in which microbial activity virtually ceases.

Impacts of frost-heave and fire on soils can be significant. Effects of fire typically limited to the top 1-2 centimeters, but can be intense enough to alter organic matter, soil conditions, and kill soil organisms.

Temperature of soil in the field is largely dependent upon three factors:

  1. Net amount of heat energy the soil absorbs
  2. Heat energy required to bring change in the soil temperature
  3. The energy required for processes such as evaporation, which occur primarily near soil surface
  
Only approximately 10% of solar incidence is actually transferred to soil. Majority is intercepted within atmosphere, by vegetation, or is loss in the evaporation of soil water.

Dark soils do not necessarily warm faster than light soils due to the fact that dark soils may often be very moist and thus solar incidence is lost through evaporation of soil moisture.

*Thermal properties of soils*

The specific heat (energy required to raise the temperature of a given mass of something by 1 degree C) is much higher for water than for mineral soil solids. Thus, evaporation of water requires a very large amount of energy (2.257 MJ per kg of water).

Thermal conductivity refers to the rate at which heat is tranferred through soil, and is dependent primarily on:

  1. Soil moisture content - heat passes through water much faster than through air
  2. Degree of soil compaction - heat passes through mineral particles even faster than through water, thus when compaction and particle-to-particle contact is high, heat is transferred quickly
  
Wet compact soils are the poorest insulators (i.e., best conductors of heat)

Given seasonal and daily variation in temperatures and the low thermal conductance of soils, there is a temporal lag between the soil temperature and atmospheric temperatures 

### Soil chemistry

The soil colloids are arguably the most important component of soil chemistry, as they control the majority of exchange of cations and anions within the soil.

The colloidal fraction of soil refers to the clays and humus particles in soil because of their extremely small size (< 1 micrometer) and colloid-like behavior.

The very large surface area to volume ratio of soil colloids is due to the platelike structure, and the present of internal surfaces within clays. The majority of soil colloids primarily carry more negative electrostatic charges than positive charges.

*Absorption of cations and anions:*

Colloids attract thousands of Al^3+^, Ca^2+^, Mg^2+^, K^+^, ^H+^, and Na^+^, which are loosely bound to the surface of the colloid, often in a hydrated state in moist soils.

"Cation exchange" is the process in which one cation breaks free from the colloid, is released into solution, and a different cation takes its place. The ions that are readily lost from the colloid surface are referred to as "exchangeable ions".

The ionic composition of soil solution (i.e., the water) resembles that of the adsorbed ions (those bound to the colloid) due to the constant exchange between the two.

Exchangeable anions are also present in the soil solution, and will exchange with soil colloids with positive charges on their surfaces.

*Types of soil colloids:*

  - Crystalline silicate clays -- dominant in most soils; layered structure like pages in a book; predominantly negatively charged; different silicate clay minerals vary widely in particle shape, however, which influences intensity of charge, plasticity, swelling behavior, etc. 
    - Examples: smectite, vermiculite, fine mica, chlorite, kaolinite (1:1 silicate)
  - Noncrystalline silicate clays -- tithgly bonded silicon, aluminum and oxygen atoms, but do not exhibit crystalline sheets; high amounts of positive and negative charge, and capable of holding large amounts of water, but exhibit little stickiness
    - Principle types: allophane and imogolite (both form from volcanic ash)
  - Iron and aluminum oxides -- found widely but are most important in weathered soils of sub-/tropics; slightly negative to moderately positive and are relatively low in plasticity and stickiness
    - Examples: gibbsite (Al-oxide) and goethite (Fe-oxide) which form crystalline sheets; others are noncrystalline
  - Organic (humus) -- convoluted chains and rings of carbon atoms bonded to H, O, and N; very high capacities to hold water, but little to no plasticity and stickiness
  
A summary of the major colloids and their properties is shown below:

```{r}
knitr::include_graphics(rep("images/colloids.bmp"))
```

*Layer silicate clay structure:* 

Silicon tetrahedral and Aluminum-Magnesium octahedral sheets:

  - Silicon tetrahedral sheets are comprised of a silicon ion surrounded by four oxygen atoms
  - Al-Mg octahedral sheets are comprised of an Al or Mg ion surrounded by 8 oxygen-hydroxy planes
  
Two to four of the tetrahedral and octahedral sheets will bind together by sharing some of the same oxygen atoms, which results in extremely tight bonds and forms a "layer". The layers then are bound together via adsorbed ions and water, which forms the clay "crystal".

The nature and combination of sheets in the layers vary, and thus produce the differences in types of clays (ant thus their resultant physical and chemical properties).

"Isomorphous substitution" is the process by which cations of comparable size to Si, Al, or Mg are substituted for the center of one of the tetrahedrons or octahedrons without significantly changing the structure of the crystal. This is the key process that induces variability in nature of silicate clays.

Isomorphous substitution is the primary cause of charge imbalances in colloids (e.g., through the substituion of a Mg^2+^ for a Al^3+^ ion).

Key difference is between 1:1 and 2:1 silicate clays:

  1. 1:1 silicate clays contain one tetrahedral and one octahedral sheet each in their layers
  2. 2:1 silicate clays contain one octahedral sheet in between two tetrahedral sheets within each layer

Kaolinite is the classic example of 1:1 silicate clays. Bonding between layers (1 tetra and 1 octa) is via hydrogen bonding, which prevents expansion when the clay is wet. Cations and H~2~O typically do not enter the structural layers of a 1:1 mineral particle and thus effective surface of kaolinite is primarily the external surface area.

Expanding 2:1 type silicate clays are formed via one octahedral sheet in between two tetrahedral sheets (e.g., smectite and vermiculite). Fine micas and chlorite, on the other hand, are relatively nonexpanding.

Smectite - predominantly have negative charges resulting from isomorphous substitution. Layers are bound via an oxygen-oxygen surface and thus the bond is weak allowing for cations and water to enter the spaces in between.

Vermiculite -- tetrahedral sheets have large amounts of Al substitution for the silicon position, which produces a CEC higher than other silicate clays

Fine-grained micas (e.g., illite and glauconite) -- very weathered micas with substitution of Al^3+^ for Si^4+^ in the tetrahedral sheets. The negative charge attracts K^+^ ions which strengthens bond between the layers and making the fine-grained micas nonexpansive. 

Chlorites -- Fe or Mg primarily occupies the place of Al in the octahedral sheets. Mg dominated octahedral sheets commonly sandwiched between two 2:1 layers, thus often said to have a 2:1:1 structure.

*Non-silicate colloids*

Fe and Al oxides are motified octahedral sheets with Fe or Al in the cation positions, and no tetrahedral sheets or silicon in their structures. 

Humus is noncrystalline and is formed of large organic molecules of varied chemical composition. Negative or positive charges on humus colloid are a result of loss or gain of H^+^ ions; a very large net negative charge is associated with humus, though sites of both positive and negative charges may exist

*Formation of colloids*

Humus is primarily formed through the breakdown or organic matter (mostly plant matter) by microorganisms and the resynthesis of this material into stable organic compounds.

Silicate clays develop through:

  1. Physical and chemical alteration of primary minerals
  2. Decomposition of primary minerals with recrystallization of products into silicate clays
  
Kaolinites and Fe and Al oxides are found primarily in warm, humid regions (e.g., tropics), whereas smectite, vermiculite and fine-grained micas are primarily in regions where weathering is less intense.

*Sources of charges on soil colloids*

Two major sources of charge imbalances:

  1. Hydroxyls and other functional groups on surfaces of colloids releasing or accepting H^+^ ions
  2. Charge imbalances as a result of isomorphous substitutions
  
Charges of all colloids are affected to a certain degree by OH^-^ groups, which are largely pH dependent. The charges of Al and Fe oxides, 1:1 type clays, allophane and humus are primarily determined by OH^-^ groups.

For 2:1 clays, surface charges are complemented by isomorphouos substitutions of cations. Charges from isomorphous substitution are not dependent upon pH, and thus are considered "permanent"

*Cation exchange reactions:*

Ions held to external surface of clays via electrostatic attraction, which is relatively weak and thus ions are in constant motion. There are points at which ions are closer versus slightly further away from the colloid. If an ion in soil solution enters space closer to the colloid, the ions "exchange".

Cation exchange reactions are reversible in that the cations can easily revert back to their original places.

Cation exchange reactions occur on charge to charge basis (i.e., 2 H^+^ for 1 Ca^2+^).

Ratios of cations within soil solution will match those of cations bound to colloids at equilibrium.

"Mass action" in soil influences the direction at which reactions may occur. If cations released from a colloid *precipitate*, *volatilize*, or *strongly associate* with other element in the solution (e.g., the C and O atoms of a CaCO~3~ molecule), then they will not exchange back to the colloid.

Different cations are held at different strengths to the colloid. Al^3+^ is held most strongly, whereas Li^+^ is held the weakest.

The likelihood of a given cation being replaced is dependent upon the neighboring cations and the strength of bond (i.e., a Mg^2+^ in between 2 Al^3+^ ions is more likely to be replaced than Al).

*Cation Exchance Capacity (CEC)* - Number of centimoles of positive charge that can be adsorbed per unit mass; measure is on a charge for charge rather than ion for ion basis.

Measurement of CEC via pH buffered solutions will yield measures of the *potential* or *maximum* CEC, whereas measures taken at the pH of the soil will only measure the *effective CEC*, which corresponds to the exchangeable cations.

*Exchangeable cations in field soils* - Varies depending on the environmental conditions of the site.

In humid regions, Ca^2+^, Al^3+^, aluminum hydroxy ions and H^+^ are most prominent, whereas Ca^2+^, Mg^2+^, and Na^+^ dominated in more arid regions.

"Saturation percentage" refers to the percent of the exchange complex held by a particular ion. Useful in identifying acid percentage saturation (H^+^ and Al^3+^) versus nonacid (base) saturation percentage.

Exchangeable cations influences the fertility of the soil, as many of the cations are common micronutrients for plants. Percent saturation of a given cation will describe its ready availability within the soil for plants to a certain degree.

*Anion exchange* - Anions are held by colloids in two primary ways:

  1. Held by anion adsorption mechanisms similar to those of cation adsorption
  2. Reactions with surface oxides or hydroxides, which forms inner-spere complexes
  
Positive charges associated with kaolinite surfaces attract anions such as SO~4~^2-^ and NO~3~^-^. Adsorption exchanges typically decrease with increasing pH for anions.

Inner-sphere complexes form when an anion replaces part of a hydroxyl group that is adsorbed by a cation on a colloid.

Anion exchange is important as it determines the mobility of certain important anions within the soil.

### Soil pH

Principle processes that determine soil acidification are:

  1. Production of H+ ions
  2. Washing away of cations by percolating water
  
Both processes are facilitated by volumes of water that percolate through soil, and thus acidity is closely related to amounts of precipitation within the region.

H^+^ ions are produced through dissociation of carbonic acid, which forms through CO~2~ dissolving in H~2~O.

$$ CO_{2}\ +\ H_2O\ \to\ H_2CO_3\ \iff\ HCO_3^{-}\ +\ H^+$$

Production of H^+ from H~2~CO^3^ is only significant for pH levels greater than 5.

In addition, several biological processes can affect the acidity of the soil:

  1. Plant balancing of cations - when excess of certain cations are taken up by roots, H^+^ often released into soil solution so maintain charge balance
  2. Accumulation of OM - cations can form soluble complexes with nonacid nutrient cations (e.g., Ca^2+^ and Mg^2+^), which may result in leaching; nonacid cations may be replaced with acid cations
  3. Oxidation of nitrogen (Nitrification) - reaction of ammonium (NH~4~^+^) with oxygen releases H^+^ ions.
  4. Oxidation of sulfur - oxidation of sulfur yields H~2~SO~4~, which is a strong acid
  
*Alkalizing processes* - Reduction of H^+^ or production of OH^-^ in the soil alkalizes the soil. Several pathways of alkalizing processes exist:

  1. Weathering of nonacid cations from minerals -- weathering of parent material may result in bonding of H^+^ and release of nonacid cations, such as Ca^2+^
  2. Accumulation of nonacid cations -- nonacid cations from weathering may accumulate in areas of low precipitation (Ca^2+^, Mh^2+^m K^+^, and Na^+^)
  3. Production of base-producing anions -- basic, hydroxyl generating anions are carbonate (CO~3~^2-^) and bicarbonate (HCO~3~^-^); these react with water to form Oh^-^ and raise pH (important for *soil buffering*)
    - The direction of carbonate reactions depends largely on the presence of CO~2~ in the soil, with greater amounts of CO~2~ acidifying the soil
  4. Excess anion uptake by roots -- roots excude bicarbonate when excess of anions are taken up by roots
  
*Role of aluminum*

Aluminum may have a strong acidifying effect on soil, largely due to its tendency to hydrolyze (split water into H^+^ and OH^-^) ions.

*Pools of soil acidity* - There are three major pools of acidity within soils:

  1. Active acidity - amount of H^+^ ion in the soil solution 
    - very small relative to other two pools
    - determines solubility of substances and provides solution environment for organisms
  2. Exchangeable acidity - associated with exchangeable Al^3+^ and H^+^ ions present in large quantities in acid soils 
    - can be released by cation exchange with unbuffered salts and thus is sometimes referred as salt-replaceable acidity
  3. Residual acidity - hydrogen and aluminum ions bound in nonexchangeable forms by organic matter and clays 
    - will dissociate with increase in pH
    - largest pool of acidity (10 - 100 thousand times greater than active acidity)
    - the amount of bound hydrogen and aluminum ions is much higher for soils rich in OM and clays

The relative amounts of exchangeable versus bound acid and nonacid cations within the soil is largely dependent upon pH. Exchangeable nonacid cations increases as pH increases.

Percent saturation of acid vs. nonacid cations refers to the relative charge percentage accounted for by acid vs. nonacid cations.

"Potential" CEC is the CEC at high pH values, whereas "effective" CEC is the CEC at the pH of the soil.

*Soil buffering* - The ability of a soil to resists changes in pH

Mechanisms of buffering are largely explained by the equilibrium between the different pools of soil acidity: active, salt replaceable and residual. Largely as a result of uptake or release of acid ions on the exchangeable complex of colloids.

In general, the higher the CEC, the higher the buffering capacity of the soil.

*Human influences on soil acidification* - Although humans may influence soil pH in many ways, three primary types exist:

  1. Nitrogen amendments - H^+^ is released upon oxidation of ammonium; often need to counter-balance additions of OM with additions of lime
  2. Acid precipitation - effects of acid rain due to combustion of fossil fuels
    - addition of strong acid anions (SO~4~^2-^ and NO~3~^-^) will bond with nonacid cations and allow for their leaching from soil; aluminum and hydrogen then become dominant on exchange complex
  3. Exposure of potential acid sulfate soils - disturbance of soils (and thus introduction of O~2~) into soils with high concentrations of sulfur can produce large amounts of acidity
    - common in wetland soils in which constant supply of sulfur is available
    - commonly contain high amounts of pyrite (FeS~2~), iron monosulfides (FeS), and elemental sulfur (S)
    - pH may be neutral but quickly shift to extremely acidic a few days after oxidation
    
*Biological toxicity* - Acidic soil conditions can create toxic environments for plants and microbes:

Aluminum is the most significant toxicity for plants, which is taken up in roots and restricts growth and functioning. Aluminum is only toxic at pH levels below 5.3

Manganese may also become toxic under acidic conditions as it is increasingly soluble and taken up in excess quantities. Mg is an essential micronutrient but becomes toxic in excess quantities

Nutrient availability in general is greatly influenced due to varying degrees of availability that depend upon pH. Several key macronutrients may be greatly limited under acidic conditions

Fungi do well across a broad range of pH values, and thus tend to dominate in acidic soils (whereas bacteria play a larger role in more neutral soils)

*Liming of soil* - Majority of liming materials raise the pH of the soil by adding oxide, hydroxide, or carbonate that reacts with carbon dioxide and water to yield calcium (or Mg) and bicarbonate. The abundance of CO~2~ in soils is typically enough to drive this reaction.

The calcium will exchange with acid cations on the exchangeable complex, which bind to hydroxides from the bicarbonate and thus reduces the percent acid saturation.

*Alternative methods to deal with acid soils* - In some cases, application of liming material is not efficient (cost restrictions, acidic soil is below surface horizon, etc.) and alternative options of dealing with ill effects of soil acidity are needed:

  1. Gypsum - ameliorates aluminum toxicity without increasing soil pH
  2. Organic matter - application of organic wastes, cover crops, or mulching helps to:
    1. humified OM may bind with Al ions
    2. organic acids can form soluble complexes with Al ions that are nontoxic to plants and microbes
    3. substantial amounts of Ca may be available that can raise pH levels and also precipitate easily down the soil profile
  3. Selecting adapted plants

*Lowering soil pH* -- Acidifying soil can be done through additions or organic or inorganic acid-forming materials. Materials from trees common to spodosols (e.g., pine needles or pine sawdust) may be effective at raising the soil pH. Additionally, inorganic chemicals may be used.

**Saline soils** - A large section exists in the chapter on saline soils, but is mostly related to agriculture and thus may not be of particular relevance here.

Measuring salinity in soils:

*Electrical conductivity* - water is a poor conductor, but salt improves its conductance; can measure conductance as an indirect measure of salinity (units: deciSiemens per meter)

*Exchangeable sodium percentage (ESP)* - the degree by which the exchangeable complex is saturated with sodium.

*Sodium adsorption ratio (SAR)* - concentration of sodium relative to calcium and magnesium

Classes of salt-affected soils include *saline, saline-sodic* and *sodic*. 

  - saline soils -- have an ESP less than 15 or EC values greater than 4 dS/m
  - saline-sodic soils -- EC greater than 4 dS/m AND ESP greater than 15 (intermediate conditions between saline and sodic)
  - sodic -- EC less than 4 dS/m, ESP greater than 15, and SAR greater than 13
    - most problematic of saline soils
    - extremely poor soil physical conditions and slow permeability of water and air

Physical degradation of sodic soils can be significant, and is largely measured in terms of ready movement of water (saturated hydraulic conductivity). In some sodic soils, K~sat~ may be reduced almost to 0 and puddles form on the soils surface.

Low permeability related to sodic conditions is due to:

  1. exchangeable sodium increases tendency of aggregates and floccules to break up when wet
  2. some expanding-type clays become highly Na+ saturated, which increases swelling capability
  3. sodic conditions may lead to soil dispersion, which is caused by:
    1. high sodium - have single charge and large hydrated size, which means they are held loosely and at far distances from soil colloids
    2. low salt concentration - low presence of ions in solution causes exchangeable ions to be spread out over greater distance and thus weakens bonds (anions are found closer to colloids)
    
In general, low salt concentrations and weakly attracted ions encourage soil dispersion and puddling, whereas high salt concentrations and strongly associated ions (calcium) promote clay flocculation and soil permeability.

### Soil organisms {#soilorganisms}

Soil organisms range from macrofauna to microfauna, and tend to aggregate in zones of favorable conditions rather than be spread out evenly through the soil.

A range of different classifications for soil organisms, but often consider trophic level (primary producer, primary consumer, etc.) or feeding patterns (fungivore, herbivore, autotrophic, etc.).

*Primary consumers* - A variety of primary consumers exist within the soil environment:

  - herbivores - feed directly on living plant tissue, for example rodents or larvae feeding on plant roots
  - detritivores - vast majority of soil organisms; feed on dead decaying tissues of plants on soil surface and within soil pores
    - "shredders" - those organisms that actually feed on organisms on decaying material (e.g., bacteria) rather than the material itself; important for shredding & processing of detritus
  - saprophytic - feeding on dead tissues by fungi and bacteria (microflora); accounts for majority of decomposition of dead plant and animal debris

*Secondary consumers* - Include microflora as well as carnivores

  - carnivores -- may include centipedes, mites, insects, nematodes, spiders, snails, etc.
    - feed on microflora and thus are important for processing of litter
    
*Tertiary consumers* - Ants, bird, or other organisms that consume the soil organisms

*Abundance of soil organisms*:

In general, the ratios of soil OM to detritus to microbial biomass to faunal biomass is 1000:100:10:1.

Total living biomass is generally well-correlated with the dry-weight or organic matter, oftentimes approximately 1-5% of the dry weight of organic matter.

Majority of activity in soils is by microorganisms (estimated at ~80% of total soil metabolism). Nematodes and protozoa play important roles in preying on bacteria and fungi, whereas earthworms can be significant and play an important role as ecosystem engineers.

*Plant roots* - Typically occupy approximately 1% of soil volume and may also be responsible for 25% - 33% of the respiration occuring in the soil.

Rhizosphere extends approximately 2 mm out from the root surface. Soil conditions within the rhizosphere can be significantly different from that of the surrounding soil.

*Fungi*:

Most versatile and persistent of any group of OM decomposers within soils. Extremely efficient nutrient cycling, with up to 50% incorporation into fungal tissue (vs 20% for bacteria). Thus, largely important for soil fertility.

Mycorrhizae - "root fungus" or an association of certain species of fungus with plant roots that greatly expands the plants ability to take up nutrients (up to 10 times); helps provides access to nutrients that are relatively inaccessible (phosphorus)

Ectomycorrhizae - primarily associated with temperate or semiarid-regions trees and plants; fungi penetrate plant roots but do not penetrate cortex cell walls (hence ecto-)

Endomycorrhizae - (arbuscular mycorrhizae are most important group); penetrate cortex cell walls and form "arbuscules," which transfer mineral nutrients to plant, and sugars from the plant to the fungus

*Soil prokaryotes* - Bacteria and archaea; cells lack nucleus surrounded by a membrane

Prokaryotes may be round, rod-shaped, or spiral; but in soils the rod-shaped prokaryotes appear to dominate.

Prokaryotes have enormous diversity (handful of soil may hold as many species as fauna in the amazon), and consist of both autotrophs and heterotrophs.

Perform important functions within the soil:

  - Break down of toxins (for example archae processing hydrocarbons, i.e., oil products)
  - Oxidation and reduction of particular elements (e.g., Iron and Mn)
    - Nitrogen fixation

*Controls on soil microorganism growth*:

  - Organic matter - almost any addition of OM will boost microbial activity and growth
    - additions of certain nutrients will prompt growth from different microorganism types
    - for example, bacteria respond well to starch and sugars whereas fungi and actinomycetes have heavier influence with additions of cellulose or other resistant material
  - Oxygen, moisture and temperature - increasing availability of oxygen, moisture and higher temperatures all promote microorganism activity
  - Exchangeable cations and pH - 
  
*Benefits to plants of soil organisms*:

  - Nutrient availability - primary benefit is the role in processing organic matter and increasing availability of key nutrients for plants
  - Breakdown of toxic compounds - can breakdown and process toxins that would otherwise accumulate within the soil profile
  - Inorganic transformations - can determine the availability of key micronutrients for plants
    - N, S, P, Fe, Mn...
    - May also determine the toxicity of particular elements (Se or Cr)
  - Nitrogen fixation - fixing of N~2~ gas by microbes which makes it available to plants
    - Cyanobacteria are important in wetlands and other anaerobic environments for fixing N; commonly occurs in root nodules or other associations with plants

*Soil organisms and damage to higher plants*:

  - Competition for nutrients
  - Herbivory - rodents, slugs, or snails. Nematodes and insect larvae are the most damaging/extensive; fungal diseases

Can control diseases or impacts on plants through management:

  - Maintaining particular pH levels - raising or lowering the pH may be able to rid the soil of particular diseases or fungal infections
  - Soil physical properties - allowing good structure and heating of soils can help to aerate soils or kill pathogens that may induce more significant problems
  
### Soil organic matter

Soil organic matter is critical to soil systems for a variety of reasons, but in particular for their increases in CEC, water holding capacity, and also importance to global carbon cycles.

Inorganic carbon may also form a substantial role of C pool in the ecosystem, and is formed largely through reaction of CO~2~ to produce carbonic acid as well as carbonates and bicarbonates of Ca, K, Mg, and Na. 

Nearly half of organic carbon in soils globally is found in histosols, gelisols, and inceptisols.

*Process of decomposition*:

Dry weight plant matter is approximately 90-95% comprised of C, H, & O.

Four broad classes of organic compounds in plant tissue:

  1. Carbohydrates - range in complexity from simple sugars and starches to cellulose
  2. Lignins and polyphenols - difficult to decompose
  4. Proteins - contain ~16% N and decompose easily
  
Decompositional process primarily consists of three steps:

  1. enzymatic oxidation of organic compounds - releases CO~2~, energy, water and biomass
  2. Nutrient specific release and/or immobilization
  3. Formation of resistant compounds either through microbial activity or modification of the compounds
  
*Mineralization* - process that releases elements from organic compounds to produce inorganic forms

Within a soil system, both r-strategist and k-strategist organisms exist in regards to OM decomposition.

  - r-strategists - "rate" organisms that decompose easily digestible OM at very fast rates
  - k-strategists - "constant" organisms that feed primarily on the more retractible OM found in the soil system
  
With addition of OM to a system, a sharp spike in populations of r-strategists will occur, and massive amounts of decomposition will occur. As OM levels are reduced, the r-strategists begin to die-off, which provides additional nutrients for surviving r-strateigists as well as k-strategists. Eventually r-strategists die-off to quiescence.

Some OM is physically protected during process, and some is left over as humus, which is a dark-colored, heterogeneous, mostly colloidal mixture of modified lignin and newly synthesized organic compounds. Oftentimes this accumulation is offset by k-strategist decomposition and thus there's little change in OM from year to year.

The factors that control rates of decomposition are primarily:

  1. Environmental conditions of the soil, for example:
    - sufficient moisture
    - sufficient aeration
    - warm temperatures (25 - 35 deg C)
  2. Quality of the added residues
    - location of the residue - soil surface vs. within the system; affects physical accessibility as well as temperature and moisture content of feedstock
    - size of particles - surface area to volume ratios are important
    - C:N ratio - can vary from 10:1 to 600:1 (sawdust)
    
Nitrogen is important in all organism for synthesis of amino acids, enzymes and DNA.

On average, soil microbes require a C:N ratio of 8:1 for synthesis of own organic compounds; only one third of C is assimilated into own tissues and thus they require C:N ratios of approximately 24:1.

Depending on the C:N ratio of the soil, availability of N for plant communities can be greatly influenced. A *nitrate depression period* is when OM of high C:N is added to the soil and microbial communities use up all available N. Until microbial communities consume all OM and N and begin to die-off, N will not be available for plants. The C:N ratio will slowly decrease as CO~2~ is lost via respiration whereas N is kept within biomass.

Consumption of microbial community by earthworms, nematodes or protozoa can result in release of NH~4~^+^ in the soil, as the macrofauna consume more N than they can consume.

**Genesis and nature of SOM and humus**:

SOM comprises of living as well as dead biomass (both recognizable and unrecognizable). The unrecognizable dead biomass is known as "humus". Humic substances are resistant to decomposition, whereas nonhumic substances are more easily degraded.

Humic substances are often ~60-80% of soil organic matter. Their complexity makes them resistant to decomposition and also induces their dark, black colors as they absorb light at many wavelengths.

Nonhumic substances contain definite chemical and physical properties and thus are more readily decomposed. 

**Influence of OM on plant growth and soils**:

Benefits of OM to plants occur largely indirectly, but some direct benefits exist:

  - Uptake of N or P as soluble organic compounds
  - Allelopathy - most allelochemicals can be included within the nonhumic OM of soil
  
Other direct effects relate to the effect of OM on soil properties:

  - Physical properties - humic fractions reduce plasticity & stickiness of clay soils; water-infiltration is improved; nonhumic substances can improve granulation and aggregate stability
  - Chemical properties:
    - Humus often accounts for 50 - 90% of cation-adsorbing power of mineral surface soils
    - Provides much of the pH buffering capacity through CEC and acid & base functional groups
    - Humic acids accelerate decomposition of soil minerals, which releases essential nutrients as exchangeable cations
  - Biological properties:
    - Provides food for much of the heterotrophic soil organism community

**Pools of organic matter**:

Depending on quality and availability, different "pools" of OM can be considered:

  - Active pool - easily decomposible OM with half-lives of days to a few years (~10-20% of OM)
    - e.g., living biomass, pieces of detritus, most of polysaccharides, other nonhumic substances
    - Can be readily enhanced through additions, but is also readily lost
  - Slow pool - finest fractions of particulate OM that is high in lignin, as well as other slowly decomposable or chemically resistant components
  - Passive pool - primarily material protected within clay-humus complexes; remains in soil for hundreds to thousands of years (~60-90% of OM)
    - closely associated with colloidal properties of soil and thus most responsible for CEC and water holding capabilities

**Other**:

Contributions to OM by plants in grasslands is often higher and more efficient than forests, given grasses greater biomass contribution roots and their greater resistance to decay.

This is interesting in that mangroves might have more significant root biomass than previously believed and thus the contributions from root biomass to SOC may be significant.

*Methane emissions* - methane emissions from salt wetlands are much lower than freshwater emissions, largely because levels of sulfate in sea water allow for sulfur-reducing prokaryotes to out-compete methanogens for organic matter within the ecosystem.

### Soil nutrients

**Key notes:**

Nitrogen is in general the most widely occurring nutritional limit on plant growth globally.

Phosphorus is generally the second most widespread nutritional limit.

**Nitrogen** - A major part of all proteins, nucleic acids, chlorophyll, etc.

Plants take up N as nitrate (NO~3~^-^) or ammonium (NH~4~^+^). Atmospheric N~2~ is abundant, but is not directly available given strong triple bond between N atoms. Microbial fixation or lightning is required to break the triple bond and make N available for plants.

Ammonium and nitrates account for the most important forms of inorganic N in the soil environment. Majority of N is found within soil OM, which is typically approximately 5% N.

N in ammonium is subject to five different fates:

  1. Immobilization - occurs via microoganisms
  2. Removal - uptake by plants
  3. Fixation - can be fixed within the layers of certain 2:1 clay minerals
  4. Volatilization - emission to atmosphere after being transformed into ammonia gas
  5. Nitrification - process in which is is oxidized to nitrite and subsequently to nitrate
  
N in nitrate is subject to five fates, some of which overlap:

  1. Immobilization -- occurs via microorganisms
  2. Removal - uptake by plants
  3. Leaching - loss to groundwater via percolation
  4. Volatilization - loss to atmosphere as several N-containing gases formed by denitrification
  5. Conversion to ammonium - dissimilatory (nitrate) reduction to NDRA by anaerobic organisms
  
*Immobilization and mineralization*

Much of N is protected in OM, but is eventually released as NH~4~^+^ through decomposition (mineralization).

Conversion of inorganic N to biological forms is termed immobilization, and is the opposite process of mineralization. May take place through either biological or nonbiological processes.

  - biological pathway occurs when microorganisms consume OM high in C relative to N, and thus scavenge nitrates and ammonium to account for deficiency from the soil; often removes mineral N from the soil (nitrogen depression period)
  - abiotic pathway is through rapid chemical reactions with high C/N ratio soil OM

Soluble organic N compounds are available for plant uptake and leaching losses; however the chemical constituents of SON are poorly known at present.

*Ammonium fixation by clays* 

Ammonium are attracted to negative surfaces of clays and humus, where they're held in exchangeable forms & are available for plant uptake. May become more permanently fixed in pockets within the crystalline structure of particular 2:1 clay types.

Although this can help preserve mineral N in the soil environment, it may also not be available to plants at rates at which it is needed.

*Ammonia volatilization*

Volatilization is conversion of NH~4~^+^ to NH~3~^+^ by reaction of ammonium with OH^-^. Given the dependence of the reaction on OH^-^, ammonia volatilization is much greater in soils of higher pH.

Ammonia volatilization in wetland soils can be quite significant, particularly during warm days when warm weather promotes photosynthetic algae consumption of dissolved CO~2~ and thus loss of carbonic acid results in increase in pH.

*Nitrification*

Conversion of ammonium to nitrates in a two-step reaction (NH~4~^+^ to NO~2~^-^ to NO~3~^-^). When ammonium is released into the soil, it is commonly rapidly converted to nitrate.

Nitrification may be reversed be several bacterial processes, some of which occur in anaerobic environments:

  - denitrification - heterotrophic bacteria reduce nitrate to NO, N~2~O or N~2~ gases
  - dissimilatory nitrate reduction to ammonia (DNRA) - reduces NO3- to NO~2~^-^ and then to NH~4~^+^
  
*Nitrate leaching* 

Nitrates readily move downwards with free water and thus are easily leached from the soil. This can result in toxicity problems for humans, livestock or other ecosystems if there is a large build up of nitrates.

*Gaseous loss by denitrification*

Denitrification of nitrates may produce NO, N~2~O and N~2~. NO (nitric oxide) and N~2~O (nitrous oxide) can be particularly environmentally damaging in at least four ways:

  - Form nitric acid, which is one of the principle components of acid rain
  - May react with volatile organic pollutants to form ground-level ozone
  - Very potent (but shortlived) greenhouse gases
  - N~2~O destructs ozone
  
May be extremely high in wetlands due to periods of wetting (denitrification) and drying (nitrates produced).

May be seen as an ecosystem service of natural wetlands in that they protect estuaries and lakes from eutrophication
  
*Biological N fixation*

Conversion of atmospheric N~2~ to forms usable by organisms by limited number of bacteria (rhizobium, actinomycetes, cyanobacteria).

Key component is the enzyme *nitrogenase*, which catalyzes reduction of N~2~ to NH~3~^+^. The ammonia is then used to form amino acids and subsequently proteins.

Nitrogenase is destroyed by O~2~, and thus the enzyme needs to be protected from exposure to O~2~. The molecule that protects nitrogenase (leghemoglobin) is virtually the same as hemoglobin that gives blood red color when oxygenated, hence root nodules on legumes are red.

Additionally, fixation of N~2~ is very energy intensive, and thus mutual relationships between N-fixing bacteria and plants occur. Host plant supplies bacteria with carbohydrates as energy sources, whereas bacteria provide the plant with reactive-nitrogen compounds.

**Sulfur**

Important as key constituent of essential amino acids, vitamins, enzymes & aromatic oils. Sulfur has three primary "natural" sources:

  1. Organic matter - microorganisms break down organic sulfur compounds to soluble forms analogous to release of ammonium and nitrate
  2. Soil minerals - inorganic forms are primarily sulfates and sulfides. Sulfates directly available to plants whereas sulfides must be oxidized to sulfate form before they may be taken up by plants
  3. Sulfur gases in the atmosphere - primarily COS, H~2~S, and S~O~2 are produced from volcanos, ocean spray, biomass fires and fossil fuel burning. These are eventually oxidized to sulfates which form sulfuric acid and result in acid rain.
  
*Sulfur cycle* -- There are four major forms of S that occur:

  1. Sulfides
  2. Sulfates
  3. Organic sulfur
  4. Elemental sulfur

Similar to N cycle, atmosphere is an important source and much of the S in the system is held within organic matter, which is subject to microbial oxidation and reduction.
  
For systems in which S uptaken by plants is eventually returned to the soil, the above three sources are typically sufficient. In agricultural systems in which biomass is removed, significant deficiencies in S may occur.

*Sulfur cycle*

  - Mineralization - Largely occurs through same pathway as N. Organic sulfur is decomposed and sulfates are released into the soil.
  - Immobilization - occurs when low-sulfur, C-rich organic forms are added; microbes will scavenge for mineral S to facilitate uptake of organic matter; temporary in that the inorganic sulfate will reappear in soil solution
  
*Sulfur oxidation and reduction*

Sulfuric acid (H~2~SO~4~) is produced through oxidation of H~2~S or S, which occurs at a wide range of soil conditions.

Sulfate ions tend to be unstable within anaerobic environments, and are reduced to sulfide ions by bacteria in two genera. THe sulfide ion will react almost immediately with reduced soluble Fe or Mn to form insoluble sulfides in anaerobic soils. 

Sulfide ions also undergo hydrolysis to form gaseous hydrogen sulfide, which gives swamps rotten-egg smell.


**Phosphorus** - Critical for ATP, DNA and RNA

Phosphorus is generally unavailable in soils due to three reasons:

  1. Total concentrations are typically low
  2. Compounds in soil often unavailable for plant uptake due to insolubility
  3. Soluble sources of P (fertilizers, etc.) are often fixed upon addition to soil
  
*Phosphorus cycle*

  - Plants absorb P typically as phosphate ions that are dissolved in soil solution. The P typically is slow moving within the soil system, and thus growth of roots or mycorrhizal connections are important in facilitating uptake of P
  - Decomposition of OM - microorganisms temporaily hold P as part of biomass, but will eventually release it through mineralization; portions of it will enter active and passive fractions of soil OM
  - Chemical forms - little P is lost due to volatilization or leaching, but rather limits of P are due to insolubility and fixation that makes it functionally unavailable. Primarily three general groups of compounds:
    1. organic phosphorus
    2. calcium-bound inorganic phosphorus - stable and insoluble at high pHs, more soluble at low pH
    3. Fe- or Al-bound inorganic phosphorus - opposite solubility of Ca-bound P
  - Gains and losses - losses are primarily due to biomass removal, erosion of P-bound soil particles, or P dissolved in surface runoff or leaching to groundwater; primary additions are through atmosphere, which is often sufficient in natural environments

*P availability*

Generally 20-80% of P is held in OM in soil systems.

For inorganic P, two primary controls on its availability exist: i) solubility of P-containing minerals, and ii) fixation or adsorption of P ions on surface of soil particles.

Phosphorus can precipitate with Fe, Al or Mn ions, which forms insoluble hydroxy phosphate precipitates. In very acidic soils, enough Al, Fe, or Mn typically present to precipitate all dissolved H~2~PO~4~^-^ ions.

Fixation of inorganic P with Fe- and Al- oxides and also 1:1 clays can be large given their large presence within many soils. Can fix P over very long term by additional precipitation of other Fe- and Al- oxides on top of the oxide bound P (termed "occluded" Phosphorus).

Iron reduction under wet conditions - anaerobic conditions can reduce Fe in complexes from Fe^3+^ to Fe^2+^, which makes the iron-phosphate complex more soluble and allows release of phosphorus into solution.

Big take-away is that P is generally present but is not available and thus there are high P limitations.

**Potassium**

Only present in soil solution as cation, and behavior in soil is thus largely effected soil cation exchange properties and mineral weathering.

Key role in organisms is as an activator of particular enzymes. Important in metabolic processes in mangroves and important in balancing Na^+^ levels.

*Potassium cycle*

Majority of K+ is in primary minerals and nonexchangeable forms. May be fixed within 2:1 clays, but overtime will become available through the exchangeable complex. 

Potassium is susceptible to leaching, but some plants with deep roots may mine K^+^ and act as "nutrient pumps" in that they move K^+^ to leaves and then return it to the soil surface upon loss of leaves. Plants take up large quantities of K^+^.

*Potassium availability*

Three factors shape availability of K^+^ within a soil system:

  - Small proportion readily available - Total presence of K^+^ is high, but the amount available through exchangeable complex at any given time may be quite small.
  - Leaching losses - quite susceptible to leaching (more than P, less than N); somewhat moderated by attraction of positive charge to negative surfaces of clays and humus
  - Luxury consumption - removal by plants is high, even without proportional increases in growth
  
Total amount of K^+^ available is largely a function of the kinds of clays present in a soil. Soils with 2:1 clays typically have more potassium whereas those with kaolinite tend to have the least.

**Calcium**

Calcium is a major component of cell walls, as well as cell elongation and division, membrane permeability, and activation of enzymes.

Deficiencies of calcium for plants is rare, except in acid soils.

Calcium is found primarily in three pools within the soil system:

  1. Ca contianing minerals
  2. Ca complexed with soil humus
  3. Ca held by CEC on clay and humus colloids
  
Much of Ca source within a soil system is through dry deposition (i.e., coal burning or wind erosion of calcareous desert soils)

**Magnesium**

Intimately involved with photosynthesis, and is taken up in relatively small quantities by plants.

Primary source of Mg is the exchangeable Mg on colloidal clay-humus complex, which is replaced by mineral weathering following depletion by leaching or plant uptake. 

**Micronutrients**

For macronutrients, the range at which nutrients can be uptaken and be used without being toxic is quite large. For micronutrients the range is quite narrow, however, and thus particular elements can become toxic for plant growth.

Micronutrients play a variety of roles in plants, though many are involved in enzyme systems. Different micronutrients are needed in different concentrations in plants.

Common micronutrients: Fe, Mn, Zn, B, Cu, Mo, Ni

*Availability of micronutrients* 

Inavailability of micronutrients may be due to lack of total concentration in soil, but is more commonly a result of unavailability of the particular nutrient.

Chemical forms of the various nutrients are important (e.g., oxides)

Factors that affect the availability of micronutrient cations include:

  - soil pH - increases in soil pH result in decrease in availability of micronutrient cations as they move to hydroxy ion forms and subsequently to hydroxide or oxide forms (insoluble)
    - general desirability of mildly acidic soil is to promote availability of micronutrients
  - oxidation state and pH - interaction of soil acidity and drainage important in determining micronutrient availability; Fe, Mn, and Cu generally more available in flooded soils
  - organic matter - decomposition of OM can result in slow release of micronutrients; additionally, binding of micronutrients to OM can protect micronutrients from binding to mineral complexes which makes them relatively more insoluble than when bound with OM
  mycorrhizae -- can be important in facilitating uptake of micronutrients

## Geomorphology {#geomorphology-soils}

### Thom, 1967

*"Mangrove ecology and deltaic geomorphology: Tabasco, Mexico"* [@thom1967mangrove]

A full summary of Thom, 1967 is provided in the "Zonation and Disturbance section." Please refer to Section \@ref(thom1967).

From a geomorphological standpoint, this is one of the most well-cited studies arguing that mangrove vegetation patterns respond primarily to the hydrological and sedimentary conditions that exist within a given site.

### Carlton, 1974

*"Land building and stabilization of mangroves"* [@carlton1974land]

**Key contribution:** This is a perhaps less seminal paper but provides a brief review of the literature surrounding the ability of mangroves to "build" and "stabilize" shorelines. Much of the reviewed work is from Florida.

**Key conclusions:** Carlton concludes more or less that mangroves do not "build" land *per se*, but are influential upon the *rate* at which sediments accrete on shorelines. In sum, two viewpoints exist:

  1. mangroves "build" land by accretion of sediments (*a la* Davis 1940) (UNSUPPORTED)
  2. mangroves modify the rate at which sediments accrete, thus affecting the speed of geomorphic change rather than the direction (Egler 1952, Gill 1971) (SUPPORTED)
  
*Further reading:*

  1. Egler 1952
  2. Gill 1971

### Thom, 1982 {#thom1982}

*"Mangrove ecology: A geomorphological perspective"* [@thom1982mangrove]

**Key contribution:** This study is best known for the five environmental settings that Thom lays out for clastic (terrigenous) sediment deposition. The five environmental settings are detailed below:

  1. *River dominated* - Deltaic or estuarine settings in which significant sediment is transported from alluvial flows to the coastal region. The landscape is marked by numerous distributaries and interdistributary basins that vary in hydrology, particularly in regards to tidal influence.
  2. *Tide dominated* - Estuarine type settings in which the dominant hydrology is the bidirectional flow of tides. Often see sandy shoals within the estuary that are shaped parallel to the flow of tides.
  3. *Wave dominated* - Environmental settings in which flat shorelines accumulate sand barriers or cheniers that create lagoons and create regions of low-energy coastlines in which mangroves may establish. Sediment is primarily autochthonous.
  4. *River & wave composite* - Settings in which both deltaic plains with strong alluvial influence as well as strong wave influences create a complex of alluvial fans and lagoons blocked by wave-distributed sand barriers from longshore flows. E.g., Tabasco, Mexico.
  5. *Drowned river valleys* - Estuaries or embayments that are underlain by impervious rock and have been flooded during the most recent period of sea level rise.
  
**Key knowledge:** Thom describes the geomorphology of mangrove from three standpoints:

  1. *Geophysical* - The dominant physical processes that affect a landscape, such as SLR or climate
  2. *Geomorphic* - The dynamic history of the land in response to the dominant geophysical factors. Can be broken into three scales:
    a. macro: clastic vs. carbonate sediments
    b. meso: dominant physical process (tide vs. riverine vs. wave)
    c. micro: microtopographic features such as drainage and elevation that produce environmental gradients
    
  3. *Biological* - Factors such as biogeography or competition that affect the landscape
  
Additionally, Thom notes that his five suggested geomorphig settings *are not comprehensive but are rather representative of dominant processes that may be relevant.* Thom goes on to discuss variations and scale in time and space and their resultant effects on mangrove geomorphology. He concludes that zonation of mangrove species responds opportunistically to geomorphic setting, and areas with high geomorphic diversity will have high degrees of habitat variation.

**Topics to think about further:**

  1. How do geomorphic settings related to soil texture and soil carbon?

### Woodroffe, 1983

*"Development of mangrove forests from a geological perspective"* [@woodroffe1983development]

**Key contribution:** This paper describes the development of an autochthnous, geomorphically inactive mangrove forests on Grand Cayman island. The study takes a geological perspective in that they examine soil depths and soil cores throughout the west peninsula, and reconstruct models of autochthonous development of the mangrove forest through time.

**Key findings:** The study concludes that there were key tidal inlets which were eventually breached my rising sea levels and allowed for expanded colonization of the calcerous bedrock by mangroves. Soil depths & cores indicate that there was a forest of landlocked mangroves at one point during low-levels (200 cm below present), but the authors are unsure of how tidal fluxes reached this area.

```{r, echo=FALSE}

knitr::include_graphics(rep("images/woodroffe1983TVS.bmp"))

```

### Thom, 1984 {#thom1984}

*"Coastal landforms and geomorphic processes"* [@thom1984coastal]

**Key contribution:** This is an extension of [@thom1982mangrove], in which Thom describes three carbonate settings. The conclusions are largely the same as [@thom1982mangrove], and thus refer to section \@ref(thom1982) for them.

**Key additions:** The carbonate settings that Thom adds to his original five clastic settings are as follows:

  6. *Carbonate platform* - Accumulation of lime muds and peat on low-energy tropical coasts. Oftentimes the fringing mangroves are against large, open, shallow bodies of water. Coral reef or sand barriers may block wave and tidal energy to produce low-energy conditions
  7. *Sand/shingle barrier* - Mangroves may form on the cemented or non-cemented coral detritus, for example patchy mangroves colozining portions of the Great Barrier Reef. Sand mostly consists of carbonate particles and sea levels play an important part on depositional characteristics
  8. *Carbonate embayments* - Embayments in which soil "crusts" were submerged by SLR during Holocene. E.g., patchy mangroves on old reef complexes that are highly variable and reflect varability in the substrate complexity.
  
**Key knowledge:** In addition, Thom discusses some of the prior models of succession and zonation in mangroves. In particular he identifies three trends of research, including one in which he situates himself:

  1. *"Classical succesional view of mangroves"* - The "Davis 1940" view of mangroves that emphasizes the role of biota rather than geophysical or geomorphological processes
  2. *"Environmental gradient view"* - The more "ecological niche" based view of vegetation patterning that results from environmental gradients; Watson 1928 was one of first studies to argue for this effect
  3. *"Physiographic ecology"* - regards physical and botanical landscapes as mosaics of landforms that result from differential effects of physical processes on different materials, and the promotion of different biological communities in response to them
    - focused at landscape scale, geomorphic variety produces variety in biological assemblages

**Other reading:**

  - Snedaker, 1982 - Mangrove species zonation: Why?

### Woodroffe, 1990

*"The impact of sea-level rise on mangrove shorelines"* [@woodroffe1990impact]

**Key contribution:** This review by Woodroffe discusses the impacts of relative sea level rise as it was projected at the time on mangroves. There has since been 27 years worth of research in this topic, and thus it's important to couple this study with more recent papers.

*Crux of the issue* - "order of rates of sea-level rise that have been predicted would in some cases lead to major disruption of mangrove forests"

**Key notes: **

  - ability of mangrove forests to respond to different levels of sea level rise depends on coastal morphology
    1. steep cliffs (e.g., mid-oceanic atolls) may not be able to respond due to lack of ability to migrate along shoreline as well as distant propagule sources for recolonization
  - Eustatic patterns during Holocene have varied regionally
    - in Indian and Pacific Oceans, rose to approximately 1-2 m above present levels, before falling and stabilizing at current levels
    - in West Indies and Southern US, increasingly risen though at decelerating rates to arrive at period of stability of today
  - mangroves are effective indicators of past sea-level if species with given preference for position in intertidal zone can be identified within paleosedimenets
    - however, often not clear because of compaction and contamination with younger aged organic matter
  - given sea level rise, the ability of mangrove forests to persist may be largely dependent upon the catchment size of deltas (i.e., size of area from which water and sediments drain)
  - carbonate settings are at higher risk of undergoing regime shifts relative to deltaic or estuarine mangroves
    
Woodroffe combines Thom's environmental settings into three primary classes of geomorphological development:
  
  1. river dominated
  2. tide dominated
  3. carbonate
  
He expounds on them in this article, whereas he summarizes them in his review in Tropical Mangrove Ecosystems (see Section @\ref(woodroffe1992)).

### Woodroffe, 1992 {#woodroffe1992}

*"Mangrove sediments and geomorphology"* [@woodroffe1992mangrove]

**Key contribution:** This is a key synthesis of knowledge related to geomorphology and environmental settings in mangroves. Woodroffe gives an overview of sedimentation and shoreline progradation / erosion, followed by a review of Thom's environmental settings and Lugo and Snedaker's mangrove habitats. He concludes the article with a review of stratigraphy and paleoecology, with particular reference to mangrove responses to relative sea level rise.

**Key findings:**

  - For Thom's clastic environmental settings, refer to Section \@ref(thom1982) 
  - For Thom's carbonate environmental settings, refer to Section \@ref(thom1984)
  - For Lugo and Snedaker's functional habitats, refer to Section \@ref(lugosnedaker1974)

Diagrams of Thom's environmental settings are provided below:

```{r, echo=FALSE}

knitr::include_graphics(rep("images/thom1982.bmp"))
knitr::include_graphics(rep("images/thom1984.bmp"))

```

*Stratigraphy and paleoecology:*

  - Quaternary period (~past two million years) comprised of:
    1. Holocene (~11,700 YBP)
      - sea level has been approximately stable for 6,000 years
      - sea level was rising rapidly from approximately 10,000 - 6,000 YBP
      - though there is likely regional variation
    2. Pleistocene (~2.5 mil to 11,700 YBP)
  - Sediments deposited during Holocene are well preserved and contain structures & fossils that are useful, whereas sediments deposited during Pleistocene are less well preserved
  - sea level rise on more gradually sloping coasts likely resulted in significant fluctuation of mangrove forests, whereas carbonate atolls (e.g., on coral) likely disappeared during high sea levels or were restablished during low sea levels
  - *transgression* - landward movement of shoreline during periods of sea level rise
  - *regression* - seaward movement of shoreline during periods of sea level subsidence
    - there are also iteractions with progradation and erosion that may reverse processes at specific locations
  - Woodroffe examines three patterns of tropical geomorphological development:
    1. tropical deltas that are river dominated (poorly studied as of 1992)
    2. estuarine embayments that are tide-dominated
      1. Progradational model - estuary shoreline gradually build into estuary and is generally fringed by mangroves
      2. Big swamp model - rapid accumulation of sediment as sea level stabilizes followed by vertical accretion and conversion to other types; differs from progradational model in that there's little age difference between mangrove sediments
      3. Barrier estuary/mud basin - sand reworked from seward shelf and forms sandy barrier or flood tidal delta; lagoon is sink for fluvial sand and mud from terrestrial sources by rivers
    3. carbonate banks or reefs
      - predominantly characterized by in situ production of sediments

*Note: * This is a key study that would be good to review again later.

### McKee, 2007

*"Caribbean mangroves adjust to rising sea level through biotic controls on change in soil elevation"* [@mckee2007caribbean]

**Key contribution: ** This is a very well-cited study that experimentally examples biological controls on soil elevation in mangroves of the Meso American Barrier Reef (in Belize). 

**Study design: ** The authors installed elevation rods with rod surface elevation tables (RSETs) that were able to measure changes in elevation from a baseline over the course of three years in fringe, transition and interior mangroves. They examined the effects of nutrient additions (N, P & a control) across three transects at three replicate sites. In addition, they laid a sand horizon in a plot adjacent to the RSETs and used cores to monitor subsurface accretion, allowing for differentiation between subsurface accretion and surface accretion of biomass to elevation. Finally, the authors examine several cores (up to 10m in depth) across sites throughout the Caribbean to examine relative SLR and mean soil accretion.

**Key results: ** The authors conclude that the fringe portion of the forests are able to keep pace with sea level rise, primarily due to subsurface accumulation of biomass (significantly correlated with elevation change). The fringe zone was found to have significantly positive gain in elevation, with majority from subsurface change, whereas interior and transition zones saw subsidence. Elevation gains were seen across all sites for P addition, whereas elevation gain was variable in directino for N (negative in interior).

The authors give three key points:

  1. Rates of modern change in elevation may vary in both direction and magnitude within forests
  2. Elevation changes vary in relation to physico-chemical gradients that may control natural productivity of mangroves
  3. Elevation changes due to variation in nutrients are variable in both direction and magnitude
  
The key results in elevation change across treatments are shown below:
  
```{r, echo=FALSE}

knitr::include_graphics(rep("images/mckee2007GlobEcolBiog.bmp"))

```

The results from the mangrove peat cores showed that peat accumulation corresponded with rates of SLR over the past 6,000-7,000 years. Although they are currently keeping pace with sea level rise, the cores show that at points (in which SLR exceeded 5 mm / yr), mangroves were lost from the area. Thus, the presence of mangroves at these sites depends on relative SLR rates (eustatic rise plus local subsidence). Rates of mangrove peat accretion, however, may underestimate ability to respond to SLR as these are often based on mean rates from historical records that likely fluctuate significantly in response to SLR.

### Woodroffe, 2016

*"Mangrove sedimentation and response to relative sea level rise"* [@woodroffe2016mangrove]

**Key contribution: ** This study is a review of sedimentary processes and geomorphology with a particular focus on their responses to relative sea level rise. The study introduces the concept of the accomodation space in the context of mangroves, which is the space available for potential sediment accumulation.

**Notes: **

Considering processes at different scales:

  1. Macroscale - processes at the landscape level (environmental setting *a la* Thom)
    - Climate, eustatics
  2. Mesoscale - habitat level (fringe, basin, etc. *a la* Lugo and Snedaker)
    - Hydrodynamics, sediment supply (allochthonous vs. autochthonous)
  3. Microscale - individual tree or plot level
    - Interactions between surface elevation, geomorphic processes (accretion vs. compaction), and biological processes (BGB production)

```{r, echo=FALSE}

knitr::include_graphics(rep("images/woodroffe2016ARMS.bmp"))

```

A nice summary of interactions between sedimentology, hydrology and biology:

"The distribution of mangrove species coincides with the intertidal accomodation space, which progressively fills through sediment accumulation. Emergent properties and biocomplexity are a function of hydroperiod and are related to salinity and flooding tolerance, the latter of which is related to root morphologies, physiology, and reproductive traits such as vivipary."

*Accomodation space* - the space available for potential sediment accumulation. Typically see highest accomodation space at shoreline in mangroves, though it may be highly variable (e.g., basin areas that are only flooded during highest tides).

  - may distinguish between mineral vs. biological accretion, and thus may consider "ecological accomodation space"
  
Rates of sedimentation depend upon the accomodation space, which depends on hydrology and other factors. In the short term, the rate of sediment deposition appears independent of rate of SLR. The accomodation space is directly constrained by upper limit of tidal flooding.

Focusing on vertical adjustment within a mangrove forests risks missing broader regional picture in that mangroves have constantly shifted position throughout time, and thus there may be natural fluctuations with sea level rise that can maintain existence as well as are potentially key climate change mitigation opportunities.

**Further reading:**

  - See the highlighted references in Mendley for additional reading -- there are quite a number worth looking at further.


## Sedimentation and hydrology

### Wolanski, 1986 {#wolanski1986}

*"Tidal mixing and trapping in mangrove swamps"* [@wolanski1986tidal]

**Key contribution:** This study primarily seeks to answer two key questions:

  1. What parameters control export of salt left after evapotranspiration of tidal waters by mangrove trees?
  2. How do rivers influence the growth of mangroves?
  
There is a lot of mathematical hydrological modeling that is over my head in this study, but there are some relevant ecological points that are worth noting.

**Key results:** Some key results from the study include:

  - The "trapping mechanism" is the dominant parameter controlling "longitudinal dispersion" in their study swamp. They conclude that it is likely the dominant mechanism in many swamps in the absence of large buoyancy effects (i.e., dry seasons)
  - Buoyancy effects are important due to trapping of freshwater in the swamp at high tide and boundary layer that enters the river banks at low tide
  - Residence time of runoff in mangrove swamps may be quite long with important ecological implications, such as nutrient uptake and settling of fine sediments
  
A diagram of the "trapping mechanism" of fresh water in the landward portion of swamps is shown below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/wolanski1986.bmp"))
```
  
**Further reading:** This study led to some questions that may be important for understanding soil formation in mangroves. In particular:

  1. What is the "trapping mechanism" in estuarine hydrodynamics?
  2. What are the specific processes of tidal mixing and what is the broader relevance for the ecology of mangroves?
  3. What is longitudinal dispersion in relation to mangroves?
  4. What are the specific implications of buoyancy effects, and how do they vary between dry and wet seasons?

### Wolanski *et al*, 1992 {#wolanski1992}

*"Mangrove hydrodynamics"* (In Robertson and Alongi, 1992) [@wolanski1992mangrove]

**Key contribution:** This is a comprehensive review of hydrodynamics in mangroves, that comes from the Robertson and Alongi book.

**Key notes: ** 

Of the 5 (6) different Lugo and Snedaker geomorphic setting classifications, hydrodynamics in the riverine forest type has been studied the most extensively.

*Tidal circulation* - Tidal circulation is often the dominant cause of water movement.

Ebb tides are shorter than flood tides, but have slightly stronger water flows.

Water movements within tidal creeks are frequently greater than 1 m/s, however water flows within the forest itself are rarely greater than 0.1 m/s.

Mathematical models of hydrodynamics in mangroves have largely focused on two components:

  1. Tidal creeks - where acceleration, sea level gradients, friction and inertia are all important
  2. Forest flood - where inertial effects can largely be neglected
  
*Tidal assymetry* - Currents in tidal creeks are strong during ebb flow relative to flood flow.

The key to understanding this is that friction within the bed of the creek as well as the forest floor produces a time lag between the time of high tide at the head of the creek, versus the time of high tide at the mouth of the creek. The higher water level at the head of the creek accelerates flows back towards the mouth of the estuary.

The stronger ebb flow relative to the flood flow is important in maintaining channels and also export of suspended matter from the mangrove; reductions in forest extent will impact ebb flow as the difference in water height between shoreward and landward zones is a function of the distance and is greatly reduced.

The creek will then silt rapidly until balance is reachieved in which scouring action can reoccur.

*Exchange with the nearshore zone* - The presence of salinity within any given location in the forest and the intermixing of coastal, tidal and freshwater flows depends on the tide, season and relative amount of each within the system.

Bounds between freshwater and saline water may exist in the uppertidal reaches following a flushing of the estuary by freshwater (e.g., heavy rain event and low tide), which is produced by influx of high tide and limited mixing with the freshwater. May take several weeks for freshwater to be flushed from the system.

*Transverse circulation* - Secondary circulation phenomena exist due to primary tidal circulation and net downstream transport of freshwater.

Longitudinal circulation - caused by saltwater inflow at creek bottom and freshwater flow at surface from density effects.

Transverse flows produce linear aggregations of leaf litter and propagules and, though understudied, are likely formed through similar processes as linear oceanographic features near coral reefs.

*Tidal currents and sediment fluxes* - At the time of this article, there were key knowledge gaps in reference to tidal currents and sediment fluxes. See the Fukukara pieces for more precise information on sedimentation rates within the forest (Sections \@ref(furukawa1996) and \@ref(furukawa1997)).

In general, flow rates are not sufficient to reentrain sediment in the water column, particularly during flood tides. This provides support of the importance of ebb flows in scouring process for maintaining channel depths.

*Groundwater flow* - Flows of groundwater are important but are poorly understood. During initial flooding of forest floor, water may first be seen through bubbling out of crab holes which is an important mechanism for maintaining lower salt concentrations.

*Climatic and tidal influences* - Amount and periodicity of precipitation greatly influence the hydrodynamics of a given site. Balance between rainfall and evaporation, when coupled with tidal variations, is a key factor that determines salinity status of various portions of the mangrove.

A diagram of physical, biological and chemical processes as they relate to mangrove hydrology is shown below:

```{r, echo = FALSE}
knitr::include_graphics(rep("images/wolanski1992.bmp"))
```


### Furukawa, 1996 {#furukawa1996}

*"Sedimentation in mangrove forests"* [@furukawa1996sedimentation]

**Key significance:** Furukawa and Wolanski show through an empirical study that a sediment pump from coastal waters to the mangrove interior does exist.

They find that the presence of a tidal pump is largely due to:

  1. High turbulence that maintains entrainment of fine sediments during flood tide.
  2. Settlement of flocculated sediment during slack high tide in which turbulence is reduced, and increased flocculation which results in order of magnitude increase in size.
  3. Insufficient ebb tide velocities to resuspend settled sediment (of now larger size).

**Key notes: **

Water velocity of tidal flooding in a mangrove forest can be described as:

$$ u = 1/n\ *\ h^{2/3}\ *\ I^{1/2}  $$

Where:

  - *u*: water velocity
  - *n*: Mannings friction coefficient
  - *h*: water depth
  - *I*: water surface slope

Typical values for the Mannings friction coefficient for sandy channels are in the range of 0.025-0.035, whereas empirical studies by Wolanski have found values as high as 0.2 - 0.4 for areas with mangrove vegetation. Thus, mangrove vegetation increases the Mannings friction coefficient by several orders of magnitude.

This results in a failure to resuspend fine sediments that are deposited during slack high tide and thus promotes build-up of sediment.

Consideration of floc size is important for sedimentation studies within mangroves. In calculating the settling velocity (via Stoke's law) for both individual particles of clay and silt, as well as that of mean floc size measured within the tidal water, settlement of flocs was found to be approximately 100 times larger than for individual particles (0.005 m/s).

Empirical measurements in the field showed that for tree species with complex matrices of aboveground roots, turbulence was intense enough to maintain flocs in suspension. For single stemmed trees without root systems (e.g., Bruguiera or Ceriops), both turbulence and flow velocity are reduced.


### Furukawa, 1997 {#furukawa1997}

*"Currents and sediment transport in mangrove forests"* [@furukawa1997currents]

**Key contribution:** A sister study to [@furukawa1996sedimentation], this study describes the empirical data collected to investigate current flows and suspended sediment characteristics of a mangrove in Australia.

**Research design:** Water level, currents, sedimentation (via sediment traps and Analite nephelometers (upward looking cameras)), suspended sediment levels (via water samples), and concentration of flocs (via microscopic slides with wells) were measured along a transect in the forest, at varying distances from a tidal creek.

The data were compared against three hydrological models to investigate the accuracy of mathematical modelling of hydrodynamics and sedimentation in mangroves, as well as analyze the empirical data for the site. The models were:

  1. 2DSWAMP - flow model based on primitive equations of motion
  2. LAKE - a lake model that assumes flat water surface and determines flow rate
  3. VORTEX - a discrete vortex model that provides information on turbulence

**Key findings:**

The LAKE model performed poorly and was not used within in the subsequent analyses (unrealistic assumption that the surface water is flat).

Suspended sediment concentration showed strong asymmetry in that concentration was higher at flood tibe and much lower at ebb tide. This is indicative of net inwelling of sediment into the mangrove.

Sedimentation rates decreased with distance from the tidal creek. In calculating the sedimentation velocity, the median floc size (90 micrometers) was approximately 4 times that of size of solid minerals (24 micrometers) given high porosity of flocs. Visual investigation of suspended sediment reveals that much of the flocs had settled out of the water column following slack high tide.

Vegetation greatly increased turbulence and created a complex set of hydrological effects (eddies, stagnant regions, and jets).

DOC was found to vary inversely with tidal depth, with the lowest DOC concentration present during slack high tide. Furthermore DOC in the forest was found to be greater than coastal waters, indicating that both creek and mangrove waters are significant sources of DOC. 

The findings suggest a net inwelling of DOC in the forest (greater DOC concentration during flood tide than ebb tide), but this does not corroborate with other analyses.

### Krauss *et al*, 2003 {#krauss2003}

*"Differential rates of vertical accretion and elevation change among aerial root types in Micronesian mangrove forests"* [@krauss2003differential]

**Key contribution:** Understanding of sedimentation and hydrology in mangroves has largely been theoretical and model based, and thus this study represents one of the only studies to investigate the differential effects of different root morphologies on sedimentation in mangroves empirically.

**Study design:** Rates of sediment accretion and elevation change for three different root types were examined in three different river basins in the Federated States of Micronesia. Experimental plots of 25 m in radius were established within the fringe, riverine and interior sections of each of the river basins, and three 1 m^2^ plots were established in the plots in each of the three root types (Rhizophora prop; Avicennia pneumatophore; and Bruguiera knee) as well as a control.

Marker horizons (50% feldspar clay, 50% sand) were laid and 5 100 cm pins were inserted 70 cms into the soil to estimate sediment accretion and surface elevation changes, respectively. Additionally, root basal areas were calculated at a height of 4 cm for comparison of root biomass to sedimentation characteristics. Fine root fraction estimates were taken via soil cores and washing of mineral soil and particulate OC to achieve just the fine root fraction, which was compared with surface elevation data.

**Key results:** The results provided several key findings:

  - The results showed that Rhizophora roots had significantly higher rates of sediment accretion relative to the others.
    - The authors attribute it to increased turbulence and thus flocculation of suspended sediment, reduction in flow which traps it within the root complex, and thus higher rates of sediment accretion during slack flows
  - No significant differences among functional root types for elevation change were found across all basins, but root type x river basin interactions were found to be significant (control plots or pneumatophores having higher rates of positive sediment surface elevation change).
    - Elevation chane rates were found to occur through series of "pulses" rather than steady through time
    - Lower rate of surface elevation change in Rhizophora was suggested to potentially having to do with the lower ratio of lignin:Nitrogen in the fine root fraction, which may have been more decomposible and thus result in elevation decreases.
  - Root area and root density did not have significant influence on monthly rates of vertical accretion.

Furthermore, found that surface elevation change in Avicennia pneumatophores was highest, and the authors conclude that organic matter contribution is thus significant -- might have implications for research questions.

### Adame *et al*, 2010 {#adame2010}

*"Sedimentation within and among mangrove forests along a gradient of geomorphological settings"* [@adame2010sedimentation]

**Key contribution:** This is one of the few studies that empirically tests for differences in sedimentation rates amongst different geomorphological settings and different forest types (fringe vs. scrub vs. algal mats) within the different geomorphological settings.

**Study design:** The study primarily employs petri dishes held to the forest floor as well as water samples to example differences in total suspended sediment, sediment accretion and the presence of glomalin (a secretion produced by terrestrial arbuscular mycorrhizal fungi) across forest types/settings.

Glomalin was used as an indicator of terrestrially versus other derived sediments, as AMF largely do not exist in mangroves and thus this would be indicative of terrestrially derived sediments.

Replicates plots in the fringe, scrub, and upper-tidal (algal mats) portions of six different sites, which were a variation of river-dominated vs. tide-dominated geomorphological settings, were established. Samples from the water column and sediment deposition on petri dishes were taken to investigate variation in deposition patterns.

Interestingly, the study also used PCA of several environmental variables to establish a quantitative measure of the geomorphic setting. Among the environmental variables included were:

  1. Salinity
  2. Catchment area (km^2^) - unsure specifically what this is
  3. DIstance to sea (km) - unsure as to how this measure was taken
  4. Entrance section area (m^2^) - unsure as to what this is
  
**Key results:** Their results found that sedimentation patterns varied between geomorphic types, as well as environmental settings within the river-dominated vs. tide-dominated settings.

*Export of total suspended sediment*:

No relationship between sediment retention and geomorphological setting was found. For those sites that did show retention of total suspended sediment (TSS), retention was found during the ebb tide in the fringe and scrub zones.

*Sedimentation*:

The fringe zone was found to have the highest sedimentation rates, which the authors attribute to the greater density of vegetation (and thus AG root complexes? though they do not specify tree species).

A relationship between sedimentation rate and geomorphological setting was not found.

In general, sedimentation was evenly distributed in riverine dominated geomorphological settings, whereas tidal mangroves found majority of sedimentation to occur in the fringe zone.

*Glomalin*:

The amount of glomalin was correlated to its age (determined via *immunoreactivity*), and riverine sites were found to have much higher and younger amounts of glomalin than tidal sites in the TSS.

In the deposited sediments, glomalin did not correlate with geomorphological setting, but did exhibit differences between intertidal zones within riverine and tidal settings. The majority of glomalin deposited in riverine settings occurred in the scrub zone, whereas the majority in tidal settings occured in the fringe zone.

### Lovelock *et al*, 2010 {#lovelock2010}

*"Mangrove forest and soil development on a rapidly accreting shore in New Zealand"* [@lovelock2010mangrove]

**Key contribution: ** This study examines the development of forest structure and soil conditions in two naturally established mangrove forests in New Zealand. The study is interesting in that it provides an interesting description of naturally establishing and developing forests in potentially newly available sites (i.e., at the ends of their range).

**Key notes: ** 

The article opens with an excellent review of succession and forest development in mangroves, for which the literature is relatively limited. Of particular note is that most field studies have focused purely on forest structure (i.e., stem density and basal area) rather than biogeochemical properties through time.

*Methods*

The study uses a chronosequence of two mangrove forests to examine development over time and test the following hypotheses:

  1. Mangroves are youngest on seaward fringe and oldest on landward edge
  2. Mangroves follow successional patterns described in Odum 1969
  3. C stocks in soils increase with age
  4. Soil respiration varies with forest age, declining with forest productivity
  5. N limitations on growth increases with forest age, similar to terrestrial forest systems
  
The site they use is at the mouth of the Thames river, and is in a macrotidal site that has large supply of fine-sediment that is redoposited along the coast in intertidal-mud flats.

Transects were laid across the intertidal (~1 km) and forest structure, soil samples (nutrients, C, redox and soil respiration), and tree cores were taken for bomb pulse radiocarbon dating.

Additionally, nutrient addition experiments were performed in the youngest section of the fringe forest and the tallest and oldest forest. Trees were fertilized directly with N and P, and growth was assessed over 14 months.

They compared the young (fringe) forest to the older (landward) forest to examine several successional criteria:

  1. Community energetics - gross production assumed to scale linearly with LAI, and thus production/biomass can be estimated as LAI/basal area
  2. Community structure - Several different metrics were used here:
    - total organic matter - soil carbon and aboveground biomass
    - ratio of porewater nutrients to total foliar nutrients
    - biochemical diversity estimated as CV in soil redox potentials
    - spatial heterogeneity as CV in tree size
  3. Selective pressure - quality and quantity of production assessed by leaf production, leaf nutrient concentrations, and specific leaf area
  4. Overall homeostasis - nutrient conservation as foliar nutrient resorption efficiency

*Results*

LAI was variable along the transect whereas other forest structure parameters (height, density and BA) were variable, but not in a linear manner or one that corresponded to distance from shore.

Within the fringe, they found trees with markedly different heights but were likely recruited at approximately the same time, indicating that they had differing rates of growth.

The radiocarbon dating indicated two different periods of broad establishment, with variation in forest structure in each of the recruitment classes.

Soil C and N increased strongly from the seaward to the landward edge. Soil C ranged from 2% at the fringe to 6% at the landward section, but then decreased sharply at the landward edge.

Leaf nutrient concentrations were highest at the fringe, and N resorption was lower whereas P resorption was higher.

Addition of N fertilizer significantly increased stem extension rates in young and old forest, whereas fertilization with P did not have significant effects.

Assessing succession showed that differences between the young and old forests tended to adhere to successional model predictions:

  - Community energetics - found to be higher in young forests for LAI per BA, but soil respiration per BA was not different
  - Community structure - soil C and biomass were lower in the younger forest; tree size was more variable; CV of soil redox potential was higher in younger
  - Selective pressure - leaf nutrient concentrations not significantly different, but younger forest had higher growth rates
  - Overall homeostasis - nutrient conservation did not differ between forests.
  
*Discussion*

They expected to find higher density and lower BA forests at the fringe, but found variable structures across both forest ages. Additionally, found two major pulses of establishment to exist.

The authors suggest a strong time component in which forest growth is pulsed and may be dependent upon favorable climatic conditions. The two periods they identify coincide with sustained El Nino activity. Given the orientation of the site, El Nino conditions would reduce wave energy and thus may favor seedling recruitment.

Patterns of forest structure seen at other sites and stylized by Jimenez and Lugo (1985) did not correspond to those seen at the site. Tree density did not differ between seaward and landward edges, although height and basal area increased from younger to older forest.

Characteristics of mature forests are believed to occur around 15 years, and persist until senescence at 55 years. The forests studied here corroborated with that, with the 30 year old forests showing mature forest traits.

Their results gave evidence that mangroves display characteristics consistent with models of succession:

  - Dissolved organic N and N:P ratios in porewater decreased towards landward edge
  - Total accumulation of soil C, N and P increased towards landward edge.

## Mangrove SOC

### Amundson, 2001 {#amundson2001}

*"The carbon budget in soils"* [@amundson2001carbon]

**Key contribution:**

**Key notes: ** This article provides a review of the interface between soils and the global carbon budget. Discussion of its significance, different methods for estimating global SOC stocks, and the impacts of human activities on it.

*Soil organic matter:*

Soil organic matter is defined as organic remains of plants and animals that is either recognizable or has decomposed to the point in which it is no longer recognizable (humus). SOM may be living or non-living, and microbial communities form a key component that is primarily responsible for the cycling of OM.

*Global soil C reservoir*

Two primary approaches to estimating it:

  1. Soil taxonomy - mean SOC values by soil taxon are extrapolated to their areal coverage globally via soil maps
  2. Ecosystem-based approaches - based on observation that SOC and plant communities are both largely a function of temperature and precipitation, and thus use the Holdridge ecological life biomes. Mean SOC values are estimated by biome, and extrapolated to the total areal coverage
  
Estimates as of 2001 estimated total SOC within the top meter of soil to be approximately 1,500 Gt C

*Controls of SOC*

State factors of soil formation + human activity (Climate, Organisms, Parent material, Topography, Time, Humans).

Different pools of SOC exist, which have varying degrees of susceptibility to being cycled:

  1. Fast (10^0^ years) - e.g., fresh plants and roots
  2. Slow (10^1-2^ years)
  3. Passive (10^3-4^ years) - physically or chemically protected OM
  
Change in SOC with time can be expressed as:

  $$ dC/dt = I -kC$$

Where I is the SOC input, k is the fraction of C lost as CO~2~ each year, and C is the mass of C per soil volume.

*Variation of SOC with time*

Two general approaches exist for determining rates of SOC accumulation and cycling over time:

  1. Chronosequences - which monitor SOC in soils of different ages but with similar environments and bedrock
  2. Mass balance approach - C cycling rates are inferred for soils near or at steady state
  
In using chronosequences, need to constrain time period in which one looks because of the importance of climate on SOC cycling. Most chronosequences of SOC are constrained to the Holocene period.

Understanding of SOC dynamics from using chronosequences includes:

  1. C accumulates in soils for thousands of years (perhaps as long as 10,000 years)
  2. Accumulation rates decrease with time, eventually approaching 0 as steady state is reached
  3. Steady-state C is related to climate and other environmental factors
  4. Rates of C accumulation in young soils are likely controlled by N availability in ecosystem
  
Understandin of SOC dynamics from mass balance approach:

At steady state, inputs can be assumed to equate to the decay of C present, thus: $I = kC$. For soils at or near steady state, total C and either losses or inputs can be measured to derive the decay constant.

*Soils in global CO~2~ cycle*

Several key questions exist per understanding the role of SOC in the global CO~2~ cycle:

  1. Are undisturbed soils sequestering atmospheric CO~2~?
    - They may be. Particularly for young soils that are not at steady state and may not be newly exposed to increasing temperatures and OM inputs (i.e., northern Canadian soils that were priorly glaciated).
  2. What is the effect of cultivation on SOC storage and atmospheric CO~2~?
    - Several key points are provided:
    - Soils initially low in C often will gain C upon cultivation.
    - At global scales, cultivation reduces SOC content by approximately 30%.
    - Rate of C loss is a function of climate
    - Agricultural soils approach new C steady states in approximately 100 years.
    - Fractional loss of C is controlled by climate.
  3. How quickly will soils recover C if reverted to original flora?
    - Initial rates are quite high, but drop off exponentially with time. Often times, inputs of OM to the soil system are limited by N availability.
    - Production and transport of N fertilizers can be greenhouse gas intensive and may reduce the GHG reduction potential of SOC inputs from both reverting ecosystems to native flora as well as enhanced agricultural management practices.
  4. How will global SOC change with increasing temperatures?
    - Only a fraction of SOC is cycled on time scales relevant for climate change, however it is expected that SOC emissions will increase with increasing temperature, and may be a significant positive feedback response for climate change.

### Bouillon, 2003 {#bouillon2003}

*"Sources of organic carbon in mangrove sediments: Variability and possible ecological implications"* [@bouillon2003sources]

**Key contribution: ** This study examines soil organic carbon at three different sites (one in India, two in Sri Lanka) and investigates various soil related parameters that may shed light on sourcing of organic carbon. The authors find high variability across the sites, and conclude that while at some sites mangrove contributions to SOC are very high, others may be dominated primarily by organic matter of marine origin.

This is one of the first studies to examine sources of organic matter in mangrove sediments.

**Key notes: **

Concentrations of organic carbon, total nitrogen, elemental ratios (C/N) and carbon stable isotopes were taken from soil samples to 5 cm depth, as well as from leaf litter in each of the sites. The geomorphology and species compositions of each of the sites were different.

C/N ratios of suspended particulate organic matter are typically much lower that those of mangrove litter and thus may act as an important source of N in sediment food webs.

*Study design*

They sampled soil and leaf litter as well as compose mixing models to investigate the relative contributions of mangrove derived organic matter versus marine derived organic matter.

Mixing models are a way of modeling relative contributions (often two end-members) to a system. In this case, it is the contribution of mangrove derived versus marine derived organic matter to total sediment organic matter.

*Key results*

Sediments underneath Rhizophora were found to have higher OC concentrations than Avicennia and Excoecaria species at two of the sites.

Lower OC concentrations coincided with lower C/N ratios and less negative C-13 isotope values, whereas soils with rich OC concentrations had higher C/N ratios and C-13 values closer to those of the mangrove leaf litter.

Appears there is an inverse relationship between OC content and C-13 values.

Although they note that organic C in mangrove sediment may originate entirely from mangrove-derived organic matter, they note that this is likely to be not the norm.

From the Coringa site, they find that suspended OM in the water column is significant and likely derived from phytoplankton-derived matter.

They describe the differences as "flow-through" systems versus "accumulation" systems, which may largely be a result of tidal amplitude (flushing) of system.

Mangrove litter has relatively high C/N ratio, and thus the influx of marine organic matter may be significant for provisioning of N rich material to microbial communities.

**Notes on stable isotopes**

Stable isotopes can be used for sourcing of organic matter and are commonly used in marine ecosystems.

Stable isotope compositions are typically expressed in "parts per thousand differences from a standard." I.e.:

$$\delta X = [(R_{sample}/R_{standard})-1]*10^{3} $$
Where:

  - X represents the isotope of interest (i.e., ^13^C)
  - R represents ratio of isotope of interest and its standard form (^13^C to ^12^C).
  
Higher values (less negative) represent increases in concentrations of the isotope of interest relative to the standard, whereas lower values represent decreases in isotope of interest relative to standard.

Carbon isotopes allow for understanding of primary production sources in an ecosystem. Transfer of ^13^C through trophic levels is relatively the same, except for slight increase (<1 ppt) in concentration.

### Krauss, 2003

*"Differential rates of vertical accretion and elevation change among aerial root types in Micronesian mangrove forests"* [@krauss2003differential]

**Key contribution: ** This study examines the differential effects of different root morphologies on sedimentation in mangroves. The study is summarized in Section \@ref(krauss2003).

### Lovelock, 2008 {#lovelock2008}

*"Soil respiration and belowground carbon allocation in mangrove forests"* [@lovelock2008soil]

**Key contribution: ** Cath sought to examine variation in mangrove soil respiration as a result of forest structure and temperature at seven different sites, from 27 deg N to 37 deg S.

Key findings were that soil respiration approximated that of terrestrial forests, and correlated with leaf area index and aboveground net primary production.

**Key notes: **

Lovelock examines soil respiration across a range of stands that represent a wide variety of forest structure, ANPP and climate. In doing so, she seeks to primarily answer:

  1. Whether soil respiration is higher in taller forests compared to dwarf forests, reflecting differences in aboveground productivity
  2. Whether total belowground carbon allocation (TBCA) in mangrove forests is related to ANPP with a similar relationship as that which exists in terrestrial forests
  3. Whether dwarf mangroves, which are nutrient limited, have proportionally higher TBCA than taller forests
  
*Study design*

Soil respiration was measured in conjunction with soil temperature at 2 cm depth, leaf area index, and tree height.

In addition, TBCA was calculated using a mass balance approach:

 $$ TBCA = F_{soil\ efflux} - F_{AG\ litter\ production} + F_{export} + F_{stored}$$
 
 Assuming that export and storage are low relative to inputs and respiration,  the mass balance is simply:
 
$$ TBCA = F_{soil\ efflux} - F_{AG\ litter\ production} $$

$F_{soil\ efflux}$ was estimated via extrapolation of soil respiration measurements.

*Key results*

Soil respiration was similar for both dwarf forests and taller forests, and was approximately equivalent to terrestrial forests (though slightly on the lower end).

Soil respiration also varied with forest LAI and soil temperature. Soil respiration increased to soil temperatures of approximately 25-27 deg C, and then decreased with further increases in temperature.

*Key findings*

Lovelock notes that the direct links between above- and below-ground processes facilitates scaling CO~2~ exchange measurements.

The correlation between LAI and soil respiration could be a result of variation in allocation of biomass to fine root components, or due to variation in heterotrophic component of respiration across sites.

LAI may be useful for estimating soil respiration over large spatial scales as it can be remotely sensed.

High rates of soil respiration in dwarf forests relative to taller forests are unlikely to be due to respiration of live roots as live root densities, root growth rates and root respiration rates are low in dwarf forests. Rather, may be due to high rates of C exudation and high levels of heterotrophic respiration.

### Lunstrum, 2014 {#lunstrum2014}

*"Soil carbon stocks and accumulation in young mangrove forests"* [@lunstrum2014soil]

**Key contribution: ** This study examines contributions of organic carbon to soil systems following mangrove afforestation of two species: Kandelia obovata and Sonneratia apetala. In addition, they examine a chronosequence of forest at different ages, and a literature review of mangrove inputs to soil organic carbon to investigate the influence of mangrove growth and species composition in soil organic carbon.

**Key notes: **

The authors hypothesize that given the realtively higher root to shoot biomass ratio in mangroves, soil C stocks will increase soon after afforestation.

They use primarily two key datasets:

  1. Data consisting of repeated measures of surface soil C concentration in two young forests (K. obovata and S. apetala) from 0 to 6 years old
  2. Chronosequence of 1 m soil cores from a mudflat, a 6 year, a 20 year, and a 70 year old forest.
  
Using the two datasets, they test the hypothesis that:

  1. Soil C increases with forest age following mangrove forestation
  2. Soil C accumulates more quickly in forested S. apetala than K. obovata plots

*Chronosequence results*

The growth of S. apetala and K obovata differed greatly. S. apetala reached 7.6 m whereas K. obovata reached just 1.4 m. The stem density between the 20 year and 70 year old forests were markedly different, with the 70 year old forest (dominated by K. obovata) stocked more densely.

Soil C generally decreased with depth in all forests. Younger forests convered towards the mudflat values below 30 cm.

The 70 year forest had the highest SOC concentration at 10-20 cm, whereas the lowest value was found at the 70-100 cm level in the 6 year forests.

N content was also found to be higher in the 70 year old forest, though the difference was not as pronounced.

SOC differences between the three young forests and the mud flat was not significantly different, whereas the 70 year old forest was significantly different from the mud flat. THis might indicate that soil C accumulation accelerates in older forests.

*Field study results*

Soil C concentration increasesd with age in the young forests (increase of approximately 0.06% per yr for K obovata and 0.08% per yr for S apetala).

Mangroves are often cited as having high average POC at 8.5% (Duarte et al., 2005), but Alongi, 2014 revised this estimated closer to 2-3% C.

*Literature results*

The literature results also showed a significant increase in SOC with age; however, the mean estimates of soil carbon percentage were lower than commonly cited values.

*Discussion*

Terrestrial forests are commonly cited as losing SOC upon afforestation, occasionally taking 20-30 years to recover initial values. Given the evidence here and the literature review, the results indicate that mangroves may behave differently and provide immediate SOC sequestration benefits.

The hypothesis that S. apetala and K. obovata would differ in SOC accumulation was not supported by the study.

Relative C:N concentrations in the biomass of different species could influence differences in SOC in forests of different species compositions, as the nutrient quality would affect the rate at which biomass is decomposed.

### Atwood, 2017 {#atwood2017}

*"Global patterns in mangrove soil carbon stocks and losses"* [@atwood2017global]

**Key contribution: ** The study provides an updated estimate of SOC patterns globally, through the lens of national scales. 

**Key notes: ** They describe the data quality (relatively poor) for SOC, although their database is likely less extensive than that of the ERL paper. They note the widespread prevalence of pedotransfer functions to estimate bulk density, as well as a common lack of SOC estimates deeper than 50 cm.

*National and global trends:*

Found a pattern in mean SOC per ha across latitudinal bands, but did not find total SOC to correlate with latitudinal change.

Found mixed species stands to have higher SOC than monospecific stands; stands with 5 genera had 70-90% higher soil C stocks than other richness levels. There are implications here, particularly for restoration efforts which typically focus on Rhizophora or Avicennia species.

Almost all variation at national level was explained by extent of mangroves, which points to the need for improved methods for estimating and mapping mangrove area.

Data did not exist for 44 countries in which mangroves occur.

They did not examine variation in mean SOC density across nations or global scales.

Indonesia, Malaysia, USA, and Brazil have the highest potential C emissions from deforestation, accounting for a total of approximately 86% of stocks. 

Other countries with very high density (higher than these four) currently have low rates of deforestation, and thus may constrain total C emissions from mangroves. Prioritizing conservation of mangroves in these countries may be critical.

Myanmar may be a highly significant country in terms of C emissions and deforestation.

## Other

### Ferreira *et al*, 2007 {#ferreira2007}

*"Redox processes in mangrove soils under Rhizophora mangle in relation to different environmental conditions"* [@ferreira2007redox]

**Key contribution: ** Upon recognizing that redox conditions may be highly variable over space and across different environmental settings, this study seeks to examine redox patterns in pure Rhizophora stands across three different environmental settings. They find that different redox patterns occur, and that particular characteristics that may be associated with particular settings (e.g., degree of periodic oxidation or soil texture) may greatly influence redox potentials.

**Key notes: **

Bacterial reduction of sulfate is commonly considered main form of respiration in tidal environments, but recent work has shown that the reduction of Fe may also be very important in mineralization of organic matter.

*Study design*

Pure Rhizophora forests in three different environmental settings (fringe, riverine and basin) were selected for examining soil conditions and redox conditions.

Half-meter cores were taken with 6 replicates, and pH and E~h~ were measured in the field.

Cores were preserved and one was cut into fine-scaled sections of variable depth, whereas the other five were cut into two depth sections (0-10 & 10-30 cm).

Pore water analysis was conducted after sampling and from the middle of each soil sample. HS^-^ and Fe^2+^ concentrations were determined.

Soild-phase Fe was partitioned to determine the contributions of seven operationally derived Fe pools:

  1. Exchangeable Fe
  2. Carbonate-associated Fe
  3. Ferrihydrite Fe
  4. Lepidocrocite Fe
  5. Crstalline oxides 
  6. Fe oxyhydroxides (goethite, hematite)
  7. Pyrite Fe

Total organic carbon, total S, total Fe, and percentage sand were determined for each soil sample.

*Results*

The different environmental settings produced differences in soil composition and physicochemical conditions.

Redox potential values decreased significantly with depth in all studied profiles.

A highly significant correlation between pH and E~h~ existed, with lowest pH values corresponding to the highest redox potentials.

For Fe^2+^ and HS^-^ concentrations, high variation over both space as well as with depth were observed. Concentrations of HS^-^ increased significantly with depth, whereas highest values of Fe^2+^ were found in surficial layers.

Iron partitioning was found to be variable between the fringe site versus the riverine and basin sites: 

  - Fringe site - Within the fringe site, Fe oxyhydroxides predominated, especially in surficial layers. Pyrite at this site increased notably with depth.
  - Basin and riverine sites - Relatively lower amounts of Fe oxyhydroxides existed. Pyrite was much higher in the riverine site both at surface and at depth, whereas pyrite approximated Fe oxyhydroxides at surface but pyrite predominated at depth

Total exchangeable Fe was small across all sites and not considered in the study.

*Discussion*

The author attribute the spatial variation in geochemical conditions to different periods of inundation in which soils are exposed, response to positions within the mangrove forest, and the composition (texturally and chemically) of the substrate.

  - Fringe forest - convex profile and high density of roots induces rapid drainage, aeration and oxidation in low tide; extremely acidic conditions in surface soils coincide with tidal cycle. High tide induces an increase in pH due to diluting and buffering effect of sea water, which favors precipitation of dissolved Fe as Fe oxyhydroxides.
  - Basin forest - strongly reduced conditions due to long periods of inundation from concave profile; anoxic conditions within the profile gavor syntehsis of pyrite rather than Fe^2+^ and HS^-^ in pore water. Predominance of pyrite synthesis is due to high availability of OM, which maintains activity of sulfate-reducing bacteria. 
    - High rates of sulfate reduction may occur with presence of reactive Fe, as different species of Fe oxyhydroxides have different surface areas (goethite and hematite are reduced), that may limit actions of microorganisms.
  - Riverine forest - highly reduced despite position along river. Total Fe and free Fe was very low relative to the other sites, and was consistent with sandy, quartz-based texture. Thus, sulfate reduction was primary reduction pathway of OM.
  
*Conclusions*

The dominant redox pathways are highly variable of space and with depth. Occurrence or dominance or a particular process over others may significantly affect chemical composition of pore water and solid phases.



