# Soils & geomorphology

## Background

### Soil formation

#### Jenny, 1994 {#jenny1994}

*"Factors of soil formation: A quantitative system of pedology"* [@jenny1994factors]

**Key contribution:** This is Hans Jenny's seminal work on the factors of soil formation. The most significant contribution of the work is his conceptual model for soil formation, which is shown below:

  $$soils = f(climate, organisms, relief, parent material, time)$$
The model allows for conceptual investigation of the various factors that influence the formation of a soil "system." Jenny defined soils as "systems" that were largely influenced by initial states (relief, parent material, time) as well as their surroundings (climate, organisms).

#### Amundson, 2003 {#amundson2003}

*"Soil formation"* [@amundson2003soil]

**Key contribution:** 


#### Amundson, 2014 {#amundson2014}

*"Soil formation"* [@amundson2014soil]

**Key contribution:** This is an extensive review on the formation of soil that extends Amundson's earlier work in the first volume of the *Treatise on Geochemistry*.

**Structure:** The chapter is structured as follows:

  1. Overview of what a "soil" is
  2. Conceptual partitioning of the earth surface
    1. Stable landforms
    2. Erosional landscapes
  3. Human dimensions of soil formation
  4. Soil chemistry in deserts
  5. Soil formation on Mars

**Notes: ** This chapter should be reviewed several times, as it is indicative of Ron's current thinking on soil formation processes.

*Conceptual issue of what constitutes a "soil"* - different views of soil thickness among differing scientific communities. For example:

  1. Earth sciences - primarily concerned with the mobile layer of weathering profile in which material no longer reflects parent rock (often restricts to A horizon as B horizon still has parent material)
  2. Geochemical perspectices - interested in the whole weathering process and thus interest extends to more considerable depths

Import to consider: What is the depth at which I am interested for mangroves?

*Soil data access* - There are several platforms for which rich data on soils in the US can be collected:

  1. Web soil survey - portal for soil maps and lab data
  2. Soil-web - accesses same database as Web Soil Survey, but uses a different architecture to interface with key technologies (Google Earth, Maps, & smartphones)
  3. Digital soil map of the world -- currently under development but needs more effort (see Sanchez, 2009, Science)
  
*Soil formation* - Can very abstractly be thought of in three classes:

  1. Stable landforms, where derivative of slope = 0
  2. Erosional landforms, where derivative of slope is negative
  3. Depositional landforms, where derivative of slope is positive

**Stable landforms** - Depositional processes can be understood largely through examination of elements relative to that of a parent material. Often this is done by normalizing that element relative to an element that is immobile in the soil weathering environment (Zr, Ti, Nb, etc.).

Often use *chronosequences* to examine changes over time due to inability to perform experiments over extremely long time scales.

Some characteristic steps of soil formation in stable landscapes are outlined:

  - Primary inputs to soil come from organisms; fixing of atmospheric CO~2~ and NO~3~ and NH~4~ bound with amino acids
  - Biotic mechanisms take place and alter the sedimentary structure of the soil (physical mixing)
  - Through chemical weathering, accumulation of secondary clay minerals in the B horizon occurs
  - Clay accumulation and soil thickening occur over long timeframes, volume decreases and bulk density increases resulting in relative "collapse" of soil
  
Whether you consider just the A or B horizon depends on discipline and particular viewpoints.

Within San Joaquin Valley, the "mound and swale" topography is due to the relative increase of clay which creates a higher water table and reduces the A horizon of soil placing pressure on gophers. In response, gophers are hypothesized to undertake directional movement of soil to the tops of mounds.

  - Note, this is the state soil of California (San Joaquin soil series)
  - "Fine, mixed, active, thermic Abruptic Durixeralfs"
  
Important point is that rates of mineral dissolution in almost all soils are transport limited in that it is not the resistance of elements or materials that determine rates, but rather flow of liquids.

Thus, the rate of weathering is relatively constant with age and thus may have weathering fronts that exist at very considerable depths.

*Weathering front* - the boundary between the parent rock and either weathered rock or regolith

**Erosional landscapes** - One of the most important conceptual changes in soil formation within last fifteen years is through modeling of soil transport on hillslopes, particularly through coupling of physical processes with geochemical ones (see Dietrich et al., 2003).

*Denudation* - processes causing wearing away of Earth's surface via moving water, ice, wind or waves; leads to reduction in elevation and relief

The chemical weathering rate can be defined as the difference between the total denudation rate and the physical erosion rate.

  $$ Chemical\ weathering = Total\ denudation - Physical\ erosion$$

The position of the weathering front will depend upon the relative balance between the denudatoin rate and the propagation rate of the weathering front. If denudation (soil loss) exceeds that of weathering front propagation rate, then the front cant be expected to be shallow, and vice versa.

Denudation is apparently insensitive to rainfall for vegetated landscapes, where as weathering front propagation rate is highly sensitive to levels of rainfall.

*Human influences* - The scale at which humans are influencing soils is massive. When defining "endangered soils" as those soil series that have lost at least 50% of their extent, considerable portions of soils in the US have been affected (particularly in the mid-West), but also much of the central valley and along the E. coast.

The effects of mechanical alteration of soils as well as through the addition of high amounts of water in arid or semi-arid landscapes greatly accelerates chemical weathering processes (e.g., removal of salts, altering of pH, secondary clay formation, etc. Additional research is needed on the effects of agriculture on undisturbed soils.

*Soil geochemistry in deserts* - Use of isotope research in arid soils reveals that *stable isotopes in soils are rarely conservative tracers of element source* (is this for all soils? what does he mean by this?), but rather reflect complex postdepositional processes caused by aqueous or gaseous transport.

An important interpretation of this is that C isotope ratios of soil CO~2~ (and thus any carbonate that forms from it) largely reflects the isotope ratio of plant respiration and the concentration of atmospheric CO~2~. Additionally, C isotope ratios are highly variable along the depth profile.

*Soil formation on Mars* - Investigation of soils in the top 10-15 cm of the surface of mars allows for interesting insight into soil formation and development processes in the absence of water; however, the conclusions of existing work conclude that much of the observation were driven by prior existence of liquids on the surface.

### Soil classification

Classifying soils includes chemical, physical, and biological properties that may be directly observable in the field or quantifiable via lab analyses. The soil taxonomic name that is associated with a particular soil series is built to reflect each of these characteristics.

**Key terms: **

*Pedon* - Smallest three-dimensional sampling unit that displays the full range of characteristic properties of a particular soil

*Polypedon* - A group of closely associated pedons that are commonly found adjacent to another within the field (i.e., slight variations in a one general pedon). A polypedon forms a *soil individual*.

*Soil series* - A class of soil individuals that exhibit the same properties and horizons, though they may be disjunct in space or time

*Diagnostic soil horizons* are often associated with precise quantitative measures and are used to resolutely identify a soil as particular series or not

Diagnostic soil horizons may occur at the surface or below the surface of the soil. Horizons that occur at the surface of the soil are termed *epipedons*. Seven epipedons exist:

  1. *Mollic* - Dark color rich in organic matter
  2. *Umbric* - Same general characteristics as mollic, but base saturation is much lower
  3. *Ochric* - Mineral horizon that is too thin and light in color to Mollic or Umbric; may be very hard when dry
  4. *Melanic* - Black, organic matter rich horizon characteristic of soils derived from volcanic ash
  5. *Histic* - 20 - 60 cm thick layer of organic soil materials overlaying mineral soil; characteristic of wet areas such as peats
  6. *Anthropic* - Same limits as Mollic epipedon, but has formed through extensive human use (e.g., agricultural soil)
  7. *Plaggen* - Human-made surface layer approximately 50 cm thick made from extensive manuring

The *diagnostic subsurface soil horizons* consist of:

  1. *Argillic* - subsurface accumulation of high-activity silicate clays that have either precipitated downwards or formed in place
  2. *Natric* - silicate clay accumulation but with high exchangeable sodium on the clay complex; common in arid or semi-arid regions
  3. *Kandic* - accumulation of Fe and Al oxides as well as low-activity clays (kaolinite)
  4. *Oxic* - highly weathered subsurface horizon very high in Fe and Al, oxides and kaolinite; found in tropics and subtropics
  5. *Spodic* - illuvial horizon (material displaced by rainwater from one horizon to another) characterized by accumulation of colloidal organic matter and Al oxides; common in highly-leached, sandy cool humid climates (think pine forests)
  6. *Sombric* - illuvial horizon dark in color due to high OM accumulation; found mostly in cool, moist plateaus and mountains of tropics and subtropics
  7. *Albic* - light-colored eluvial (formed in place) low in clay and Fe and Al oxides which have been transported downwards
  8. *Calcic* - accumulation of carbonates (white chalk-like nodules)
  9. *Gypsic* - accumulation of gypsum
  10. *Salic* - accumulation of soluble salts
  
*Soil moisture regimes*:

  1. *Aquic* - soil saturated with water
  2. *Udic* - moisture sufficiently high to meet plant needs
  3. *Ustic* - intermediate moisture between udic and aridic; significant periods of drought may occur
  4. *Aridic* - soil is dry for at least half of the growing season and moist for less than 90 days (torric indicates the same soils that may be hot in summer but not hot in winter)
  5. *Xeric* - moisture regime common to Mediterranean climates; cool, moist winters and warm, dry summers
  
The *soil temperature regimes* primarily refer to:

  1. Frigid (cold)
  2. Mesic (moderate) 
  3. Thermic (hot)

**Soil taxonomy:**

  1. Order - Defined as one of 12 orders
  2. Suborder - Often integrate the moisture regime of the region (e.g., Aquent)
  3. Great group - Often indicate presence of a diagnostic horizon (e.g., Sulphaquent)
  4. Subgroup - Further define diagnostic horizon to add nuance to the classification (e.g., Halic Sulphaquent)
  5. Family - identifies subsets of subgroups that are similar in texture, mineral composition and mean soil temperature at depth of 50 cm
    - e.g., "Typic Argiaquolls, fine, mixed, mesic, active"
    - Typic Argiaquoll that is fine in texture, mixed clay mineral content, mesic in temperature (8 - 15 deg C), and with clays active in cation exchange
  6. Series - named after geographic feature where they were first identified
  
There are 12 *soil orders* globally:

  1. *Alfisols* - argillic, natric or kandic horizon; high to medium base saturation (alf)
  2. *Andisols* - soil formed from volcanic ash (and)
  3. *Aridisols* - dry soils with ochric epipedon (id)
  4. *Entisols* - little profile development, ochric epipedon is common (ent)
  5. *Gelisols* - permafrost; frozen soils (el)
  6. *Histosols* - peatland or waterlogged soils with > 20% OM (ist)
  7. *Inceptisols* - embryonic soil with few diagnostic features; ochric or umbric epipedon (ept)
  8. *Mollisols* - grassland soil rich in organic matter with mollic epipedon (oll)
  9. *Oxisols* - oxic horizon, highly weathered (ox)
  10. *Spodosols* - spodic horizon common with Fe and Al oxides (od)
  11. *Ultisols* - argillic or kandic horizon, low base saturation, very weathered (ult)
  12. *Vertisols* - high in swelling clays, cracks when dry (ert)
  
Mangrove soils are likely to be *Aquents* (frequent saturation with water) or Fluvents (parent material of recent alluvium, more commonly associated with floodplains)

  - likely to be "halic sulphaquents," "sulfic halaquents", or "halic sulphaquepts"
  
*Suborders* commonly integrate the moisture regime of the region into the soil order (e.g., Aquent).

More than 400 *Great Groups* exist, defined primarily by presence of diagnostic horizons, which supply formative elements for the names (suphic diagnostic horizon in Sulphaquents, halic diagnostic horizon in Halaquents)

*Subgroups* - central concepts of a Great Group makes up one subgroup (*Typic*). Others introduce secondary diagnostic characteristics that better situate the soil between different Great Groups. (e.g., Sulfic Halaquents as an Aquent with Halic diagnostic properties but also Sulfuric properties).

*Families* - Approximately 8,000 families have been identified; integrate information on the physical and chemical properties affecting the growth of plant roots.

  - Information on particle size, mineralogy, CEC activity, temperature, and depth of soil

### Soil architecture

**Soil color:** Soil color is classified using the Munsell charts and is a function of:

  1. Hue - Redness or yellowness
  2. Chroma - Lightness or darkness
  3. Saturation - Intensity of brightness of the color
  
Soil color is indicative of many properties of the soil, including presence of organic matter, moisture content, or presence of Fe or Mn oxides.

*Gleying* refers to reduction of Fe or Fe depletion from the soil and is indicated by a gray soil color.

**Soil texture:** Soil texture ranges greatly (e.g., from clay to boulders), but often we refer to the fine-earth fraction (sand, silt & clay) in referring to soil texture.

  1. *Sand* - Smaller than 2 mm in size but greater than 0.05 mm; most common material constituent is quartz (SiO~2~), or other silicate materials
  2. *Silt* - Between 0.05 mm and 0.002 mm; Similar in shape and mineral composition to sand but much smaller in size and thus drains water more slowly
  3. *Clay* - Less than 0.002 mm; Extremely large surface areas which results in great propensity to hold water
    - The finest sizes of clays act as colloids (i.e., where particles suspended in a liquid do not readily settle out)
    - More commonly flat and plate-like than round which further inhibits passing of liquids and gasses

*Textural classes* - Soils exist on a continuum of texture, but can classified generally given their relative percent composition of sand, silt or clay

Central concept of a *loam* is defined as a mixture of sand, silt and clay that exhibits properties of those separates in approximately equal proportions.

Adding peat or compost to soils does not technically change the texture, as the texture criteria is only referring to the mineral particles.

**Soil structure:** Soil particles aggregate together to form *peds* or *aggregates*, which greatly influence the structure and functioning of the soil.

*Peds* may vary in size from 1 mm to 1 m; Commonly used to refer to the whole structure of the soil (as aggregates exist at different scales), but is synonymous with "aggregate". Peds often vary depending on the horizon in which they are found. The principal ped shapes are:

  1. Spheroidal
  2. Platy
  3. Prismlike
  4. Blocklike
  
Soil aggregates consist of macroaggregates that are a combination of microaggregates. Both biological and physical-chemical processes are important in their formation, though PC processes matter primarily for microaggregates whereas biological processes are more critical for macroaggregates.

*Flocculation* is the mutual attraction between clay and organic molecules, and is key for formation of microaggregates (particularly for sedimentation in mangroves).

  - Form through a covalent bond in which positive charged ions compressed against one another attract negative charges on the opposites platelet.
  
Organisms (e.g., earthworms) and plant roots are also important in forming of casts through consumption of soil as well as production of sticky substances (glomulin) that are important in forming aggregates.

**Soil density:** There is a key distinction between particle density and bulk density that should be noted. In particular:

  - Particle density is defined as the mass of the soil particles per unit volume of *soil solids* (i.e., no pore space is considered)
    - Particle densities vary little and are typically between 2.6 - 2.75 Mg/m^3^
  - Bulk density  is the mass of the soil particles per unit volume of soil (including the pore spaces)
    - Bulk density values vary to a much larger degree
    
One of the key reasons for measuring bulk density is for the calculation of *total porosity* (i.e., through comparison of particle density and bulk density). Provides a measure of total pores, but pores are comprised of both *macropores* (> 0.08 mm) as well as *micropores* (< 0.08 mm).

  - *Macropores* readily allow movement of gas and water within the soil, as well as growth of plant roots and movements of soil organisms.
  - "Micropores" are typically filled with water. Even when not filled with water, they are small enough to restrict passage of gasses. Water in these pores is restricted to plants

### Soil hydrology

**Key notes**:

*Properties of water molecules* - Ability of water to influence soil processes is largely due to the structure of water. In particular:

  - Polarity - The H atoms of water line up in a V-shape of 105 degrees which induces a polarity to the molecule.
  - Hydrogen bonding - Hydrogen atom of one water molecule is attracted to the H of another water molecule forming a low-energy bond that results in polymerization of water. Hydrogen bonding thus accounts for the following behaviors of water
    - Cohesion - Adherence of water molecules to one another
    - Adhesion - Adherence of water molecules to the surfaces of solid particles
    - Tension - Results from greater attraction of water molecules to one another than to gases, which results in tight bonding of water molecules at the air/water interface
    
Capillary action is important for the movement of water within soils. Capillary action occurs through the dual action of adhesion and cohesion.

*Soil water energy concepts* - Various forces come into play in determining movements of water in the soil.

Water transfers from states of high energy to low energy, and various forces that determine the potential energy of water are important:

  - Matric forces -- forces caused by adhesion that are responsible for adsorption and capillarity.
  - Osmostic forces -- forces driven by differences in ion concentrations in solutes
  - Gravity -- self explanatory; constant downwards force on water
  
*Water potential* is determined as the difference in potential energy between water within a given soil at a particular location with water at a reference location at standard pressure and temperature with no effects of soil.

Total soil water potential is a function of:

  1. Gravitational potential
  2. Matric potential -- extremely important for supply of water to plant roots
  3. Hydrostatic potential -- pressures from the weight of water in saturated soils
  4. Osmotic potential 
    - Important for mangroves
  5. Other more minor potentials...
  
Matric and hydrostatic potentials are part of the "pressure potential," whereas gravitational and osmotic are considered alone.

*Soil water content and water potential* - An inverse relationship exsists between the water content of soils and how strongly water is held within soils.

In describing water content of soil, need to describe both i) the water content as well as ii) the water potential. Behavior of water in soil is typically determined primarily by water potential rather than water content.

"Water content" is defined as a volume of water per volume soil. Typically measured as difference between wet and dry weights of soil.

"Water potential" can be measured in different ways; Matric potential can be measured via a tensiometer, which is a semi-porous rod filled with water that allows water to flow out of the rod and creates a vacuum within the rod. Tensiometers are useful over a range of 0 to -85 kPa.

*Flow of water* - Flow can be broken into three primary types:

  1. Saturated flows - Saturated flow occurs when soil pores are completely filled. Hydrostatic potentials play a part here and the hydraulic conductivity is largely determined by macropore spaces
    - Flow rates are proportional to the fourth power of the radius, and thus macropore space greatly determine flows during saturated flow.
    - Preferential flow is the faster flow of water through biopores (e.g., pores and channels created by organisms) or large cracks
  2. Unsaturated flows - Large pores are free of water but smaller pores hold water
    - Primary driver of water flow is "matric potential gradient"
    - Soil texture is highly important as clays will moderate flows whereas sandy loams are more "flashy"
  3. Vapor movements - Result of vapor pressure differences in relatively dry soils
  
*Infiltration and percolation* - Texture of soil and in particular presence of macropores are critical here.

*Qualitative descriptions of soil wetness* - Several classes of soil wetness can be described:

  1. Maximum retentive capacity - all pore spaces are filled, matric potential is close to zero and gravitational forces dominate
  2. Field capacity - soil will drain via gravitational forces within 1 - 3 days, and matric forces will dominate (generally a range of -10 to -30 kPa). Field capacity is an important condition given:
    1. Maximum level of water available to plants is present
    2. Soil is at optimal "pliability" in that below field capacity it will begin to crumble and above that it will easily turn to mud
    3. Sufficient pore space is available to allow aeration for microbial activity and plant growth
  3. Plant available water - from 10 - 30 kPa to 1500 kPa - water potentials at which water is largely available to plants
  4. Permanent wilting percentage - point at which plants wilt during both day and night due to inability to extract water from soil (<1500 kPa)
  5. Hygroscopic coefficient - point at which water is only available in vapor form due to extreme aridity

The amount of organic matter within a soil can influence the amount of plant available water greatly. Organic matter has a high water-holding capacity, particularly relative to mineral soils.

"Least limiting water range" - Saturation of soil may produce anoxic conditions which restrict root growth. Often times this is the point at which only 10% of pore space is filled with air. Additionally, soils may become too dry for plant roots to push through them during growth (2000 kPa)

*Relevance for mangroves* - My own interpretation of how this might matter for mangroves is as follows:

For mangroves, matric potentials and osmotic potentials likely drive water flows within mangroves. Given the extremely find sediment of many mangrove forests, there is likely very little pore space for which water and air may enter the soil. Water flushing and aeration may only exist within top 15-20 cm of soil.

### Soil aeration

Aeration is the process of soil ventilation, with gases moving both into and out of the soil. Requires both the supply of oxygen as well as the removal of CO~2~ from the soil.

Three primary processes control the availability of oxygen within soils:

  1. Soil macroporosity
  2. Soil water content
  3. O~2~ consumption

Plants that have adapted to waterlogged soils (i.e., in wetlands) are *hydrophytes*, and have *aerenchyma* which are hollow tissues used for the transport of oxygen to the roots.

Aeration of soil occurs through two primary mechanisms:

  1. Mass flow -- much less significant on total exchange of gases
  2. Diffusion -- most significant process; gases moving in response to their partial pressures
    - partial pressures is the pressure a particular gas in a gas mixture will exert
    - O~2~ will generally move into soil whereas water vapor and CO~2~ will typically move from soil to the atmosphere
    
There are three primary ways of characterizing aeration of soil:

  1. Directly describing the gas constituents in the soil
  2. Air-filled soil porosity (i.e., percentage of soil pores that are filled with air)
  3. Redox potential as a measure of the oxidation and reduction of different elements (i.e., a measure of aerobic metabolism by organisms)
  
Reduction-oxidation (redox) describes the chemical processes by which electrons are transferred. Reduction and oxidation occur together, as electron is transferred from one (the reductant) to the other (the oxidant). Micro-organisms drive majority of redox reactions, often consuming electrons from C or accepting electrons from oxygen. Respiration is largely the consumption of C based electrons from carbohydrates.

Basic example of respiration in which electrons are derived from an organic molecule:

$$CH_{2}O + H_{2}O = CO_{2} + 4e^-$$

The electrons cannot exist in independent form, and thus are run through metabolic cycles in organisms and dumped onto electron acceptors (e.g., O~2~).

Oxygen is important within soils as it rapidly accepts electrons from other elements, which allows for metabolic processes. Aerobic respiration is performed on O~2~ (oxidize organic carbon) and release energy for metabolic activity.

The redox potential of a soil is a measure of how well much oxidation-reduction it is able to perform (i.e., a measure of its ability to accept or donate electrons), and depends upon presence of electron acceptors as well as pH. Redox potential typically decreases as the pH increases.

*Oxidizing soils* - accept electrons easily
*Reducing soils* - donate electrons easily

Standard Eh levels for drained, well-aerated soils is in the 0.4 - 0.7 Volt range, whereas flooded conditions can reduce Eh levels to as low as -0.3 V.

Other electron acceptors besides oxygen are:

  1. Nitrate - accepts 2 electrons when reduced from N(V) to N(III); reduces to N~2~
  2. Manganese (VI) -> reduces to Mn(II)
  3. Iron (III) -> reduces to Fe(II)
  4. Sulfur (from SO~4~^2-^ to S~2~^-^ (reduced))
  5. Carbon (reduction of CO~2~ produces CH~4~)
  
Electron acceptors are typically used up in this order (i.e., sulfur reduction is a process that occurs largely under very anaerobic soils). Anerobic systems result in the build up of reduced species

The form of particular elements is important because it will change availability for uptake by plants (i.e., oxidized vs. reduced states). 

*Factors that affect soil aeration:*

  - Presence of macropores -- most important factor in well-drained soils; function of soil texture, bulk density, aggregate stability, OM content, and biopore formation
  - Microbial activity -- rates of respiration in the soil will greatly influence gas composition, and is dependent upon availability of organic matter
  - Soil heterogeneity -- lower oxygen concentrations in subsoils; often due to slow diffusion, reduced pore space (from compaction), and greater proportion of pore space filled by water
    - plant roots may reduce or increase O~2~ content (diffusion of O~2~ from root aerenchyma)
  - Seasonal differences
    - Changes in precipitation patterns or temperature
    
*Ecological effects of soil aeration:*

  - Nature and rates of microbial activity are determined largely by O~2~ concentrations
  - Redox of elements:
    - Determines form and thus availability of elements, which may be beneficial or toxic
    - Methane production, which occurs at Redox potentials of approximately -0.2
  - Activities of higher plants:
    - Root growth may be curtailed
    - Absorption of water and nutrients may be decreased
    - Formation of certain inorganic compounds may occur
  
*Soil temperature*

Growth rates of plants are more strongly controlled by soil temperature than the air temperature. Biological zero is 5 deg Celcius, in which microbial activity virtually ceases.

Impacts of frost-heave and fire on soils can be significant. Effects of fire typically limited to the top 1-2 centimeters, but can be intense enough to alter organic matter, soil conditions, and kill soil organisms.

Temperature of soil in the field is largely dependent upon three factors:

  1. Net amount of heat energy the soil absorbs
  2. Heat energy required to bring change in the soil temperature
  3. The energy required for processes such as evaporation, which occur primarily near soil surface
  
Only approximately 10% of solar incidence is actually transferred to soil. Majority is intercepted within atmosphere, by vegetation, or is loss in the evaporation of soil water.

Dark soils do not necessarily warm faster than light soils due to the fact that dark soils may often be very moist and thus solar incidence is lost through evaporation of soil moisture.

*Thermal properties of soils*

The specific heat (energy required to raise the temperature of a given mass of something by 1 degree C) is much higher for water than for mineral soil solids. Thus, evaporation of water requires a very large amount of energy (2.257 MJ per kg of water).

Thermal conductivity refers to the rate at which heat is tranferred through soil, and is dependent primarily on:

  1. Soil moisture content - heat passes through water much faster than through air
  2. Degree of soil compaction - heat passes through mineral particles even faster than through water, thus when compaction and particle-to-particle contact is high, heat is transferred quickly
  
Wet compact soils are the poorest insulators (i.e., best conductors of heat)

Given seasonal and daily variation in temperatures and the low thermal conductance of soils, there is a temporal lag between the soil temperature and atmospheric temperatures 

### Soil chemistry

The soil colloids are arguably the most important component of soil chemistry, as they control the majority of exchange of cations and anions within the soil.

The colloidal fraction of soil refers to the clays and humus particles in soil because of their extremely small size (< 1 micrometer) and colloid-like behavior.

The very large surface area to volume ratio of soil colloids is due to the platelike structure, and the present of internal surfaces within clays. The majority of soil colloids primarily carry more negative electrostatic charges than positive charges.

*Absorption of cations and anions:*

Colloids attract thousands of Al^3+^, Ca^2+^, Mg^2+^, K^+^, ^H+^, and Na^+^, which are loosely bound to the surface of the colloid, often in a hydrated state in moist soils.

"Cation exchange" is the process in which one cation breaks free from the colloid, is released into solution, and a different cation takes its place. The ions that are readily lost from the colloid surface are referred to as "exchangeable ions".

The ionic composition of soil solution (i.e., the water) resembles that of the adsorbed ions (those bound to the colloid) due to the constant exchange between the two.

Exchangeable anions are also present in the soil solution, and will exchange with soil colloids with positive charges on their surfaces.

*Types of soil colloids:*

  - Crystalline silicate clays -- dominant in most soils; layered structure like pages in a book; predominantly negatively charged; different silicate clay minerals vary widely in particle shape, however, which influences intensity of charge, plasticity, swelling behavior, etc. 
    - Examples: smectite, vermiculite, fine mica, chlorite, kaolinite (1:1 silicate)
  - Noncrystalline silicate clays -- tithgly bonded silicon, aluminum and oxygen atoms, but do not exhibit crystalline sheets; high amounts of positive and negative charge, and capable of holding large amounts of water, but exhibit little stickiness
    - Principle types: allophane and imogolite (both form from volcanic ash)
  - Iron and aluminum oxides -- found widely but are most important in weathered soils of sub-/tropics; slightly negative to moderately positive and are relatively low in plasticity and stickiness
    - Examples: gibbsite (Al-oxide) and goethite (Fe-oxide) which form crystalline sheets; others are noncrystalline
  - Organic (humus) -- convoluted chains and rings of carbon atoms bonded to H, O, and N; very high capacities to hold water, but little to no plasticity and stickiness
  
A summary of the major colloids and their properties is shown below:

```{r}
knitr::include_graphics(rep("images/colloids.bmp"))
```

*Layer silicate clay structure:* 

Silicon tetrahedral and Aluminum-Magnesium octahedral sheets:

  - Silicon tetrahedral sheets are comprised of a silicon ion surrounded by four oxygen atoms
  - Al-Mg octahedral sheets are comprised of an Al or Mg ion surrounded by 8 oxygen-hydroxy planes
  
Two to four of the tetrahedral and octahedral sheets will bind together by sharing some of the same oxygen atoms, which results in extremely tight bonds and forms a "layer". The layers then are bound together via adsorbed ions and water, which forms the clay "crystal".

The nature and combination of sheets in the layers vary, and thus produce the differences in types of clays (ant thus their resultant physical and chemical properties).

"Isomorphous substitution" is the process by which cations of comparable size to Si, Al, or Mg are substituted for the center of one of the tetrahedrons or octahedrons without significantly changing the structure of the crystal. This is the key process that induces variability in nature of silicate clays.

Isomorphous substitution is the primary cause of charge imbalances in colloids (e.g., through the substituion of a Mg^2+^ for a Al^3+^ ion).

Key difference is between 1:1 and 2:1 silicate clays:

  1. 1:1 silicate clays contain one tetrahedral and one octahedral sheet each in their layers
  2. 2:1 silicate clays contain one octahedral sheet in between two tetrahedral sheets within each layer

Kaolinite is the classic example of 1:1 silicate clays. Bonding between layers (1 tetra and 1 octa) is via hydrogen bonding, which prevents expansion when the clay is wet. Cations and H~2~O typically do not enter the structural layers of a 1:1 mineral particle and thus effective surface of kaolinite is primarily the external surface area.

Expanding 2:1 type silicate clays are formed via one octahedral sheet in between two tetrahedral sheets (e.g., smectite and vermiculite). Fine micas and chlorite, on the other hand, are relatively nonexpanding.

Smectite - predominantly have negative charges resulting from isomorphous substitution. Layers are bound via an oxygen-oxygen surface and thus the bond is weak allowing for cations and water to enter the spaces in between.

Vermiculite -- tetrahedral sheets have large amounts of Al substitution for the silicon position, which produces a CEC higher than other silicate clays

Fine-grained micas (e.g., illite and glauconite) -- very weathered micas with substitution of Al^3+^ for Si^4+^ in the tetrahedral sheets. The negative charge attracts K^+^ ions which strengthens bond between the layers and making the fine-grained micas nonexpansive. 

Chlorites -- Fe or Mg primarily occupies the place of Al in the octahedral sheets. Mg dominated octahedral sheets commonly sandwiched between two 2:1 layers, thus often said to have a 2:1:1 structure.

*Non-silicate colloids*

Fe and Al oxides are motified octahedral sheets with Fe or Al in the cation positions, and no tetrahedral sheets or silicon in their structures. 

Humus is noncrystalline and is formed of large organic molecules of varied chemical composition. Negative or positive charges on humus colloid are a result of loss or gain of H^+^ ions; a very large net negative charge is associated with humus, though sites of both positive and negative charges may exist

*Formation of colloids*

Humus is primarily formed through the breakdown or organic matter (mostly plant matter) by microorganisms and the resynthesis of this material into stable organic compounds.

Silicate clays develop through:

  1. Physical and chemical alteration of primary minerals
  2. Decomposition of primary minerals with recrystallization of products into silicate clays
  
Kaolinites and Fe and Al oxides are found primarily in warm, humid regions (e.g., tropics), whereas smectite, vermiculite and fine-grained micas are primarily in regions where weathering is less intense.

*Sources of charges on soil colloids*

Two major sources of charge imbalances:

  1. Hydroxyls and other functional groups on surfaces of colloids releasing or accepting H^+^ ions
  2. Charge imbalances as a result of isomorphous substitutions
  
Charges of all colloids are affected to a certain degree by OH^-^ groups, which are largely pH dependent. The charges of Al and Fe oxides, 1:1 type clays, allophane and humus are primarily determined by OH^-^ groups.

For 2:1 clays, surface charges are complemented by isomorphouos substitutions of cations. Charges from isomorphous substitution are not dependent upon pH, and thus are considered "permanent"

*Cation exchange reactions:*


### Soil pH

### Soil organisms

### Soil organic matter

### Soil nutrients

### Redox



### Sulfur cycle

Sulfur is particularly important in wetland ecosystems due to the anaerobic conditions of the soil and thus reliance on sulfur for metabolic activities.





## Geomorphology {#geomorphology-soils}

### Thom, 1967

*"Mangrove ecology and deltaic geomorphology: Tabasco, Mexico"* [@thom1967mangrove]

A full summary of Thom, 1967 is provided in the "Zonation and Disturbance section." Please refer to Section \@ref(thom1967).

From a geomorphological standpoint, this is one of the most well-cited studies arguing that mangrove vegetation patterns respond primarily to the hydrological and sedimentary conditions that exist within a given site.

### Carlton, 1974

*"Land building and stabilization of mangroves"* [@carlton1974land]

**Key contribution:** This is a perhaps less seminal paper but provides a brief review of the literature surrounding the ability of mangroves to "build" and "stabilize" shorelines. Much of the reviewed work is from Florida.

**Key conclusions:** Carlton concludes more or less that mangroves do not "build" land *per se*, but are influential upon the *rate* at which sediments accrete on shorelines. In sum, two viewpoints exist:

  1. mangroves "build" land by accretion of sediments (*a la* Davis 1940) (UNSUPPORTED)
  2. mangroves modify the rate at which sediments accrete, thus affecting the speed of geomorphic change rather than the direction (Egler 1952, Gill 1971) (SUPPORTED)
  
*Further reading:*

  1. Egler 1952
  2. Gill 1971

### Thom, 1982 {#thom1982}

*"Mangrove ecology: A geomorphological perspective"* [@thom1982mangrove]

**Key contribution:** This study is best known for the five environmental settings that Thom lays out for clastic (terrigenous) sediment deposition. The five environmental settings are detailed below:

  1. *River dominated* - Deltaic or estuarine settings in which significant sediment is transported from alluvial flows to the coastal region. The landscape is marked by numerous distributaries and interdistributary basins that vary in hydrology, particularly in regards to tidal influence.
  2. *Tide dominated* - Estuarine type settings in which the dominant hydrology is the bidirectional flow of tides. Often see sandy shoals within the estuary that are shaped parallel to the flow of tides.
  3. *Wave dominated* - Environmental settings in which flat shorelines accumulate sand barriers or cheniers that create lagoons and create regions of low-energy coastlines in which mangroves may establish. Sediment is primarily autochthonous.
  4. *River & wave composite* - Settings in which both deltaic plains with strong alluvial influence as well as strong wave influences create a complex of alluvial fans and lagoons blocked by wave-distributed sand barriers from longshore flows. E.g., Tabasco, Mexico.
  5. *Drowned river valleys* - Estuaries or embayments that are underlain by impervious rock and have been flooded during the most recent period of sea level rise.
  
**Key knowledge:** Thom describes the geomorphology of mangrove from three standpoints:

  1. *Geophysical* - The dominant physical processes that affect a landscape, such as SLR or climate
  2. *Geomorphic* - The dynamic history of the land in response to the dominant geophysical factors. Can be broken into three scales:
    a. macro: clastic vs. carbonate sediments
    b. meso: dominant physical process (tide vs. riverine vs. wave)
    c. micro: microtopographic features such as drainage and elevation that produce environmental gradients
    
  3. *Biological* - Factors such as biogeography or competition that affect the landscape
  
Additionally, Thom notes that his five suggested geomorphig settings *are not comprehensive but are rather representative of dominant processes that may be relevant.* Thom goes on to discuss variations and scale in time and space and their resultant effects on mangrove geomorphology. He concludes that zonation of mangrove species responds opportunistically to geomorphic setting, and areas with high geomorphic diversity will have high degrees of habitat variation.

**Topics to think about further:**

  1. How do geomorphic settings related to soil texture and soil carbon?

### Woodroffe, 1983

*"Development of mangrove forests from a geological perspective"* [@woodroffe1983development]

**Key contribution:** This paper describes the development of an autochthnous, geomorphically inactive mangrove forests on Grand Cayman island. The study takes a geological perspective in that they examine soil depths and soil cores throughout the west peninsula, and reconstruct models of autochthonous development of the mangrove forest through time.

**Key findings:** The study concludes that there were key tidal inlets which were eventually breached my rising sea levels and allowed for expanded colonization of the calcerous bedrock by mangroves. Soil depths & cores indicate that there was a forest of landlocked mangroves at one point during low-levels (200 cm below present), but the authors are unsure of how tidal fluxes reached this area.

```{r, echo=FALSE}

knitr::include_graphics(rep("images/woodroffe1983TVS.bmp"))

```


### Thom, 1984 {#thom1984}

*"Coastal landforms and geomorphic processes"* [@thom1984coastal]

**Key contribution:** This is an extension of [@thom1982mangrove], in which Thom describes three carbonate settings. The conclusions are largely the same as [@thom1982mangrove], and thus refer to section \@ref(thom1982) for them.

**Key additions:** The carbonate settings that Thom adds to his original five clastic settings are as follows:

  6. *Carbonate platform* - Accumulation of lime muds and peat on low-energy tropical coasts. Oftentimes the fringing mangroves are against large, open, shallow bodies of water. Coral reef or sand barriers may block wave and tidal energy to produce low-energy conditions
  7. *Sand/shingle barrier* - Mangroves may form on the cemented or non-cemented coral detritus, for example patchy mangroves colozining portions of the Great Barrier Reef. Sand mostly consists of carbonate particles and sea levels play an important part on depositional characteristics
  8. *Carbonate embayments* - Embayments in which soil "crusts" were submerged by SLR during Holocene. E.g., patchy mangroves on old reef complexes that are highly variable and reflect varability in the substrate complexity.
  
**Key knowledge:** In addition, Thom discusses some of the prior models of succession and zonation in mangroves. In particular he identifies three trends of research, including one in which he situates himself:

  1. *"Classical succesional view of mangroves"* - The "Davis 1940" view of mangroves that emphasizes the role of biota rather than geophysical or geomorphological processes
  2. *"Environmental gradient view"* - The more "ecological niche" based view of vegetation patterning that results from environmental gradients; Watson 1928 was one of first studies to argue for this effect
  3. *"Physiographic ecology"* - regards physical and botanical landscapes as mosaics of landforms that result from differential effects of physical processes on different materials, and the promotion of different biological communities in response to them
    - focused at landscape scale, geomorphic variety produces variety in biological assemblages

**Other reading:**

  - Snedaker, 1982 - Mangrove species zonation: Why?

### Woodroffe, 1990

*"The impact of sea-level rise on mangrove shorelines"* [@woodroffe1990impact]

**Key contribution:** This review by Woodroffe discusses the impacts of relative sea level rise as it was projected at the time on mangroves. There has since been 27 years worth of research in this topic, and thus it's important to couple this study with more recent papers.

*Crux of the issue* - "order of rates of sea-level rise that have been predicted would in some cases lead to major disruption of mangrove forests"

**Key notes: **

  - ability of mangrove forests to respond to different levels of sea level rise depends on coastal morphology
    1. steep cliffs (e.g., mid-oceanic atolls) may not be able to respond due to lack of ability to migrate along shoreline as well as distant propagule sources for recolonization
  - Eustatic patterns during Holocene have varied regionally
    - in Indian and Pacific Oceans, rose to approximately 1-2 m above present levels, before falling and stabilizing at current levels
    - in West Indies and Southern US, increasingly risen though at decelerating rates to arrive at period of stability of today
  - mangroves are effective indicators of past sea-level if species with given preference for position in intertidal zone can be identified within paleosedimenets
    - however, often not clear because of compaction and contamination with younger aged organic matter
  - given sea level rise, the ability of mangrove forests to persist may be largely dependent upon the catchment size of deltas (i.e., size of area from which water and sediments drain)
  - carbonate settings are at higher risk of undergoing regime shifts relative to deltaic or estuarine mangroves
    
Woodroffe combines Thom's environmental settings into three primary classes of geomorphological development:
  
  1. river dominated
  2. tide dominated
  3. carbonate
  
He expounds on them in this article, whereas he summarizes them in his review in Tropical Mangrove Ecosystems (see Section @\ref(woodroffe1992)).

### Woodroffe, 1992 {#woodroffe1992}

*"Mangrove sediments and geomorphology"* [@woodroffe1992mangrove]

**Key contribution:** This is a key synthesis of knowledge related to geomorphology and environmental settings in mangroves. Woodroffe gives an overview of sedimentation and shoreline progradation / erosion, followed by a review of Thom's environmental settings and Lugo and Snedaker's mangrove habitats. He concludes the article with a review of stratigraphy and paleoecology, with particular reference to mangrove responses to relative sea level rise.

**Key findings:**

  - For Thom's clastic environmental settings, refer to Section \@ref(thom1982) 
  - For Thom's carbonate environmental settings, refer to Section \@ref(thom1984)
  - For Lugo and Snedaker's functional habitats, refer to Section \@ref(lugosnedaker1974)

Diagrams of Thom's environmental settings are provided below:

```{r, echo=FALSE}

knitr::include_graphics(rep("images/thom1982.bmp"))
knitr::include_graphics(rep("images/thom1984.bmp"))

```

*Stratigraphy and paleoecology:*

  - Quaternary period (~past two million years) comprised of:
    1. Holocene (~11,700 YBP)
      - sea level has been approximately stable for 6,000 years
      - sea level was rising rapidly from approximately 10,000 - 6,000 YBP
      - though there is likely regional variation
    2. Pleistocene (~2.5 mil to 11,700 YBP)
  - Sediments deposited during Holocene are well preserved and contain structures & fossils that are useful, whereas sediments deposited during Pleistocene are less well preserved
  - sea level rise on more gradually sloping coasts likely resulted in significant fluctuation of mangrove forests, whereas carbonate atolls (e.g., on coral) likely disappeared during high sea levels or were restablished during low sea levels
  - *transgression* - landward movement of shoreline during periods of sea level rise
  - *regression* - seaward movement of shoreline during periods of sea level subsidence
    - there are also iteractions with progradation and erosion that may reverse processes at specific locations
  - Woodroffe examines three patterns of tropical geomorphological development:
    1. tropical deltas that are river dominated (poorly studied as of 1992)
    2. estuarine embayments that are tide-dominated
      1. Progradational model - estuary shoreline gradually build into estuary and is generally fringed by mangroves
      2. Big swamp model - rapid accumulation of sediment as sea level stabilizes followed by vertical accretion and conversion to other types; differs from progradational model in that there's little age difference between mangrove sediments
      3. Barrier estuary/mud basin - sand reworked from seward shelf and forms sandy barrier or flood tidal delta; lagoon is sink for fluvial sand and mud from terrestrial sources by rivers
    3. carbonate banks or reefs
      - predominantly characterized by in situ production of sediments

*Note: * This is a key study that would be good to review again later.


### McKee, 2007

*"Caribbean mangroves adjust to rising sea level through biotic controls on change in soil elevation"* [@mckee2007caribbean]

**Key contribution: ** This is a very well-cited study that experimentally examples biological controls on soil elevation in mangroves of the Meso American Barrier Reef (in Belize). 

**Study design: ** The authors installed elevation rods with rod surface elevation tables (RSETs) that were able to measure changes in elevation from a baseline over the course of three years in fringe, transition and interior mangroves. They examined the effects of nutrient additions (N, P & a control) across three transects at three replicate sites. In addition, they laid a sand horizon in a plot adjacent to the RSETs and used cores to monitor subsurface accretion, allowing for differentiation between subsurface accretion and surface accretion of biomass to elevation. Finally, the authors examine several cores (up to 10m in depth) across sites throughout the Caribbean to examine relative SLR and mean soil accretion.

**Key results: ** The authors conclude that the fringe portion of the forests are able to keep pace with sea level rise, primarily due to subsurface accumulation of biomass (significantly correlated with elevation change). The fringe zone was found to have significantly positive gain in elevation, with majority from subsurface change, whereas interior and transition zones saw subsidence. Elevation gains were seen across all sites for P addition, whereas elevation gain was variable in directino for N (negative in interior).

The authors give three key points:

  1. Rates of modern change in elevation may vary in both direction and magnitude within forests
  2. Elevation changes vary in relation to physico-chemical gradients that may control natural productivity of mangroves
  3. Elevation changes due to variation in nutrients are variable in both direction and magnitude
  
The key results in elevation change across treatments are shown below:
  
```{r, echo=FALSE}

knitr::include_graphics(rep("images/mckee2007GlobEcolBiog.bmp"))

```

The results from the mangrove peat cores showed that peat accumulation corresponded with rates of SLR over the past 6,000-7,000 years. Although they are currently keeping pace with sea level rise, the cores show that at points (in which SLR exceeded 5 mm / yr), mangroves were lost from the area. Thus, the presence of mangroves at these sites depends on relative SLR rates (eustatic rise plus local subsidence). Rates of mangrove peat accretion, however, may underestimate ability to respond to SLR as these are often based on mean rates from historical records that likely fluctuate significantly in response to SLR.

### Woodroffe, 2016

*"Mangrove sedimentation and response to relative sea level rise"* [@woodroffe2016mangrove]

**Key contribution: ** This study is a review of sedimentary processes and geomorphology with a particular focus on their responses to relative sea level rise. The study introduces the concept of the accomodation space in the context of mangroves, which is the space available for potential sediment accumulation.

**Notes: **

Considering processes at different scales:

  1. Macroscale - processes at the landscape level (environmental setting *a la* Thom)
    - Climate, eustatics
  2. Mesoscale - habitat level (fringe, basin, etc. *a la* Lugo and Snedaker)
    - Hydrodynamics, sediment supply (allochthonous vs. autochthonous)
  3. Microscale - individual tree or plot level
    - Interactions between surface elevation, geomorphic processes (accretion vs. compaction), and biological processes (BGB production)

```{r, echo=FALSE}

knitr::include_graphics(rep("images/woodroffe2016ARMS.bmp"))

```

A nice summary of interactions between sedimentology, hydrology and biology:

"The distribution of mangrove species coincides with the intertidal accomodation space, which progressively fills through sediment accumulation. Emergent properties and biocomplexity are a function of hydroperiod and are related to salinity and flooding tolerance, the latter of which is related to root morphologies, physiology, and reproductive traits such as vivipary."

*Accomodation space* - the space available for potential sediment accumulation. Typically see highest accomodation space at shoreline in mangroves, though it may be highly variable (e.g., basin areas that are only flooded during highest tides).

  - may distinguish between mineral vs. biological accretion, and thus may consider "ecological accomodation space"
  
Rates of sedimentation depend upon the accomodation space, which depends on hydrology and other factors. In the short term, the rate of sediment deposition appears independent of rate of SLR. The accomodation space is directly constrained by upper limit of tidal flooding.

Focusing on vertical adjustment within a mangrove forests risks missing broader regional picture in that mangroves have constantly shifted position throughout time, and thus there may be natural fluctuations with sea level rise that can maintain existence as well as are potentially key climate change mitigation opportunities.

**Further reading:**

  - See the highlighted references in Mendley for additional reading -- there are quite a number worth looking at further.


## Sedimentation and hydrology

### Preface

### Wolanski, 1980 {#wolanski1980}

*"Hydrodynamics of a tidal creek-mangrove swamp system"* [@wolanski1980hydrodynamics]



### Wolanski, 1986 {#wolanski1986}

"*Tidal mixing and trapping in mangrove swamps"* [@wolanski1986tidal]

**Key contribution:** This study primarily seeks to answer two key questions:

  1. What parameters control export of salt left after evapotranspiration of tidal waters by mangrove trees?
  2. How do rivers influence the growth of mangroves?
  
There is a lot of mathematical hydrological modeling that is over my head in this study, but there are some relevant ecological points that are worth noting.

**Key results:** Some key results from the study include:

  - The "trapping mechanism" is the dominant parameter controlling "longitudinal dispersion" in their study swamp. They conclude that it is likely the dominant mechanism in many swamps in the absence of large buoyancy effects (i.e., dry seasons)
  - Buoyancy effects are important due to trapping of freshwater in the swamp at high tide and boundary layer that enters the river banks at low tide
  - Residence time of runoff in mangrove swamps may be quite long with important ecological implications, such as nutrient uptake and settling of fine sediments
  
A diagram of the "trapping mechanism" of fresh water in the landward portion of swamps is shown below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/wolanski1986.bmp"))
```
  
**Further reading:** This study led to some questions that may be important for understanding soil formation in mangroves. In particular:

  1. What is the "trapping mechanism" in estuarine hydrodynamics?
  2. What are the specific processes of tidal mixing and what is the broader relevance for the ecology of mangroves?
  3. What is longitudinal dispersion in relation to mangroves?
  4. What are the specific implications of buoyancy effects, and how do they vary between dry and wet seasons?

### Wolanski, 1992

*"Mangrove hydrodynamics"* (In Robertson and Alongi, 1992)

### Furukawa, 1996 {#furukawa1996}

*"Sedimentation in mangrove forests"* [@furukawa1996sedimentation]

**Key significance:** Furukawa and Wolanski show through an empirical study that a sediment pump from coastal waters to the mangrove interior does exist.

They find that the presence of a tidal pump is largely due to:

  1. High turbulence that maintains entrainment of fine sediments during flood tide.
  2. Settlement of flocculated sediment during slack high tide in which turbulence is reduced.
  3. Insufficient ebb tide velocities to resuspend settled sediment.

**Key notes:**

Water velocity of tidal flooding in a mangrove forest can be described as:

$$ u = 1/n\ *\ h^{2/3}\ *\ I^{1/2}  $$

Where:

  - u: water velocity
  - n: Mannings friction coefficient
  - h: water depth
  - I: water surface slope

Typical values for the Mannings friction coefficient for sandy channels are in the range of 0.025-0.035, whereas empirical studies by Wolanski have found values as high as 0.2 - 0.4 for areas with mangrove vegetation. Thus, mangrove vegetation increases the Mannings friction coefficient by several orders of magnitude.

This results in a failure to resuspend fine sediments that are deposited during slack high tide and thus promotes build-up of sediment.

Consideration of floc size is important for sedimentation studies within mangroves. In calculating the settling velocity (via Stoke's law) for both individual particles of clay and silt, as well as that of mean floc size measured within the tidal water, settlement of flocs was found to be approximately 100 times larger than for individual particles (0.005 m/s).

Empirical measurements in the field showed that for tree species with complex matrices of aboveground roots, turbulence was intense enough to maintain flocs in suspension. For single stemmed trees without root systems (e.g., Bruguiera or Ceriops), both turbulence and flow velocity are reduced.


### Furukawa, 1997 {#furukawa1997}

*"Currents and sediment transport in mangrove forests"* [@furukawa1997currents]

**Key contribution:** A sister study to [@furukawa1996sedimentation], this study describes the empirical data collected to investigate current flows and suspended sediment characteristics of a mangrove in Australia.

**Research design:** Water level, currents, sedimentation (via sediment traps and Analite nephelometers (upward looking cameras)), suspended sediment levels (via water samples), and concentration of flocs (via microscopic slides with wells) were measured along a transect in the forest, at varying distances from a tidal creek.

The data were compared against three hydrological models to investigate the accuracy of mathematical modelling of hydrodynamics and sedimentation in mangroves, as well as analyze the empirical data for the site. The models were:

  1. 2DSWAMP - flow model based on primitive equations of motion
  2. LAKE - a lake model that assumes flat water surface and determines flow rate
  3. VORTEX - a discrete vortex model that provides information on turbulence

**Key findings:**

The LAKE model performed poorly and was not used within in the subsequent analyses (unrealistic assumption that the surface water is flat).

Suspended sediment concentration showed strong asymmetry in that concentration was higher at flood tibe and much lower at ebb tide. This is indicative of net inwelling of sediment into the mangrove.

Sedimentation rates decreased with distance from the tidal creek. In calculating the sedimentation velocity, the median floc size (90 micrometers) was approximately 4 times that of size of solid minerals (24 micrometers) given high porosity of flocs. Visual investigation of suspended sediment reveals that much of the flocs had settled out of the water column following slack high tide.

Vegetation greatly increased turbulence and created a complex set of hydrological effects (eddies, stagnant regions, and jets).

DOC was found to vary inversely with tidal depth, with the lowest DOC concentration present during slack high tide. Furthermore DOC in the forest was found to be greater than coastal waters, indicating that both creek and mangrove waters are significant sources of DOC. 

The findings suggest a net inwelling of DOC in the forest (greater DOC concentration during flood tide than ebb tide), but this does not corroborate with other analyses.

### van Maren, 2013

*"The role of flow asymmetry and mud properties on tidal flat sedimentation"*



## Soil carbon processes

### Amundson, 2001

*"The carbon budget in soils"*

### Bouillon, 2003

*"Sources of organic carbon in mangrove sediments: Variability and possible ecological implications"*

### Krauss, 2003

*"Differential rates of vertical accretion and elevation change among aerial root types in Micronesian mangrove forests"*

### Lovelock, 2008

*"Soil respiration and belowground carbon allocation in mangrove forests"*

### Breithaupt, 2012

*"Organic carbon burial rates in mangrove sediments: Strengthening the global budget"*

### Lunstrum, 2014

*"Soil carbon stocks and accumulation in young mangrove forests"*



## SOC spatial patterns

### Preface

### Twilley, 1992

*"Carbon sinks in mangroves and their implications to carbon budget of tropical coastal ecosystems"*

### Lacerda, 1995

*"Biogeochemistry of mangrove soil organic matter: A comparison between Rhizophora and Avicennia soils in southeastern Brazil"*

### Atwood, 2017

*"Global patterns in mangrove soil carbon stocks and losses"*
