# Remote sensing

## Mangrove RS reviews

### Adam, 2010

*"Multispectral and hyperspectral remote sensing for identification and mapping of wetland vegetation: A review"* [@adam2010multispectral]

**Key contribution: ** This review examines the status quo of mapping wetland vegetation using spectral (multi and hyper) imagery, with a particular focus on discriminating wetland vegetation as well as monitoring biophysical and biochemical characteristics.

**Key findings: ** 

The spectral and spatial composition of mangrove species and vegetation assemblages complicates use of multispectral (broadband) imagery

  - Hyperspectral (narrowband) imagery is promising but more difficult to access

  - Spectral signatures of different wetland vegetation types are often similar, and are "muddied" by spectral reflectance from soil, hydrologic regime or atmospheric conditions

Spectral signatures affected most strongly by biochemical and biophysical properties of leaves, which vary greatly across species

Discrimination of wetland species from Landsat TM or SPOT imagery has generally been insufficient due to:
  1. Lack of fine, ecological divisions between vegetation species
  2. Broad nature of spectral wavebands
  3. Lack of high spectral and spatial resolution to detect subcanopy vegetation types

Aritifical Neural Networks and fuzzy classification techniques have shown promise in better classifying mixed species pixels

Most important biochemical and biophysical properties characterizing wetland species are:
  1. Chlorophyll - green pigment
  2. Biomass concentration - biophysical prop
  3. Leaf water content

Major issue of using vegetation indices (e.g., NDVI) to estimate biomass is the saturation problem
  - Mutanga and Skidmore (2004) propose one mechanism to resolve saturation problem
  
Building spectral libraries via hyperspectral sensors is critical for discriminating wetland species, but also needs investigation at different sites & scales as reflectance within species may vary depending on environmental conditions

  - Vaiphasa and Li Wang have done much of the work in building hyperspectral libraries for mangroves (see \@ref(vaiphasa2005) & \@ref(wang2009))
  
**Research ideas:**

  1. Spectral imagery can be effective in monitoring plant water content, which may vary significantly across mangrove species due to differing water use efficiencies
  

### Heumann, 2011

*"Satellite remote sensing of mangrove forests: Recent advances and future opportunities"* [@heumann2011satellite]

**Key contribution: ** One of several reviews of mangroves and remote sensing techniques that focuses primarily (i.e., solely) on vegetation mapping. 

**Key findings: **
  
Spectral discrimination of different species not successful using multispectral (broadband) imagery (Ramsey and Jensen, 1996).
  
  - Mixed pixels also an issue for discriminating species as well as estimating canopy structure

Spectral saturation at high biomass values is a particular problem for mapping biomass and volumes

Very high resolution imagery effective at reducing mixed pixels just due to smaller size

OBIA has typically been applied to species discrimination rather than extent of mangrove mappings

Quickbird and IKONOS sensors are most common for very high resolution imagery due to long-term archiving of data

  - Multispectral sensors but show improvement over Landsat TM or ETM+ due to reduced numbers of mixed pixels

Use of spectral data only is often insufficient for discrimination of species

  - Classification accuracies are greatly improved when other image properties are considered, such as texture
  - Texture analaysis has been applied to species discrimination, but has not been investigated (as of 2011) for other forest characteristics

C-band polarimetric SAR has lowest ability to parse forest structure of SAR bands
  
*Mangrove biomass specific findings:*

  - Canopy height and biomass have been shown to be strongly correlated in mangroves (e.g., Fromard et al., 1998; Smith and Whelan, 2006)
    - Most studies of height and biomass have employed SAR data
    - Different polarizations (based on wavelength) interact differently with different components of the forest
      - Trunks and soils may produce single or double bounce interactions (typically woody components best detected by L-band)
      - Signal may scatter within canopy
        - P-band polarimetric SAR or HV polarization of L-band SAR perform best in mapping aboveground biomass

*Very-high resolution satellites:*

  - IKONOS and Quickbird are most commonly used, due to archiving of data and plans for long-term consistency
    1. IKONOS > GeoEye-1 (successor) (multispect @ 1.65 m and panchrom @ 0.41 m)
    2. Quickbird > Worldview-2 (successor) (<2 m multi, <0.5 m panchrom)
      - Note that Worldview-2 has 8 multispectral channels, including those in yellow and red edge ranges designed for vegetation studies

**Research ideas:**

  - Coupling of species maps with species-specific allometric equations for better estimates of mangrove biomass
    - Might make sense in Americas where just three species exist (easier!)

### Kuenzer, 2011

*"Remote sensing of mangrove ecosystems: A review"* [@kuenzer2011remote]

**Key contribution: ** This is a very extensive review of remote sensing and mangroves. It may not provide much more detail beyond that of the other two reviews of remote sensing and mangroves.

**Notes: **

*Optical imagery: *

The spectral signature of mangrove leaves are largely a function of age, vitality, phenological properties and physiological properties. Greater discrimination between species (at least in Vietnam) may be seen in the NIR, but little differentiation seen across the visible light wavelengths:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/kuenzer2011RemSens.bmp"))
```

*Radar imagery: *

  - SAR imagery provides intensity values of the "backscatter coefficient". Given different polarizations and wavelengths of SAR scanning, different interactions with mangrove landscape occur.
  - The different SAR bands are as follows:
    1. X-band - 3 cm wavelength, effective for direct backscatter from canopy
    2. C-band - 6 cm, plant/soil multiple and volume scattering
    3. L-band - 24 cm, direct backscatter from soil & water
    4. P-band - 68 cm, longest wavelength
    
```{r, echo=FALSE}
knitr::include_graphics(rep("images/kuenzerRemSensFig4.bmp"))
```
  
*Sensors:*

See Table 2 in the study for an extensive list of sensors used within remote sensing of mangrove studies

Aerial imagery is limited in that it cannot be scaled beyond immediate site/locality, but has value in that it can provide high-resolution data particularly for hard to access sites without field data.

High spatial resolution data (IKONOS & Quickbird):

  - Ultimate goal is to map mangroves at the species level
  - A comparison of IKONOS and QuickBird found IKONOS (4 m) to be more effective at discriminating in species because of the higher spectral resolution. QuickBird has higher spatial resolution (1 m) but was less useful in this particular investigation (Greg's paper).

Hyperspectral imagery
  - Identifying relevant bands is a time intensive process and many bands are highly correlated

EnMAP sensor
  - Satellite hyperspectral data that will provide more access to hyperspectral datasets globally
    - Is this online? I'm not sure yet.
  
Several tables are given in the paper that describe the various data sensors and platforms and their relative benefits and limitations for remote sensing of mangroves.

*Needs within RS of mangroves space:*

The authors call for several needs within the remote sensing of mangrove space:

  1. Need for homogenized classification schemes
    - "Ontologies", e.g., what is a sparse versus dense mangrove
  2. Need for standardized data-processing models 
    - Standardiziing preprocessing, processing and post processing such that enhanced comparability of studies is enabled
  3. Need for homogenized transparent accuracy assessment
  4. Need for further investigations of synergistic data use 
    - Coupling of new, novel and disparate RS datasets to investigate novel research questions
  5. Need for ecosystem service assessments
  6. Need for interdisciplinary and well-networked research teams

### Giri, 2016

*"Observation and monitoring of mangrove forests using remote sensing: Opportunities and challenges"* [@giri2016observation]

**Key contribution:** This is an introductory piece to a special issue on monitoring mangroves using RS. There are potentially interesting pieces in the special issue but most are focused on vegetation dynamics rather than geomorphology.

Giri is one of the key researchers working on mapping of mangrove forests using RS data, and offers some succinct advice on steps for classifying mangrove forests:

  1. Identify clear and concise objective
  2. Develop classification system and legend
  3. Satellite and secondary data acquisition
  4. Data pre-processing
  5. Training data collection
  6. Image classification
  6. Results validation
  8. Dissemination of results
  9. Application of classification results
 
*Random notes: * 

  - Important to be clear on consideration of purely true mangroves, or true mangroves with mangrove associates.
  - Entire Landsat archive available in Amazon Web Services as well as Google Earth Engine.

Article synopses:

  - Zhu et al, 2016: 
    - Looked at back-propagation artificial neural network to map mangrove forest biomass using WorldView-2 imagery. 
    - Find that inclusion of species dummy variable for *S. apetala* vs *K. candel* highly significant in improving biomass estimation accuracy
  - Cuogo et al, 2016: 
    - Examined SAR backscattering from Radarsat-2 C-band data for correlations with forest structure and species data in Amazon. Might be interesting to look into their methods for sites in Thailand.
  - Jones et al, 2016: 
    - Used Giri's data to assess changes in mangrove cover at national level for Madagascar. Produced higher accuracy maps with improved detail relative to Giri's data. Potentially interesting for Thailand.

## Vegetation

### Ramsey, 1996 {#jensen1996}

*"Remote sensing of mangrove wetlands: Relating canopy spectra to site-specific data"* [@ramsey1996remote]

**Key significance: ** This is one of the earliest comprehensive attempts to relate multispectral wavelength data to mangrove forest structure characteristics. In particular, the study seeks out to answer two primary questions:

  1. What are the spectral and structural changes within and between mangrove species and community types?
  2. What are the relationships between the canopy spectral variability at the blue, green, red and NIR wavelength regions and canopy species composition and structural variability?
  
Their results largely showed that remote sensing of species composition is quite difficult (no relationships between species composition and spectral indices exist), but LAI and a few other parameters (canopy height) were found to be significantly correlated with some spectral bands/vegetation indices.

**Key notes: **
  
*Research design:*

Field structure and species composition data was collected and coupled with remotely sensed spectral data. Regression and other statistical techniques were used to relate the data to the various spectral measures and their arithmetic transforms. Spectral measurements were taken in the lab, from a helicopter platform, and also from SPOT satellite.

*Key results: *

The following results and key conclusion of the study should be noted:

  - LAI was calculated as a transform of canopy closure, which was collected with a hemispherical mirror
  - Significant correlations of any of the bands with percent species composition was not significant
    - The strongest correlations between spectral signatures and forest structure data was with LAI
  - LAI was not significantly correlated with species composition, but was significantly correlated with forest height, which indicates that canopy depth expands with increasing canopy height.


### Green, 1998 {#green1998}

*"Remote sensing techniques for mangrove mapping"* [@green1998remote]

**Key significance: ** The study reviews common methods of remote sensing for mangrove mapping, with a particular emphasis on the fact that accuracy assessments are rarely performed following classification of RS mangrove imagery. 

The authors couple the most commonly used techniques with accuracy assessments using field-based data to examine the relative strengths & weaknesses of the different techniques.

**Research design:** The study employs three spectral image datasets (Landsat 6 TM, SPOT, and Airborne multispectral data at 1 m resolution) for generating RS imagery of mangroves in the Turks and Caicos Islands of the Bahamas. The review of classifying spectral imagery for mangroves identifies 5 different methodologies:

  1. *Visual interpretation* - the user classifies areas, though often with a radiometric enhancement applied to the imagery
  2. *Vegetation index* - different vegetation indices can be employed such as ratio indices (NDVI) or orthogonal indices (tasseled cap transformations)
    - This study uses NDVI with a ISODATA clustering algorithm producing 50 classes that are identified and edited to match mangroves
  3. *Unsupervised classification* - may or may not involve image enhancement 
    - This study used ISODATA clustering algo on unenhanced imagery
  4. *Supervised classification* - most common method of mangrove classification and may or may not include image enhancement (e.g., enhancing edges between stands or species types) 
    - This study uses 3 x 3 edge enhancement filter applied to imagery, useful in mangroves given linear nature
  5. *PCA and band ratios* - ratios of different bands can be effective in distinguishing different landscape features 
    - This study uses the ratios of various Landsat and SPOT bands, runs PCA, and then generates an image based on the most significant PCAs (which they then classify)
  
Accuracy assessment was performed using three metrics (following Congalton, 1991): 

  1. Overall accuracy - overall agreement as sum of correctly labelled sites divided by total number of sites
  2. User's accuracy - probability that a classified pixel actually represents that category on the ground
  3. Producer's accuracy - how often real features on ground are shown correctly on the map (NOT used in this study)
  4. Tau coefficient - indicates the percent of pixels classified correctly over what would be expected by chance alone

**Key results:**

The authors largely justify the work based on:

  1. A lack of widespread accuracy assessments within the literature
  2. A discussion of the limitations of climate on spectral RS imagery classification (although their methods only examine spectral datasets; they give SAR a head nod in the review section)

*Discrimination of mangrove vs. non-mangrove vegetation:*

SPOT XS imagery was not effective at producing accurate classifications of mangrove extent. Compact Airborne Spectral Imager  (CASI; airborne) data was highly effective at identifying mangrove vegetation (with all five methods producing overall accuracies greater than 50%), whereas methods performed on Landsat data improved markedly with increased computational sophistication (i.e., method 5).

Accurate classification of Landsat data required the calculation of PCs and band ratios, rather than processing of just raw band data.

In general, visual identification and ratio index based classifications are *not* effective at classifying mangroves.

Additionally, merging of Landsat TM and SPOT XP (i.e., down-sampling Landsat data) was found to be effective at improving mapping classifications (using both a principle components merge, as well as an intensity-hue-saturation merge).

*Discrimination of mangrove habitats:*

The results show that the airborne spectral data (CASI) was most effective at accurately classifying 9 different mangrove classes. Reasonable classification accuracies could be achieved with the Landsat data for a reworking of just three mangrove classes (tall Rhizophora, other mangrove, and non-mangrove), but were highly inaccurate for a 9-class map.

**Note: ** There is a nice table of other works on mangrove mapping and remote sensing in the introduction. It may be good to review this table again at a later date.

### Proisy, 2000{#proisy2000}

*"Interpretation of polarimetric radar signatures of mangrove forests"* [@proisy2000interpretation]

**Key contribution:** This study reviews the interaction of backscattering signals of different polarizations of SAR data with components of a mangrove forest.

**Research design:** The study took field-based data on forest structure and species composition, as well as collected data from the P-, L-, and C-Bands of AIRSAR data (NASA JPL; airborne SAR). They calculated a number of backscattering coefficients for each of the polarizations (HH, VV and HV), as well as the copolarization ratio (VV/HH) as well as two cross-polarization ratios (HH/HV and VV/HV).

The different bands were acquired at a mean incidence angle of 35 degrees at:

  - P-band: 0.44 GHz
  - L-band: 1.25 GHz
  - C-band: 5.3 GHz

**Key findings:**

Within the L-band, surface scattering or double-bounce effects dominate up until approximately 150-200 Mg/ha, whereas volume scattering (i.e., within the canopy) dominates for biomass values greater than 150-200 Mg/ha.
  
The dominant contributer to the backscattering coefficient depends on the type of forest being monitored. For example:

  - Mature or declining stands with biomass greater than 300 Mg/ha have volume scatter at all frequencies.
  - Mature or declining stands with biomass between 100 - 300 Mg/ha have dominant volume scattering at C-band, volume and double-bounce scattering at L-band, and double-bounce scattering at P-band
  - Pioneer stages with less than 100 Mg/ha biomass have volume scattering present in all three polarizations of C-band, double-bounce or surface scattering at L-band, and all three present (though interaction term is strongest) at the L-band for all three polarizations.
  
Overall, P-band provides most pronounced polarimetric signatures, and is the most useful for distinguishing between different scattering mechanisms and thus different forest stages.

A nice summary figure of the interactions between forest structure and radar returns is below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/proisy2000.bmp"))
```

**Notes: ** Prior studies have found that there is a strong positive relation between SAR backscattering coefficients and total biomass, particularly for P-HV and L-HV.

The main contributions to the total backscattering coefficient response are: 
  
  1. the soil scattering component
  2. the volume scattering component
  3. the double-bounce scattering component


### Proisy, 2002 {#proisy2002}

*"On the influence of canopy structure on radar backscattering in mangrove forests"* [@proisy2002influence]

**Key significance: ** One of a series of studies by Proisy *et al*, this particular study examines two mangrove forest stands of near identical aboveground biomass, but markedly different forest structure. They examine the polarimetric SAR data that they collect from the two plots, and examine them in relation to the two different forest structures.

The forest structures both contain ~75 Mg biomass/ha, but are structured as:

  1. A pioneer stand of *L. racemosa* that is extremely dense (~12,000 trees/ha) and has much higher basal area (21.0 m^2^/ha)
  2. A declining stand with two canopy strata:
    - *A. germinans* that are dying off and form an open, tall canopy (250 trees/ha, 4.2 m^2^/ha & 19 m height)
    - *R. mangle* that are replacing the *A. germinans* (31 trees/ha, 0.2 m^2^/ha & 8 m height)

**Research design:** The same research methods as [@proisy2000interpretation] (see Section \@ref(proisy2000)) were employed, but this particular study focused on an in-depth comparison of two plots with similar aboveground biomass values, but markedly different forest structure characteristics.

**Key results:** 

Overall, the backscattering from the open stand was higher than that of the closed (pioneer) stand.

With the help of a polarimetric scattering model, they found that the three primary contributers to backscattering (volume vs. interaction vs. surface scattering) differed between the two stands for select polarizations.

Additionally, they note two key concluding remarks:

  1. Correlating backscatter to biomass relationships should be restricted to homogeneous closed canopies.
  2. Distinguishing between stands of different developmental status should be possible via differences in backscattering coefficients of the different SAR bands and polarizations

A key figure that shows some of these differences is shown below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/proisy2002.bmp"))
```

### Mutanga, 2004 {#mutanga2004}

*"Narrow band vegetation indices solve the saturation problem in biomass estimation"* [@mutanga2004narrow]

**Key significance:** This study addresses the issue of vegetation index saturation at high biomass levels, and examines the effectiveness of hyperspectral data for resolving the issue. 

The study is largely motivated given that the "major limitation of using vegetation indices is asymptotic approach of a saturation level after a certain biomass density or LAI"

*Red edge* - The rise of reflectance at the boundary between chlorophyll absorption in red wavelengths and leaf scattering in NIR wavelengths (680-750 nm).

*Red edge position* can be defined as the point of maximum slope on the red infrared curve, and can be calculated by identifying the maximum of the first derivative

**Research design:** The study primarily focuses on two approaches to resolving the saturation issue with biomass and vegetation indices:

  1. Use of narrow band vegetation indices, which are simply NDVI, Simple Ratio (SR) and Transformed Vegetation Index (TVI), but for 637 bands of hyperspectral data (350-2500 nm)
  2. Use of the red edge position
  
A grass was grown in a greenhouse with two different nitrogen treatments (and one control group), and hyperspectral signatures were repeatedly taken. The various indices are computed for narrow band vegetation indices as well as indices using the red edge position.

**Key results:** 

*NDVI* - The results clearly indicate an improvement in a relationship between biomass and an NDVI index derived from red edge spectral bands rather than standard (NIR and red) bands.

*SR* - The results from the SR tests are similar to that of NDVI; reconstruction of SR using all possible hyperspectral band combinations indicates that bands from the red edge portion are most sensitive to changes in biomass

*TVI* - Although distribution of the R^2^ values for reconstruction of this index differed visually from that of NDVI and SR (which were quite similar), the most efficient TVI was found to be constructed from bands in the red edge portion of the spectrum. 

The green portion of the spectrum also did well (e.g., second "best") in reconstructing the vegetation indices.

*Red edge position* - The red edge position was found to have a high correlation coefficient with biomass, indicating there is a significant correlation

Of the three vegetation indices, SR was found to have the strongest correlation with biomass. Thus, narrow band derivations of SR from the red edge portion of the wavelength spectrum may be most efficient in estimating biomass from spectral data.

### Wang, 2004a {#wang2004a}

*"Comparison of IKONOS and QuickBird images for mapping mangrove species on the Caribbean coast of Panama"* [@wang2004comparison]

**Key contribution:** This study employs high resolution IKONOS (Space Imaging) and QuickBird (DigitalGlobe) imagery for improved mapping of mangrove species at Punta Galeta, Panama. In particular the study seeks to compare the imagery sources across three primary criteria:

  1. Spectral quality using subjective visual inspection and overall spectral statistics
  2. Classification effectiveness using multispectral bands with and without panchromatic bands
  3. Classification effectiveness following inclusion of textural information
  
**Research design:** IKONOS (4 m multi, 1 m pan) and QuickBird (2.8 m multispec, 0.7 m panchrom) imagery were bought for the region and geometrically corrected. Training and validation plots for the major land cover classes (red, white, and black mangroves, roads, lagoons, and rainforest) were established and data were collected.

MLC was used for classification of imagery using four multispectral bands, as well as a resampled panchromatic band at the resolution of the multispectral band resolution.

Additionally, spectral information using the high resolution data of the imagery were computed and included within the classification algorithms. Two "order" textural layers were derived:

  1. A first order layer that was a simple measure of local variance within different sized rectangular neighborhoods that were coupled with the multispectral bands.
  2. A second order layer that uses a gray-level co-occurrence matrix to derive three features: contrast, correlation and entropy (see Clausi, 2002)
  
Finally, an object-based approach was employed using the second order texture data. The image was segmented into objects, and then classification was applied to the resultant objects.

**Key results:** The results generally show that the high resolution multispectral datasets were effective in accurate classifications of mangrove species. In particular:

  - Standard deviation and entropy of the multispectral data for IKONOS was consistently higher than that of QuickBird, indicating that IKONOS captures richer spectral information than QuickBird
  - Classification using purely multispectral data was slightly better for IKONOS than QuickBird
    - Addition of the panchromatic band did not significantly improve accuracies
  - Inclusion of a first order texture band improved accuracies for the IKONOS image; HOWEVER, accuracies decreased for QuickBird imagery, with the exception of the largest window size
  - Second order texture data did not produce accurate classifications by themselves
  - Under the object-based approach (using textural information in panchromatic and multispectral bands), almost equivalent classification accuracies were achieved

It is important to note that the results do not support the expectation that the slightly higher spatial resolution of QuickBird imagery would result in "better" textural information for classification accuracies.

**Further reading:**

  - Textural analyses:
    - Clausi, 2002 - An analysis of co-occurrence texture statistics as a function of grey level quantization
    - Haralick *et al*, 1973 - Textural features for image classification

### Wang, 2004b {#wang2004b}

*"Integration of object-based and pixel-based classification for mapping mangroves with IKONOS imagery"* [@wang2004integration]

**Key significance:** This is a sister study to the study reviewed in Section @\ref(wang2004a). This study sought to examine three different classification methods:

  1. MLC at the pixel level
  2. Nearest neighbor classification at the pixel level
  3. A hybrid classification that combines pixel and object based methods (MLCNN)
  
They consider the conundrum that with increasing spatial resolution, the number of detectable sub-class elements increases as well and thus classification at the pixel level may deteriorate. Thus, they examine object-based classification as a means of getting around this particular issue.

**Research design:** For the site and imagery description see Section \@ref(wang2004a). The different classification approaches used in this study are described here:

  1. MLC at pixel level - maximum likelihood classification of raster imagery using pixels
  2. Object-based classification - Used eCognition 3.0 software (region merging technique that uses a homogeneity criteria to detemine whether pixels and regions should be merged).
    - The merging "breaks off" at a predetermined parameter (small parameter = small regions, large parameter = large regions), and there are in principle infinite choices of break-off parameters\*
  3. The authors develop a joint procedure in which MLC at pixel level is first used, in which spectrally inseparable classes are identified, and all spectrally separable classes are masked out. An object-based classification procedure is then performed on the spectrally inseparable classes.
    
\* To deal with the issue of infinate scale parameters, Wang *et al* develop a method to identify an optimal value. They use the training set data from the MLC image, reclassify all other pixels as 0s, and then use the image segmentation approach for several scale parameters. They then derive the pairwise Bhattacharya Distance values for the seven land cover classes at each scale, and select the scale at which BD reaches a maximum.
  
**Key results:** In general, the accuracy of the classifications were relatively high for all three methods; however, the mixed pixel and object-based method performed the best. The key results are:

  - The pixel based approach generally did well in correctly classifying the land cover classes (88.9% overall accuracy), with the exception of the red mangroves (71.3% accuracy), black mangroves (81.8%) and rainforest (81%)
  - The object-based approach improved classification of the rainforest (100.0%), red mangroves (77.5%) and black mangroves (89.8%), but greatly misclassified white mangroves (21.9%) due to their existence along edges with red and black and thus being pulled into "objects"; overall accuracy was relatively lower relative to the pixel based approach as a result (80.4%)
  - The mixed pixel and object-based method performed the best because white mangrove, road, gap and lagoon all were highly separable using MLC, but distinguishing between red mangrove, black mangrove and rainforest benefitted from the object-based method (due to high confusion between these classes)
    - The mixed method approach achieved overall classification accuracies of 91.4%.
  
  
**Key conclusions:** The following conclusions are of particular note:

  1. The pixel-based method retains the spectral information at a fine spatial scale, but does not excel at distinguishing between spectrally similar classes (rainforest, red & black mangrove)
  2. The object-based method averages spectral information, but in doing so is able to improve discrimination between spectrally similar regions (the difficult classes mentioned above)
    - However, you run the risk of averaging unrelated pixels (likely unavoidable)
    
### Vaiphasa, 2005 {#vaiphasa2005}

*"Tropical mangrove species discrimination using hyperspectral data: A laboratory study"* [@vaiphasa2005tropical]

**Key contribution:** At the time this study was written, hyperspectral studies of mangrove species classifications were relatively limited, and fairly inconclusive. This study seeks to examine whether mangrove canopy leaves have sufficient spectral information to discriminate mangroves at the species level. Sixteen mangrove species were examined.

**Research design:** Canopy leaves from 16 species were collected in Chumphon and hyperspectral signatures at 2,151 spectral bands from 400 - 2,500 nm were collected in laboratory conditions.

One way ANOVA was used to examine for differences of individual species from one another across all spectral bands, and spectral separability of the species using a Jeffries-Matusita distance analysis metric. Given the issue of a massive feature space (2,151 bands), four "optimal" bands for assessing spectral separability were identified using a wrapper feature selection approach.

**Key results:** The results generally show that discrimination between the sixteen species is generally achievable. Some particular key results include:

  - 1,941 of the 2,151 bands were found to be spectrally dissimilar across all species (ANOVA analysis) at th p < 0.05 level. Of these 1,941, 477 bands were significant at the < 0.01 level. The exceptions were primarily at the UV region (350-400 nm) and the shortwave infrared region (1,800-2,500 nm), where lighting conditions may have played an influential role.
  - The optimal band combination for classification was found to be:
    - 720 nm, 1,277 nm, 1,415 nm and 1,644 nm
    - These results indicate that other properties beyond pigment (which would correspond to the 720 nm band) may play important roles in discrimination, and may be a result of other ecophysiological factors (water content, salinity, sugar, protein, oil, lignin, cellulose, leaf structure, etc.)
  - Pair-wise separability of mangrove species was possible for many of the species using these four bands and the J-M distance metric, but reduced separability was seen for species of the Rhizophoraceae family. 
    - There was high confusion both amongst species within Rhizophoraceae, as well as with *S. ovata*, *X. granatum* and *Pluchea indica*.
    
**Further reading:**

  - Wayne and Le's study of lab hyperspectral indices: 
    - Wang *et al*, 2009 - Distinguishing mangrove species with laboratory measurements of hyperspectral leaf reflectance


### Vaiphasa, 2006 {#vaiphasa2006}

*"A post-classifier for mangrove mapping using ecological data"* [@vaiphasa2006post]

**Key contribution:** This is one of the first studies to employ environmental data (pH) to refine classification maps of mangrove species. The study finds that classification accuracies are increased significantly, from 76.04% to 88.21%.

**Research design:** The study coupled an existing mangrove species map produced from classification of ASTER imagery with a map of pH throught Pak Panang mangroves to examine the effectiveness of a post-classifier. In particular, they collected 200 pH measurements distributed throughout the forest, 150 of which were used for interpolation of a salinity profile, and 50 of which were used for validation. They additionally collected species composition and forest structure data at over 500 plots of 15 x 15 m^2^ in the forest, and used these for accuracy assessment of the maps.

The post-classifier was based on a Bayesian approach that employed probabilities of species given a pH level, and pH level given a species.

**Key results:** Accuracy of the mangrove species map was improved significantly with the addition of the pH data. In particular:

  - Overall accuracy was improved from 76.04 - 88.21% following inclusion of pH data
  - Speckling within the original mangrove species map was greatly reduced following inclusion of the pH layer
  - Confusion between *R. mucronata* and *S. caseolaris* persisted despite the inclusion of the pH data
  - In reference to the results from Vaiphasa (2005), they found that the addition of the pH data was able to improve the separability of species within the Rhizophoraceae family (which exhibited high confusion following classification of hyperspectral data)
  
The authors conclude that the method presented (classification followed by updating with ecological gradient data) may be an efficient means for producing detailed mangrove species maps in areas that are not too large (e.g., < 100 km^2^).

### Wang, 2009 {#wang2009}

*"Distinguishing mangrove species with laboratory measurements of hyperspectral leaf reflectance"* [@wang2009distinguishing]

**Key contribution:** Le and Wayne largely repeat the Vaiphasa, 2005 study (see Section \@ref(vaiphasa2005)) but for the three mangrove species common to central America. In addition, they examine the ability of hyperspectral data to estimate stress conditions in mangrove trees.

**Research design:** They collected leaves from all three species in both stressed and unstressed conditions, and also collected leaves from the top and bottom portions of the canopy. They then subjected each of the leaves to hyperspectral measurements in a laboratory setting before analyzing the data.

To reduce the feature space prior to spectral discrimination, they first used the ANOVA across all spectral bands to remove any bands that were non-significant at the p < 0.01 level (*a la* Vaiphasa, 2005). They then use Linear Discriminate Analysis (LDA) to examine for differences amongst the species.

In addition, they use several indices of spectral bands identified by prior work as indices of leaf stress. Chlorophyl a and b contents drop dramatically in stressed leaves, greatly changing the reflectance properties of leaves. Studies have found that ratios are best in identifying stress characteristics, and they use these within ANOVA to compare means of ratios between stressed and unstressed leaves.

**Key results:** The results generally find that hyperspectral data is effective in species discrimination and also identification of stressed mangrove trees. Some particularly key results are as follows:

  - 116 of the 215 bands were significant at the p < 0.01 level for distinguishing between species and were clustered in the 350-510 nm, 610-690 nm, 760-810 nm, 1,370-1,550 nm, and 1,850-2,500 nm ranges of the spectrum
  - Red, white and black mangroves were capable of being discriminated using the hyperspectral data
  - Use of the ratios for identifying stress in mangrove trees was effective, with different ratios performing better for different species (3/4 for red, 1/4 for white, and 4/4 for black)
  - The specific band widths identified as significant in this study were different from those of Vaiphasa, 2005

**Key conclusions:** Some key, broader-level conclusions from the study are as follows:

  1. Conventional challenges to analysing hyperspectral data are the large number of highly correlated bands available coupled with relatively smaller numbers of training and test samples.
  2. Wang and Sousa conclude that use of hyperspectral data is effective in overcoming the shortcomings of the multispectral datasets that they use in their 2004 sister studies.

### Kamal, 2011 {#kamal2011}

*"Hyperspectral data for mangrove species mapping: A comparison of pixel-based and object-based approach"* [@kamal2011hyperspectral]

**Key contribution:** This study employs high resolution, hyperspectral data to map mangroves and wetland land classes at a landscape scale in Australia using three different methods. There are relevant points they make but the study feels somewhat lacking in general.

**Research design:** CASI-2 hyperspectral data were used as the imagery (30 bands and 4 m spatial resolution) in this study.

An existing digital map of mangrove species produced via photographic interpretation of aerial imagery, site data, and traverses was used for endmembers selection and accuracy assessments of the classifications.

They use three different classification procedures to produce different maps of the landscape:

  1. Spectral angle mapper -- uses a spectral library representative of different end members (i.e., the 9 targeted wetland classes) and assigns pixel to each based on a similarity measure
  2. Linear spectral unmixing -- assumes each spectral signature is a linear combination of area weighted sum of reflectance of endmembers; total areas or pixels may then be assigned based on probabilistic basis
  3. Object-based mapping -- used OBIA to segment the image before applying classification algorithms on the spectral signatures
  
**Key results:** Of the three methods examined, OBIA performed the best (overall classification accuracy of 76%). Linear spectral unmixing produced the lowest overall classification accuracy. 

**Conclusion:** The classification accuracy is relatively lower than those found in other studies, and thus the results feel perhaps a bit short-sighted. Other metrics that could have been interesting to explore would have been texture given the high resolution nature of the dataset. Additionally, they do not dig into the hyperspectral bandwidths that are relevant, and thus there is valuable information that is missing from the study.

Additionally, it is not entirely clear to me whether they used the coarse level reference map for their accuracy assessment, or whether field data was collected. If the reference map is of low-quality, their results are rather meaningless.

### Dronova, 2012 {#dronova2012}

*"Landscape analysis of wetland plant functional types: The effects of image segmentation scale, vegetation classes and classification methods"* [@dronova2012landscape]

**Key contribution:** This study examines the classification accuracy of plant functional types (PFTs) under different image segmentation scales and classification algorithms for Poyang Lake in China. The key results generally show that use of a coarser image scale (on average 5-8 Landsat pixels) significantly improved image classification of the different PFTs. Classification between the different algorithms were less significant.

**Research design:** Field data for training and test validation were collected, and Landsat 5 TM data (bands 1-5 & 7) were used for all image classification procedures. Image segmentation was performed at 9 different scales from (2 - 10), and six different classification algorithms were used:

  1. Bayesian
  2. Logistic
  3. Neural network
  4. Decision trees
  5. K-Nearest neighbors
  6. Support vector machines
  
The image classification procedure was performed under two scenarios, the first being for generalized PFTs (3 classes), and the second being for specific PFTs (6 classes). Water, sand and other development were excluded prior to classification. Cluster analysis was performed on morphological and physiological data of the 51 species within the landscape to identify the different classes.

Additionally, texture metrics within objects were computed and included in the classification procedure to examine their utility.

**Key results:** The results found that OBIA generally improved the results of the classification for both the 3-class and 6-class sets Some specific results include:

  - A scale of 8 was found to produce the highest classification accuracies for both classification sets (i.e., objects of 5-8 Landsat pixels)
    - However, at the specific PFT level, the optimal scale parameter was found to be variable
  - Texture metrics were not found to significantly improve classification accuracies, indicating that discrimination between the various classes was primarily spectrally based
  - Overall classification accuracy was relatively similar for the different classification algorithms
  
The authors conclude that smoothing of spectral signatures within objects lends to greater discrimination between PFTs, aiding in overall classification.

Texture metrics may have proven to be insignificant in improving classification accuracies due to the relatively coarse nature of the Landsat data. Most other studies that have found texture to be significant have used high resolution imagery.

Human interference and ecological conditions that produced mixed pixels resulted in reduced classification accuracy, and the authors conclude that mixed approaches (i.e., object-based followed by pixel-based, *a la* [@wang2004integration]) may be an interesting means of resolving this issue. However, propagation of uncertainty has been under-explored and will be important in such scenarios.

## Geomorphology {#rs-geomorphology}

### Hess *et al*, 1990 {#hess1990}

*"Radar detection of flooding beneath the forest canopy: A review"* [@hess1990radar]

**Key significance: ** This reviews many of the earliest studies that employ SAR imagery for mapping of inundated versus non-inundated forest types.

**Key notes: **

Of key importance for mapping of flooded vs. non-flooded forests is the fact that SAR imagery has a much more intense return off of water versus dry soil or trunks. Thus, the "double-bounce" effect off of a water surface is much stronger than that from soil & then tree.

An excellent schematic of this effect is shown below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/hess1990.bmp"))
```

*Accuracy of flood detection*

Some have found an issue in that canopy that is too dense may attenuate the signal at too great of a degree, and thus the difference between the more intense return from flooded forests may not be apparent (L-band).

Additionally, presence of herbaceous vegetation (e.g., salt marshes) adjacent to flooded forests may erroneously appear as "flooded" forest using L-band. The bright return may be a result of the fleshy, vertically-oriented leaf stalks. These areas are still flooded, however, and thus it may still be effective for mapping of flooding extent, perhaps just not between flooded forest vs. flooded non-forest.

*Magnitude of enhancement*

Enhancement from flooded forest may range from approximately 2 - 10 dB.

*Effect of radar parameters*

Of primary importance are three considerations:

  1. Incidence angle - angles at which imagery is collected vary greatly, and may be an important parameter for efficacy of detecting flooding
    - The most effective angle may depend on the ecosystem type. Flooded forests versus swamps may exhibit different dominant backscattering mechanisms.
  2. Polarization - the polarization of the radar may effect the efficacy of the return in identifying inundated landscapes 
    - Majority of studies have used HH polarization (Seasat, SIR-A, SIR-B)
    - Qualitative assessments are consistent in findings that like-polarized scenes (HH or VV) are more effective at identifying flooded forests via L-band
  3. Frequency - use of P-, L-, or X-band may have different dominant backscattering mechanisms, and thus may influence detection of inundation.
    - L-band commonly used due to ability to penetrate dense canopies, however may have some limitations
    - P-band may be promising for mapping of inundated forests in forests that are very tall or with very dense canopies

*Applications in flooded forest research*

Critical to examine the temporal and spatial variability of water cover. An important component is availability of ground data on extent and time of flooding at the time of imaging, particularly tricky for tidally inundated forests.

Additionally, for forests with aboveground root structures (i.e., mangroves), scattering of the signal by aboveground roots may reduce the signal and reduce ability to identify inundated forests.

### Walsh, 1998 {#walsh1998}

*"An overview of scale, pattern, process and relationships in geomorphology: A remote sensing and GIS perspective"* [@walsh1998overview]

**Key contribution: ** This study exemplifies the "scale, pattern and process paradigm" for exploration of geomorphological relationships in two study systems: alpine and fluvial environments. The authors note the value of remote sensing and GIS, as well as their integration, for examining processes that may occur over different scales.

The authors note the basic intent of the study is to examine use of space-time scales for organizing and examining geomorphic and biogeographic processes via a RS-GIS system framework.

**Key notes: **

The study is motivated by early theoretical work in landscape ecology, which functions largely under the premise that ecological patterns and processes are largely dictated by the spatial arrangement and patterning of a landscape.

The authors note several key values of GIS work for landscape processes, in particular:

  1. Examination of spatial vs. non-spatial relationships via explicit tools (overlays, etc.)
  2. Represent an array of landscape perspectives through integrating geographically registered spatial coverages
  3. Efficient display of information
  4. Examine co-occurrence of spatial and non-spatial data
  5. Display singular thematic or composite coverages
  6. Can model location and response of phenomena

Critical consideration is the scale at which ecological processes happen, and the scale of the dataset (e.g., 30 x 30m pixels) that one is working with.

The paper provides a review of remote sensing and GIS, which I have largely omitted here except for a few key points.

GIS can be used to produce spatial metrics (quantification of the composition and spatial pattern of landscape units), which may help assess degree of spatial organization within a landscape. As the authors refer to it, these are largely the precursor to what is now termed object based image analysis.

Landscape level pattern indices may include: dominance, contagion, fractal dimension, or lacunarity indices.

Lacunarity incorporates overall fraction of map covered by a phenomenon, degree of contagion, presence of self-similarity, scale of randomness and presence of hierarchical structure. 

In essence, lacunarity describes the degree of similarity of regions within a geographic area at a predefined scale.

Spatial metrics at landscape level can be generally understood to provide information on landscape heterogeneity indices due to measurements of overall landscape structure within defined area.

GIS can also be useful for assessing scale dependencies. Semivariance and fractal dimensions are used to assess nature and degree of spatial autocorrelation within landscape elements over different spatial scales.

Examples of overall scale dependence that may influence a study are:

  1. The precision with which scales and the region of interest are defined
  2. Nature of the variables that are examined (local scale may be physical variables, regional may be climate or land use)
  3. Degree of scaling (either up or down) that is attempted
  
Explicit recognition of the importance of scale has led to the use of hierarchy theory within landscape ecology studies (may want to read up more on this).

*Fluvial case study*

The authors describe the different scales of environmental processes that shape the meandering and development of alluvial systems. They note the importance of both spatial and temporal scales.

Fine spatial scales may be seen in channeling or sedimentation along particular points within a river, and are likely due to interactions of flow and sediment properties at fine scales.

At coarse spatial scales, development of abandoned channels are controlled by water flow across floodplains and proximity of site to the channel.

Additionally, for temporal scales, sharp pulses in water flow or longer term gradual shifts may have different effects on an alluvial system.

### Imhoff, 1990 {#imhoff1990}

*"The derivation of a sub-canopy digital terrain model of a flooded forest using synthetic aperture radar"* [@imhoff1990derivation]

**Key significance: ** This is one of the earliest uses of SAR imagery to map digital terrain models and understand tidal inundation in a mangrove forest. The study couples both SAR imagery as well as field measurements of tidal inundation in a novel manner, and provides discussion of syncing the timing of imagery acquisition with hydrological records.

**Key notes: **

The study employed data from both the SIR-B mission as well as tide surface information to create a digital terrain model in southern Bangladesh.

They note the importance of hydrology and geomorphology on mangrove health, citing Lugo and Snedaker's discussion in which they note that "tidal dynamics" and "water chemistry" are the two most important factors on mangrove health.

The radar data used for the study consisted of three HH polarized L-band radar images, acquired at 26, 46 and 58 degree angles of incidence.

The field data comprised of tide gauge data from five stations throughout the region slightly larger than the survey area, a 1,200 m topographic survey transect, and test plots in the forest that were visited during the radar data acquisition period.

DN counterparts to mean backscatter coefficients were used to select break points for classifying flooded versus non-flooded forests.

*Results*

All three datasets were coregistered, and field data were used to train a model on flooded vs. non-flooded forest areas.

A spatial filter was applied using a 5 by 5 and then 11 by 11 median filter to remove speckling. The processed image allowed for better discriminating between classes at a spatial frequency appropriate to the formation.

The 58 and 46 deg images were combined to produce an inundation map (the 26 deg data had issues during time of collection, but was used for validation of land edge boundaries), and subsequently into a topographic map.

In doing so they produced numbers along the inundation boundaries, which took on values of elevation level relative to mean sea level. The values of the pixels were used as points and an interpolation procedure was used to produce a continuous DTM.

The authors then discuss various sources of error, which include resolution issues, measurement error, as well as imprecise measures of "flooded" vs "non-flooded" forest recognition by the L-band SAR imagery.

### Souza-Filho & Paradella, 2002 {#souzafilho2002}

*"Recognition of the main geobotanical features along the Bragan\c{c}a mangrove coast (Brazilian Amazon Region) from Landsat TM and RADARSAT-1 data"* [@souza2002recognition]

**Key contribution: ** This study exemplifies integration of Landsat and SAR data for mapping of different ecosystem types and successional classes along the Bragan\c{c}a coastline at a landscape scale. Incorporation of landsat and SAR data enables mapping of not only different ecosystem types, but also successional classes (e.g., young vs. mature intertidal mangroves, degraded and renegerated mangroves, etc.).

**Key notes: **

The authors process Landsat TM and RADARSAT-1 SAR data, and build a composite product using an "Intensity, Hue and Saturation" transform. The transform uses a RGB/IHS transform that incorporates channels formed from PCA of Landsat bands 1, 3 & 3; bands 5 & 7; band 4; as well as an intensity band from the RADARSAT-1 data (described in Harris et al., 1994, Paradella et al, 1998, Pohl, 1998).

The resulting map incorporates the spectral data as well as the intensity data corresponding to SAR topography, with an array of colors and values across the landscape. Using field data and prior knowledge, the authors then visually interpret the map into a variety of different land features.

*Key criticism* - they do not provide a full accuracy assessment and thus it is unclear how accurate their results are. The map provides much more information (colors) than is interpreted, and thus it is unclear how legible the map necessarily is (classifications may be too fine scale).

They produce a map with 19 classes of the landscape, but it is unclear how they arrive at this point. Assuming that they used the integrated product and supervised classification.

Nevertheless, if we take the procedure for granted, they are able to distinguish between a variety of geobotanical features, including: channels, open water, mudflats, coastal dunes, young intertidal mangroves, intertidal mangroves, supratidal mangroves, degraded mangroves, outer salt marshes, and urban areas.

The final map classification as they provide it is shown below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/filho2002.bmp"))
```

*Conclusion*

The methods are a little black-boxey and may require digging into the associated PhD thesis (in Portuguese), but the map is an interesting approach to pulling information from both spectral (Landsat TM) as well as radar-based (RADARSAT-1) information.

### Souza-Filho, 2003 {#souzafilho2003}

*"Use of synthetic aperture radar for recognition of coastal geomorphological features, land-use assessment and shoreline changes in Bragan\c{c}a coast, Para, Northern Brazil"* [@souza2003use]

**Key contribution: ** This study is a sister study to that of [@souza2002recognition], and examines mapping of geomorphology as well as changes in coastal extents over time. The study uses SAR data acquired from over a 30 year period to investigate areas of coastal progradation versus erosion.

**Key notes: **

The difference in SAR return from land versus water is largely a function of contrast of dielectric properties as well as surface roughness, and allows for striking interface along coastal landscapes.

The two SAR imagery datasets used were airborne X-HH band data from 1972, and orbital (satellite-based) C-HH band data from 1998. Both images were acquired at high tide conditions.

Imagery were ortho-rectified based on ground points that were obtained during a field campaign.

The SAR imagery from 1972 was used to distinguish between terrain and open water, converted to a vector dataset, and overlaid on the 1998 imagery to examine changes in shoreline over time.

The authors provide a nice overview of the steps that must be taken for use of SAR data (i.e., processing):

  1. Input data
    - Reading of image orbit data and 16 bit RADARSAT data
  2. Preprocessing phase
    - Radiometric correction - scaling of 16 bits to 8 bits, ground control point selection
    - Geometric correction - math model for geometric correction, geometric ortho-correction + speckle reduction
  3. Enhancement
    - Linear stretch
  4. Assessment of SAR product
    - Radargeologic interpretation
    
*Results - Mapping of land cover classes*

The authors used visual interpretation to identify six prominent geomorphological coastal features. Intertidal vs. supratidal mangroves can be determined given different microwave responses to vegetation. Supratidal mangroves are shorter and more open, and thus result in a reduced double-bounce effect relative to intertidal mangroves.

They used aerial imagery in conjunction with the SAR data to help validate some of their land classifications.

*Results - Mapping of shoreline change*

Other studies have noted that mapping shoreline change during Spring High Tide Level is most appropriate, thus both images were acquired at same high tide condition.

The authors conclude that the shorelines has been subject to severe erosion, but I'm also not sure how they distinguish between this and sea level rise during the time period. They identify both areas of erosion as well as deposition, but with a net effect of erosion in the study region.

They note that deposition of sand has killed mangrove vegetation, and thus the mangrove terrace has largely been eroded.

They also note that in other areas, sediment supply and protection from wave attacks has allowed for accretion of sandbanks, muddy deposition and establishment of mangroves.

The full map or erosion and accretion in the area is shown below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/filho2003aabc.bmp"))
```

The authors use the maps of erosion and accretion coupled with ecosystems to understand processes that may be occuring within the landscape. The authors also note, however, that much of the findings are speculations given the absence of wind, wave, and tidal currents data within the region.

Time series analysis of SAR data thus may be informative in investigating long-term trends along coastlines, particular for identifying regions of erosion or deposition.

### Knight, 2009 {#knight2009}

*"Exploring LiDAR data for mapping the micro-topography and tidal hydro-dynamics of mangrove systems: An example from southeast Queensland, Australia"* [@knight2009exploring]

**Key contribution: ** This is one of the very few studies that employs airborne LiDAR to derive a high resolution digital elevation model and examine inundation and hydrology characteristics within a mangrove forest.

**Key notes: **

The study was motivated by the need to map habitats for a disease vector mosquito.

The digital terrain model that they obtained had a vertical accuracy of 0.5 m, and was validated by using hydrological modeling and field data of tidal gauges over ten days.

The accuracy of the DTM and that of field data indicating habitats for mosquitos were found to validate one another.

Of key geomorphological importance is the fact that they were able to identify a ridge that ran through the majority of the site as well as back basins and fringing mangroves. The back basin was blocked from tidal waters during all times except for the highest tides. Such understandings may be important for understanding the spatial ecology of a site.

The authors also note the importance of the LiDAR derived DTM for understanding effects of SLR on the mangrove forest.

*Study design*

LiDAR was flown over the area, producing a shapefile with 3 million points. The points were classified as ground versus non-ground by removal of points not contributing to ground terrain points by +/- 0.1 m. The ground point data was validated against GPS base-station data. 150 points were used to assess vertical accuracy across the entire area.

The tidal inundation data was collected from 10 points in the field, recording water depth, pressure and temperature every 5 minutes continuously for 10 days. Locations of the tidal gauges were informed by the DEM, selecting for areas near to tidal sources, in depressions or on hummocks. The tidal depths were used to validate inundation patterns expected given the LiDAR DTM.

Tidal inundation was modeled in ArcMap by "flooding" of the landscape in 0.05 m increments, with flooding constrained to adjacent to water bodies (i.e., the back basin was not flooded until the berm was breached).

*Results*

The use of the DTM was able to identify the presence of a berm as well as back basin. The below figure displays the findings. The top figure displays the elevation along a transect that was conducted along the center of the berm. A minimum elevation can be seen that must be breached for flooding of the back basin.

```{r, echo=FALSE}
knitr::include_graphics(rep("images/knight2009fig2.bmp"))
```

The total DTM for the site can be seen below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/knight2009fig1.bmp"))
```

The authors conclude that use of LiDAR data for derivation of the DTM was successful, and that the approach that they took may work well for modeling of shifts in hydrology that may occur with sea level rise.

### Smith, 2009 {#smith2009}

*"Applications of remote sensing in geomorphology"* [@smith2009applications]

**Key contribution:** A review of the use of remote sensing for geomorphology, without a particular focus on mangroves. The article reviews the current platforms and data types, and maps them to their relevant uses within geomorphology.

**Of particular note:**

Remote sensing particularly well placed to aid studies of geomorphology in:

  1. location and distribution of landforms
  2. land surface elevation
  3. land surface composition
  4. subsurface characterization

```{r echo=FALSE}
knitr::include_graphics(rep("images/smith2009ProgPhysGeog.bmp"))
```

The authors remark consistently on the added value of digital elevation models (both as products, as well as automatic classification by sensors) for geomorphology.

Among most important techniques for postprocessing of LiDAR point clouds for geomorphologists is removing "surface clutter" to extract actual ground surface.

**Note:**

  - Look further into mangrove sites on [OpenTopography](http://opentopo.sdsc.edu/datasets)

**Further reading:**

  1. Clark, 1999 - Spectroscopy of rocks and minerals, and principles of spectroscopy
  2. Sithole and Vosselman, 2004 - Experimental comparison of filter algorithms for bare-Earth extraction from airborne laser scanning point clouds


### Balke & Friess, 2016 {#balke2016}

*"Geomorphic knowledge for mangrove restoration: A pan-tropical categorization"* [@balke2016geomorphic]

**Key contribution: ** This study uses remotely sensed datasets to characterize geomorphic settings across the tropics with the goal of identifying differences that are relevant for restoration projects. The study uses spatial layers of both suspended sediment as well as tidal range to primarily classify the settings, and emerges with three different classes:

  1. Minerogenic and high tidal range (allochthonous)
  2. Minerogenic and low tidal range (allochthonous)
  3. Organogenic and low tidal range (authochthonous)
  
**Key notes: **

Primarily three different datasets (soil organic matter, total suspended matter, and tidal range) were used for the classifications of different geomorphic settings.

*Study design*

The first dataset is that of soil organic matter, and was compiled from the academic literature.

The second dataset is total suspended matter, which was derived from the MERIS spectrometer (Medium Resolution Imaging Spectrometer) aboard the European Envisat satellite. The data employed was a product of TSM available from ESA's Global Ocean Colour for Carbon Cycle Research Project.

Tidal range data used was a layer built from the FES2012 (Finite Element Solution-Global tide) model, which is one of the more accurate models for shallow continental shelf locations.

For each SOM value, a 50 km radius was used to average TSM and tidal data to characterize the site. K-means clustering was used to separate the data into three different groups and thus define borders between sites of different character.

*Results*

TSM values were highest near outlets of major rivers.

Multiple linear regression showed that SOM was negatively associated with increasing tidal range and total suspended matter. 

The k-means clustering and classification of sites from which they had data are shown below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/balke2009epsl.bmp"))
```

*Key findings*

Their results show that sites with SOM > 60% were found mostly in the Caribbean, Hawaii and Florida. Sites elsewhere had average OM values of ~10%, and thus broad description of mangrove soils as "peat" is misleading.

The broad majority of mangrove soils globally are minerogenic.

They then describe a variety of relevant mangrove restoration and management techniques relevant for the various mangrove geomorphic settings.

In organogenic settings, key risk is collapse of peat following disturbance. Influences on root productivity such as nutrient availability as well as inundation levels are of key importance. 

They note that low tidal ranges often indicate the importance of seasonal flooding to the site. Management and restoration in microtidal sites is thus often concerned with preserving networks of terrestrial freshwater sources.

Minerogenic sites require high levels of sediment input to maintain intertidal surfaces at correct elevation relative to tidal frame, and thus proper management of hydroperiod at rates which mangroves prefer. Disturbances in sediment supply may influence long term health of mangroves. Use of bamboo fences in Vietnam and Thailand may help to reduce tidal energy and wave energy and promote sediment deposition.

Interesting that they note the Chao Phraya river has undergone a ~75% reduction in sediment yield, with severe coastal erosion resulting.

## Other

The following studies provide key information on use of remote sensing as it relates to wetlands, but do not necessarily fit into the mangrove-specific studies of mapping "vegetation" or "geomorphology".

### Lucas, 2007 {#lucas2007}

*"Potential of L-band SAR for quantifying mangrove characteristics and change: Case studies from the tropics"* [@lucas2007potential]

**Key contribution: ** As a precursor to the Lucas *et al* (2014) study on L-band SAR and Global Mangrove Watch, this study examines the use of L-band SAR data in mangroves in Australia, French Guiana and Malaysia to characterize forest structure and changes in extent.

**Key notes: **

*Overview notes on SAR data*

SAR operates at X-, C-, L-, or P-band frequencies in both polarimetric and interferometric formats.

"Polarimetric systems measure the full scattering matrix from which the intensity and phase of scattered fields can be derived for any polarization". The backscattering coefficient relates to intensity whereas the phase information indicates phase difference and correlation coefficients between signals.

SAR data may be used for *polarimetry* or *interferometry*:

  - Polarimetry - measurement of phase difference and correlation between different polarizations; measurement of phase allows entire complex electromagnetic field to be recomposed
  - Interferometry - measurement of phase different and correlation of two identical radar signals from slightly different locations; allows for construction of DEMs
  
The backscattering coefficient of SAR data depends upon interactions of microwaves of varying configurations with components of vegetation that differ in qualities (size, dimension, density, dieletric constant).

Interactions of microwaves with surfaces is often described in terms of the scattering mechanism, which ranges from direct (i.e., direct return with little diffuse scattering) to diffuse (rough surfaces that scatter the signal in many directions).

*Case studies*

In using SAR data, detection of zonation at the species level is more complicated because the information provided is purely structural in nature.

Differences in signal from truly different structural forms versus differences in growth stage and form should be considered when distinguishing species or zones from SAR data.

Mapping of mangrove zones may be complicated by variations in growth stage/form and tidal inundation at the time of observation.

*Forest structure*

May be expressed in terms of integrative metric of aboveground biomass, or more specific metric such as height, basal area, density. 

For SAR data, the viewing angle may be an important criteria in identifying differences in different structures. 

In the case studies, the backscattering coefficient was found to be enhanced towards the seaward zone due to both a more open canopy as well as the double-bounce effect from tidal inundation.

Backscattering coefficient for approximating biomass may saturate at high biomass values (120-140 Mg/ha). 

The different interactions of different bands of microwave and various components of vegetation have been hard to untangle due to the interactions of the components.

*Potential of L-band PALSAR for mangrove assessment*

ALOS PALSAR is collecting data in HH and HV polarizations, which allows for better understanding of the landscape.

"Quad-polarized" - implies availability of both amplitude and interchannel phase information

"Selective polarization" or "selective single polarization" - imply availability of amplitude data but no interchannel phase information

The authors note that use of ALOS PALSAR across sites will be greatly complicated and thus the use of site-specific ancillary data such as species maps, or optical imagery will be important in understanding variation in forest structure and composition.

### Lucas, 2014 {#lucas2014}

*"Contribution of L-band SAR to systematic global mangrove monitoring"* [@lucas2014contribution]

**Key contribution: ** This paper provides an overview of the Global Mangrove Watch program, which primarily employs L-band SAR data to monitor and report on local to global changes in extent of mangroves.

**Key notes: **

The study justifies the use of SAR under the premise that spectral data is limited in tropical regions and thus ill-suited for global monitoring efforts. 

Long-term datasets that are available for C-band SAR data are: 

  1. European Space Agency's (ESA) European Remote Sensing (ERS) Satellite 1 (1991-2000) 
  2. European Space Agency's (ESA) European Remote Sensing (ERS) (1995-2011) Advanced Microwave Instrument.
  3. ENVISAT advanced SAR (2002-2012)
  4. Canadian Space Agency (CSA) RADARSAT-1 (1995-2013)
  5. CSA RADARSAT-2 (2007-present)

However, C-band SAR is relatively limited for monitoring of mangroves; structural studies limited to just the canopy.

Tandem-X (German Space Agency) is the key source for X-band SAR data, and may be coupled with the SRTM DEM to derive canopy height changes.

Primary L-band SAR datasets that have been operating at global scales are:

  1. Japan Aerospace Exploration Agency's - Japanese Earth Resources Satellite (JAXA JERS-1; 1992 -1998)
  2. JAXA Advanced Land Observing Satellite Phased Array L-band SAR (ALOS PALSAR-1; 2006-2011)
  3. JAXA ALOS PALSAR-2 (2014-present)
  
Interestingly, the authors note that mangroves may be an indicator ecosystem for impacts from regional to global climate change.

The study reviews several relevant environmental policies for the GMW program:

  1. Ramsar Convention
  2. Convention on Biological Diversity
  3. IPCC REDD+
  4. IPBES
  5. Convention on Conservation of Migratory Species of Wild Animals (CMS)
  
*Addressing the needs*

Early efforts for producing global maps of mangrove coverage include efforts by: 

  - Spalding et al. 1997, based on expert maps, some RS, and point estimates
  - Spalding et al., 2010, updated effort largely employing the same data sources/types
  - Giri et al., 2011, based largely on RS Landsat
  
Other efforts that seek to fill other desires exist:

  - Mangrove Capital - seeks to bring awareness to value of mangroves
  - GLOMIS - Global Mangrove Database and Information System
    - searchable database of scientific literature relating to mangroves

However, still no system to provide accurate and reliable trends in *changes* of mangrove extent and structure at regional to global scales.

*Value of L-band SAR for mangrove mapping*

Generation of early maps from optical imagery is critical as baseline effort since SAR data cannot distinguish between mangroves and terrestrial forests well.

Due to SAR's ability to provide information on the structure of a forest, estimates of biomass as well as segments of forest with or without prop roots can be achieved.

Furthermore, given the ability to penetrate cloud cover and global coverage of some datasets, change detection analyses are facilitated through the use of SAR. Change may be a function of natural processes, which can be elucidated well in protected areas (e.g., USA or Australia), as well as more anthropogenic or immediate impacts.

*Validation*

In validating change, main variables to consider are extent, structure and AGB of mangroves. Often use RS data of different types (high resolution imagery, LiDAR, SAR, multispectral imagery) for validation given the difficult field conditions that exist within mangroves.

In future, important for validation activities to co-occur with collection of SAR data to help validate the RS datasets.

*Conclusion*

GMW uses primarily L-band SAR data from JERS-1 SAR, ALOS PALSAR, and ALOS-2 PALSAR-2 (all Japanese) to monitor and map changes in mangrove extent and structure.

The key advantage of L-band SAR over C-band and X-band SAR is its sensitivity to woody components rather than vegetative components. 

### Dronova, 2015 {#dronova2015a}

*"Mapping dynamic cover types in a large seasonally flooded wetland using extended principal component analysis and object-based classification"* [@dronova2015mapping]

**Key contribution: ** This study addresses the issue that remote sensing imagery is often collected at a single point in time, but ecosystems are inherently variable and are constantly changing over multiple time scales (hourly, seasonally, yearly). The authors propose a new mechanism for embracing the dynamic nature of the landscape and improving classification results based on that.

**Key notes: **

The authors use different forms of PCA to examine temporal patterns in space (S-mode PCA), prevalent spatial patterns over time (T-mode PCA), as well as shared spatial and temporal patterns among different datasets (extended PCA). 

One of the key interests is distinguishing between short-term changes (phenology) versus longer-term changes that may indicate changes in ecosystem health, state parameters, or ecosystem function.

"Dynamic cover types" (DCTs) - distinct sequences of wetland cover states and transitions observed within a given period of change cycle.

The study focuses on one flood cycle from summer 2007 to spring 2008 and characterizes seven dominant DCTs that are representative of different water coverages, surface compositions and plant phenology that were identified during prior research.

*Study design*

Their methods were organized as such:

  1. Performed T- and S-mode extended PCA transforms of input RS data and examined which spatial and temporal patterns corresponded to DCTs
  2. Segmented input image and applied and assigned training and test data for DCT mapping
  3. Examination of statistics to determine DCT class differences
  4. Supervised object-based classification of DCTs
  5. Accuracy assessments of classification and uncertainty
  
Within the Poyang Lake area, there is massive change in lake and inundation levels, and five very broad coarse land covers may be described: live vegetation, dead vegetation, water, sand and mudflats.

Note that this study site is primarily herbaceous in cover, and thus the degree of dynamism that occurs (during wet season) may be more pronounced than in mangroves in which dominant vegetation is tree cover.

*Dynamic cover types*

Seven different dynamic cover types were defined, based on key ecological processes as well as the interests of the study:

  1. Ephemeral mudflats, non-vegetated
  2. Ephemeral mudflats, vegetated
  3. Emergent C3 grassland, flooded
  4. Mixed C4-C3 grassland, partially flooded
  5. Winter aquatic macrophytes
  6. Permanent water
  7. Permanent sand
  
The RS datasets that were used were Beijing-1 microsatellite multispectral data (7 scenes), and Envisat advanced SAR C-band in like polarization (HH or VV). The multispectral data was the primary imagery whereas the SAR imagery allowed for better delineation between murky waters and mudflats.

The spectral data were used to create NDVI and NDWI indices for the landscape.

The NDVI, NDWI and SAR data were used as inputs to the extended PCA analysis. They "performed standardized EPCA on these three time series using spatial (S-mode) and temporal (T-mode) forms of PCA orientation". 

T-mode analysis outputs principal components as images for each time series that highlight the dominant spatial patterns over time.

S-mode analysis reveals recurrent temporal patterns in the study area.

*Classification*

Classification was performed following OBIA, which helps to reduce the salt and pepper effect that is common to wetlands.

*Results*

The results produced maps with quite high classification accuracies (>90%). Thus, the authors conclude that use of prior ecological information and the mapping of dynamic cover types is deemed to be an appropriate approach for producing maps of ecological classes in dynamic landscapes.

The T-Mode EPCA primarily explained variance through the first principal component (94.7% of variation), which correlated with differences in prevalence of green emergent vegetation for the NDVI data, and gradients of submersion time for the NDWI and ASAR data.

The S-mode EPCA highlighted temporal changes in the landscape. The first component (37.7% of variation) highlighted changes between August 2007 and late spring 2008, whereas the second component (35.4%) highlighted change between mid-winter and late spring 2008.

The major sources of error were mutual confusion of emergent C3 grassland with mixed C4/C3 grassland, as well as ephemeral non-vegetated mudflats with vegetated mudflats, permanent water, and winter aquatic macrophytes.

The DCTs helped identify different "schedules" for similar classes, i.e., different trajectories over time that occurred.


### Dronova, 2015 {#dronova2015b}

*"Object-based image analysis in wetland research: A review"* [@dronova2015object]

**Key contribution:** This is an excellent and comprehensive review of OBIA as it relates to wetlands, and provides an overview of how researchers are currently using OBIA in wetland analyses, some of the relative strengths of it, and some of the challenges.

**Key notes: **

There are several extensive sections in the paper per different topics, and I follow the structure in my summarization of the article here.

*Background on general OBIA principlies*

OBIA segmentation may be performed through top-down or bottom-up algorithms, or a combination of the two. 

Given that "typical" objects might be difficult to identify within complex landscapes such as wetlands, "primitive objects," which are just a first pass at removing some of the variation and creating objects slightly larger than the pixel level may be used. The "primitive objects" then become the unit of analysis. 

Segmentation algorithms are controlled by parameters that are subjective and variable across studies. This is a key concern and area for additional research not only within wetlands & OBIA, but OBIA more broadly.

Classification of objects occurs under the same conceptual pathway of pixel-based classification, but the feature space is much larger as textural and spatial characteristics (proximity, size, shape) can all be incorporated into the classification algorithm.

Accuracy assessments for segmentation focus primarily on degree of over- or under-segmentation of the target reference units.

*Using segmentation to address heterogeneity and noise*

Speckle arises due to heterogeneity of intrinsic spectral variation, or mixed areas with fine-scale patchiness and diversity. The influence increases with high resolution data and may expose non-relevant landscapes.

OBIA can be a key means of reducing the speckle within the landscape by smoothing out local variation.

This may be particularly useful for radar based datasets, as interaction with vertical and horizontal structure of wetlands augments speckling. 

For most studies that were reviewed, "primitive objects" are the initial units of analysis, and objects at "higher" hierarchical levels can be recovered later from classification of the primitives.

*Detection and delineation of wetlands as landscape units*

OBIA can help in delineation by providing physically meaningful non-spectral attributes such as shape, size, spatial relationships and adjacency metrics.

OBIA may also smooth variation and help to delineate between "drier" vs. "wetter" regions within the landscape.

Supplementary datasets can also be incorporated into OBIA, such as digital terrain models, which may further help to delineate wetlands.

A major issue for delineating wetlands is the "fuzziness" of their boundaries, which are often unclear and may change with time. Potential strategies to reduce fuzziness may include:

  1. Examining imagery at dates with different hydrological patterns to obtain "generalized" boundaries
  2. Incorporating high-resolution imagery or ancillary data
  3. Focus on detection of wetland presence and prevalent extent rather than precise boundaries
  
*Hierarchical relationships in wetland classification*

Given complexity in cover types, may need step-wise classifications (i.e., classifications at multiple steps) or analyses at different spatial scales in order to improve accuracies.

In using hierarchical OBIA, researchers can examine "nestedness" within wetlands, which may allow for different scales of physical or ecological processes to emerge (biocomplexity).

*Diversity of object level variables for class delineation*

At the object level, there are a variety of different variables that can be employed for classification and subsequently merging of objects. Three primary types of information exist:

  1. Spectral variables - most commonly employed and may be spectral wavelengths, but may also be radar backscatter that indicates the height or "texture" of the landscape
  2. Texture variables - of non-spectral variables, texture-based ones are most commonly employed. Here texture refers to patterns within the object, and may be simple measures such as standard deviation or may become increasingly complex.
  3. Shape and contextual variables - shape, adjacency, proximity, perimeter to area ratios are all examples of informative metrics. Landscape objects that have a known shape may be able to be classified as a function of such parameters.

*OBIA for wetland change analysis*

Use of OBIA for change analysis is less common than single data segmentation or classification efforts, but may be valuable for investigating shifting parameters within particular regions of interest in a landscape.

The author notes the value of an "object fate analysis," in which investigation of spatially explicit changes in the shape geometry can be informative.

In employing wetland change analysis, magnitude of transitions should exceed the sizes of individual pixels or primitive objects.

*Accuracy in wetland OBIA*

Most studies employed the use of a traditional confusion matrix. Several studies used fuzzy accuracy assessment (?).

In general, the accuracies of the OBIA classifications increased with more than 4 classes, though studies rarely employed more than 10 classes.

The accuracy assessment brought to light important issues that may impinge upon accuracy of OBIA-based classification. In particular, the following may influence the classification accuracy:

  1. Spatial scale and spectral properties of image data and inputs to segmentation
    - object dimensions should significantly exceed scales of local noise and irrelevant heterogeneity
    - may be spectral, radar based, or supplementary in nature; timing of data acquisition is also critical 
  2. Choice of segmentation parameters for object generation
    - depending on segmentation parameters, different objects will be generated; additional research into standardization of them should be performed to identify "optimal" segmentation
  3. Choice of object attributes for classification
    - OBIA generates additional features and thus the selection amongst them can be particularly difficult; research into automation of optimizing the feature space is needed
  4. Choice of classification approach
    - different algorithms may perform better than others; more research needed here. SVM is quite good so far

*Summary points*

The author provides six key summary points:

  1. OBIA is effective for smoothing microvariation that often corresponds to "noise" and variance that is not ecologically meaningful; induces the unit of analysis to be more ecologically meaningful and effective for classifying into larger landscape classes.
  2. OBIA is effective for mapping isolated wetlands
  3. OBIA allows for hierarchical approaches - can delineate wetlands from other land types, and then begin to parse apart ecologically meaningful units within the wetland
  4. Flexibility in segmentation allows for both customization of analysis but may also hinder generalization and may induce biases into the analysis
  5. OBIA for change detection can be developed greatly and may be useful in examining the "fate" of particular landscape objects
  6. Wetland-specific challenges for RS inference remain in OBIA:
    - spectral similarities of diverse classes
    - subpixel mixing
    - difficulties in field sampling
