# Ecoinformatics

## Ecoinformatics

### Michener *et al*, 2001

*"Defining and unraveling biocomplexity"* [@michener2001defining]

**Key contribution:** This study's primary contribution to the literature is a formal definition of biocomplexity, and an investigation of its emergence within the scientific disciplines. The article primarily discusses:

  1. Defining biocomplexity
  2. Examining the characteristics of biocomplexity
  3. Speculating on the future of biocomplexity
  
*Biocomplexity* - "properties emerging from the interplay of behavioral, biological, chemical, physical, and social interactions that affect, sustain, or are modified by living organisms, including humans," as defined by [@michener2001defining]
  
**Key notes: **

  - Biocomplexity exists in contrast to the Popperian type of reductionalist science (causality & falsification), and recognizes that reduced causal mechanisms often are not robust within real-world situations
    - "myriad confounding factors operating at different temporal and spatial scales"
  - Two features of "biocomplexity"
    1. biocomplexity occurs as temporal, spatial and conceptual boundaries are broken-down and worked across
    2. emergent & unexpected properties may (or likely will) arise in response to biocomplexity

**Characteristics of "biocomplexity":**

  1. Relevant for a wide range of organisms; from microbial to megafauna (including humans)
  2. Relevant for a wide range of environments
  3. Reflected in nonlinear, chaotic, or unpredictable behaviors -- which likely necessitates sophisticated and innovative quantitative procedures
  4. Likely to span many hierarchical levels as well as across multiples scales in time and space
    
**Key research trajectories for ecology (as posed by Thompson 2001):**

  1. "Dynamics of coalescence in complex communities" - how do environmental constraints and species availability result in current complex ecological communities
  2. "Ecological memory" - how do past events & conditions influence current ecological structure and conditions
  3. "Emergent properties" - examining whether the first principles of "hard" sciences can predict composition, structure & function
  4. "Ecological topology" - identifying "rules" that occur over different scales of time and space, and examining how they interact to produce ecosystems

**Further reading:**

  1. [@thompson2001frontiers] - Frontiers of ecology: As ecological research enters a new era of collaboration, integration, and technological sophistication, four frontiers (listed above) seem paramount for understanding how biological and physical processes interact over multiple spatial and temporal scales to shape the Earth's biodiversity

### Jones *et al*, 2006

*"The new bioinformatics: Integrating ecological data from the gene to the biosphere"* [@jones2006new]

**Key contribution:** This paper makes the case for "locating, accessing and integrating" data within ecology. They provide examples of the successes associated with doing so in alternative fields (genetics), and make a case for the need within ecology. They also outline vertically integrated databases (data warehouses) versus metadata-driven databases (data collections), and the relative pros/cons of the two.

**Examples of metadata and data collections in ecology:**

```{r, echo=FALSE}
knitr::include_graphics(rep("images/jones2006AREES.bmp"))
```

**Key notes**: Some relevant points from the article:

  - A focus on "experimental" design and testing of multiple hypotheses makes ecological data inherently heterogeneous.
  - Ecoinformatics importance is housed in ability to "allow a broader perspective over time and space, and across many disciplines, than is possible from on or a few studies," and in allowing data to be reused for questions that may emerge in the future.
    - Would appear to be well-suited to address questions of biocomplexity
  - Current standard for storing of data is in heterogeneous spreadsheets
  - *Data warehouses (vertically-integrated)* - databases that coalesce data from many investigators, but focus on one theme and are searchable often through an online interface 
    - Often represents a "least-common denominator" approach in that they are less complex than original dataset
    - Examples: GenBank, VegBank or TreeBase
  - *Data collections (metadata-driven)* - collection of datasets that are loosely structured together by well-crafted metadata. Some advantages are:
      1. Data of disparate types can be housed within a single location
      2. Focuses on project-level data and thus is familiar to researchers
      3. Metadata often more detailed
      4. Metadata often more precise and thus can be used in search functions
  - *Metadata standards* - "consistent and rigorous sets of definitions for metadata categories that are broadly adopted by a research community." Several examples exist within ecology:
      1. Ecological Metadata Language (EML)
      2. Biological Data Profile (BDP)
      3. Geographic Markup Language (GML)
      4. Knowledge Network for Biocomplexity (KNB) - a metadata-drive repo that provides mappings from EML to BDP (but not from BDP to EML as of time of article publication)
  - *Data islands* - data derived from researchers working on disparate, autonomous topics that focus on specific, predetermined hypotheses
    - Often subject to entrenched domain-specific vocabularies and also assumptions; ontologies may be a way of resolving this limitation
  - *Ontologies* - well-defined concepts or terms that have precise meanings within a specific domain or field, as well as formal mappings of how the terms interrelate
    - *Science Environment for Ecological Knowledge (SEEK)* - one example of a system developing ontologies within ecology

**Further reading:**

  1. Look into metadata languages (e.g., EML)
  2. Look into some of the earlier work on databases by Michener

### Michener, 2006

*"Meta-information concepts for ecological data management"* [@michener2006meta]

**Definitions:**

  - *Metadata* - information about data, i.e. higher level descriptions of data
  - *Meta-information* - information derived from subsequent data processing and analyses; are these data products?

**Key contribution:** Michener outlines some of the key criteria, challenges and opportunities for metadata within an ecological context.

**Notes on metadata: **

Some benefits of metadata include:

  1. Longevity of the data is greatly increased
  2. Reuse of the data both by the original researcher and others is facilitated
  3. Well-documented data may be used to expand scale of ecological inquiry
  
The primary functionality of metadata is:

  1. Supports data discovery (most basic level)
  2. Facilitates the acquisition and use of data by humans (requires additional metadata)
  3. Enables automatic discovery, ingestion and analysis of data (requires more sophisticated metadata), which needs:
    - research context (hypothesis, study site, experimental design, etc.)
    - status of dataset
    - physical structure of dataset
    - info to facilitate use and comprehension
      
**Ecological Metadata Language (EML):**

  - EML is a standard for metadata that provides structure that supports automated functions of metadata listed above and is implemented in eXtensible Markup Language (XML)
  - Organized into modules, such as "dataset", "access", "physical", "party coverage", "project", "methods", "datatable", "attribute", and "physical"
    - Additional info at [ecoinformatics.org](http://ecoinformatics.org/)
  
### Bekker *et al*, 2007

*"Long term datasets: From descriptive to prescriptive data using ecoinformatics"* [@bekker2007long]

**Key contribution:** This is the lead article to a special issue in *Journal of Vegetation Science* that focuses on ecoinformatics driven approaches to understanding complexity within vegetation systems. The article is less informative in its own right as much as a good pointer to other key studies working with vegetation science and ecoinformatics. It appears to be one of the earliest applications of ecoinformatics to vegetation science.

**Further reading:**

  1. Le Duc, 2007 - A database application for long-term ecological field experiments
  2. Peacock, 2007 - The RAINFOR database, for monitoring forest biomass and dynamics
  

### Hale & Hollister, 2009

*"Beyond data management: How ecoinformatics can benefit environmental monitoring programs"* [@hale2009beyond]

**Key contribution: ** This is a less seminal study. They describe the relevance of ecoinformatics for environmental monitoring, but do not go much beyond some of the other pieces on ecoinformatics and its relevance for ecology. Some relevant points that stood out to me:

  - Standardization of taxonomic classes may be facilitated by ecoinformatics and application of ontologies in ecoinformatics
  - Establishing standards and protocols can help in synthesizing studies and results across time and space

**Note: ** Likely better to focus on other studies within the literature than this one as they don't bring up many points beyond those expressed in more seminal works.

### Feller *et al*, 2010

*"Biocomplexity in mangrove ecosystems"* [@feller2010biocomplexity]

**Key contribution:** Aim to elucidate the factors playing in at different hierarchical levels in mangrove ecosystems (biocomplexity). They use emergent properties (which they define as "patterns or processes that occur at multiple hierarchical levels within ecosystems") to examine underlying ecological processes within mangroves.

Biocomplexity appears to be thinking of ecology through a hierarchical lens, in which processes that exist at the "hard science" level (physics, chemistry, etc.) result in ecological processes > functions > services.

**Key findings:** The article doesn't differ much from other reviews of controls on mangrove processes, apart from the structure in which they examine:

  1. Controls of emergent properties (namely specializations for variable salinity, flooding and nutrient availability)
    - e.g., salinity tolerance, flooding tolerance, reproductive traits/regeneration, nutrient availability
    - perhaps analogous to "ecosystem processes"
  2. Collective properties of forests (e.g., zonation, productivity, nutrient cycling, food webs)
    - perhaps analogous to "functions"
  3. Emergent/collective properties of ecosystems & landscapes (e.g., habitat stability & connectivity of ecosystems)
    - perhaps analogous to "services" or "benefits"

*Emergent properties* - concept that the whole is greater than the sum of its parts when examining ecological phenomena and the underlying processes that drive them. Emergent properties can exist at any scale, and may build upon themselves (e.g., morphology and physico-chemical conditions leading to ecophysiology (emergent property) which then leads to spatial assemblages (2^nd^ emergent property), etc.). It may be helpful to think of different emergent properties at different scales of time and space.

Three primary characteristics of emergent properties exist:

  1. Emergent properties are not simply the sum of their parts
  2. Emergent properties exist as a type that is different from their underlying components
  3. They cannot be easily predicted from individual components

Examples of emergent properties within mangroves are ecophysiological traits (as listed above) that emerge from the interaction between biochemical and morphological traits with environmental factors.

Individual Based Models (IBMs) are discussed as useful though limited tools for examining biocomplexity within mangroves, but note that there is relatively little work in this field to date. See section \@ref(chentwilley1998) for an example of using IBMs within mangroves.

```{r, echo=FALSE}
knitr::include_graphics(rep("images/feller2010ARMS.bmp"))
```

**Relevance to research:**

  1. Are IBMs relevant for research questions at hand?


### Michener & Jones, 2012

*"Ecoinformatics: Supporting ecology as a data intensive science"* [@michener2012ecoinformatics]

**Key contribution: ** This is a seminal study on the relevance of ecoinformatics written by two of the leading thinkers in the field.

**Key notes: ** Several key notes from the study are:

  - The authors describe the "data life cycle,"" with 8 steps that do not necessarily occur in sequential order and depend upon the analysis and research question at hand.
  - The steps are:
    1. *Plan* - derive a data management plan that is often a living document as needs and uses for data will change over time
    2. *Collect* - there are highly variable mechanisms under which this occur given the vast amount of data types and collecting methods that exist; standardization in collection procedures is key for synthesizing data at later times
    3. *Assure* - measures taken to ensure data quality both before and after collection of the data; may consist of standardization of codes, units or metadata (e.g., using EML)
    4. *Describe* - sufficiently describing the data via metadata (who, what, when, where, why)
    5. *Preserve* - depositing of data in a repository such that they may be verified, maintained and accessed through time
    6. *Discover* - identifying appropriate data for research question of interest; double-edged sword in that you can get millions of hits or alternatively have tons of distinct data sources that exist only on disparate, local machines. DataONE is one such tool that drills down into data to identify datasets of actual interest; this is where ontologies come into play
    7. *Integrate* - understanding methodological differences, transforming to common representations and converting and recoding data to comparable semantics
    8. *Analyze* - reproducibility in analysis, with a particular focus on scientific workflows

*DataONE:* DataONE is a research platform that seeks to support the entire data life cycle, as well as integrate many different forms of data. There are three principle components:

  1. Member Nodes - contribute data, computing resources and services (e.g., data repositories, universities, computing centers)
  2. Coordinating Nodes - support network-wide services such as indexing, replication, and interactability of Member Nodes
  3. Investigator Toolkit - tools for researchers to employ in all steps of the data life cycle
    
**Schematic of the data life cycle:**

```{r, echo=FALSE}
knitr::include_graphics(rep("images/michenerjones2012.bmp"))
```

### Soranno, 2014 {#sorrano2014}

*"Cross-scale interactions: Quantifying multi-scaled cause -- effect relationships in macrosystems"* [@soranno2014cross]

**Key significance:** This study calls for a need for more explicit investigation of cross scale interactions (CSIs) both in time and space. They lay out a framework for investigating CSIs that consists of i) conceptual mapping, ii) construction of an appropriate database (ecoinformatics), and iii) modeling strategies that specifically investigate interactions (e.g., interaction terms in hierarchical modeling).

**Notes:** 

The authors define two primary types of *cross scale interactions* (CSIs).

  1. Interactions among driver variables at different spatial scales that influence a focal resonse variable
  2. Interactions between transport processes that link fine- and broad-scaled processes
  
The authors explicitly model the interaction using Bayesian hierarchical models via an interaction term between regional and local driver variables.

*Steps for quantifying CSIs*:

  1. Conceptual model of cross-scale relationships that explicitly considers interactions and interdependencies. Development of the conceptual model is key to the process and entails four key components:
    1. Identifying relevant spatial and temporal extents and resolutions of driver variables
    2. Considering the types of the interactions that are expected
    3. Determining whether relationships are linear or nonlinear
    4. Identifying critical assumptions and uncertainties that may generate additional testable hypotheses
  2. Compilation of multi-scaled, multi-themed databases
    - Summarize independent variables at multiple scales
    - Resolve differences in data resolution
    - This is where ecoinformatics is worked in
  3. Test for CSI as interaction term using Bayesian hierarchical models
  
The authors exemplify these steps in examining P nutrient additions in lake ecosystems in the US. Specifically, they encounter the issue in that wetland cover surrounding lakes leads to *both* increases and decreases in concentrations of P. They hypothesize that there are interactive effects between agricultural presence and wetland presence on lake P, and model the cross spatial scale interactivity between the two as an interaction term.

The results are significant and lead the authors to additional hypotheses. The particular case study is facilitated by standardization of lake environmental quality data.

### Ruegg, 2014 {#ruegg2014}

*"Completing the data life cycle: Using information management in macrosystems ecology research"* [@ruegg2014completing]

**Key contribution:** This article promotes the premise that macrosystem ecology requires integrated information management at its inception. In calling for "completing the data cycle," they refer to well-documenting datasets such that data may be re-used to answer new and different questions from those that are conceived by those involved in the original projects.

The article is largely an extension of Michener and Jones (2012), and argues largely for extensive documentation of datasets to enhance longevity.

**Notes: ** The study notes several key areas within the data life cycle:

  1. Data collection: Incorporating information management early within the data collection process; critical to consider the structure and longevity of the dataset prior to collection such that it may be built upon and reused in the future
  2. Data integration (1): Integrating data from many sources. Currently integration is not possible through an automated manner, though tools exist to help in data integration
  3. Data integration (2): Integrating data from local observations to models and back requires unique considerations. Must understand spatial resolution of different datasets (e.g., field and RS data). Spatial and temporal issues exist for interpolation of field (e.g., point) datasets.
  4. "Completing the data cycle": Documenting well and sharing facilitate analysis. Tools that show full data provenance and work flow can be effective in showing the full analysis and collection of a particular dataset.

## Ontologies

### Williams, 2006 {#williams2006}

*"Ontologies for ecoinformatics"* [@williams2006ontologies]

**Key contribution:** This is an early piece that gives an overview of what purpose ontologies may serve within ecology. They discuss it in particular relation to food webs within ecology, which are inherently complex with multiple definitions or system-specific interpretations for terms.

**Key notes: ** Ontologies should be employed to describe the following characteristics of a dataset:

  1. Where, when and by who the data are collected
  2. A description of what was observed, including identity and traits of the organism
  3. Sampling protocol, including collection procedures and associated experimental manipulations
  
The architecture of the ontology can be designed intelligently such that the ontology is extensible, and allows for both system-specific interpretations as well as across-system relationships. For example, they discuss the importance of separating "entities" and "traits".

The piece by Madin, 2008 (see Section \@ref(madin2008)) provides a more comprehensive overview of ontologies within ecology and should be referred to in conjunction with this piece.

### Madin, 2007 {#madin2007}

*"An ontology for describing and synthesizing ecological observation data"* [@madin2007ontology]

**Key significance:** Madin *et al* present a formal ontological framework within ecology (SEEK's OBOE) that captures the essential semantic information of observational datasets, which will aid discovery and integration of data. The paper is structured:

  1. A discussion of ontologies and their relevance in ecology
  2. An introduction and overview of SEEK Extensible Observation Ontologies (OBOE)
  3. Several examples of use of OBOE for i) data discovery, ii) data summary, and iii) data integration

**Notes:** 

*Ontology* - "A representation of the knowledge within a domain of interest, defined via the concepts within the domain as well as the properties and relationships among domain objects"

The work of the authors of the study is founded within the SEEK project. SEEK aims to develop technology to discover, access, integrate and analyze distributed ecological information, primarily through extension of EML to support semantic annotation of ecological datasets. This is done primarily through OWL-DL (description logic) ontologies.

*Extensible Observation Ontology (OBOE)* - aims to provide a core ontology framework for semantically annotating observational data sets. OBOE differs from other ecological ontologies in that it:

  1. Provides a robust framework for describing generic scientific observations
  2. Facilitates and structures the creation and sharing of domain-specific ontology extensions
  3. Provides discovery and integration services by semantic annotations to the ontology across variable domains of ecological observation data

The core structure of OBOE is diagrammed below:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/madin2007.bmp"))
```

The five key characteristics of OBOE are:

  1. *Observations* - formal statement that an entity of a particular type was observed
  2. *Entities* - all *Observations* are composed of exactly one *Entity*
  3. *Measurements* - *Observations* can be composed of *Measurements*, which represent measurable *Characteristics* of whatever entity is being observed
    - *Measurements* are always associated with *Observations*
    - Can have a variety of different characteristics, which are defined via the *ofCharacteristic* property
  4. *Characteristics* - "qualities" of the entity being observed
  5. *Measurement standards* - the units, scales, categories, catalogs and lists used when measuring characteristics. Two subclasses of *Measurement standards* exist:
    1. *Entity naming standard* - defines uniques names for unique identifies (e.g., "California")
    2. *Entity classification standard* - entities associated with a classification have one or more instances (e.g., "red", "5 m")
    - Additionally, for ratio or other complex units (e.g., ct / ha), the individual components may be assigned *Measurement standards* such that automatic conversions or data discovery may be performed.

The different classes of the above can be used to describe the data, and facilitate machine-intelligent discovery, data summary and data integration. For example, data searches for density data may be able to identify datasets consisting of explicitly reported "density" data, as well as datasets that report the constituents of "density" (e.g., count & area)

### Madin, 2008 {#madin2008}

*"Advancing ecological research with ontologies"* [@madin2008advancing]

**Key significance:** This article gives an overview of ontologies in ecology, with particular focus on:

  1. What issue they are capable of resolving -- i.e., integration of heterogeneous data
  2. What ontologies are and the forms they take
  3. Ongoing efforts to advance ontologies within ecology

**Ontologies** provide a formal mechanism for defining terms and their relationships, and can improve the location, interpretation and integration of data based on its inherent meaning."

**The issue:** Lack of critical consideration of "concepts" within ecology complicates interpretation and synthesis of findings, as well as the location and reuse of existing datasets.

**The solution:** Ontologies provide a formal framework for establishing concepts within a discipline, and can resolve some of the issues that arise from multiple or nuanced interpretations of the same "concept" within ecology.

**Notes: ** Ontologies within ecology are becoming more common, with three particular trends emerging:

  1. *Domain-specific ontologies* that focus on capturing terminology used in specialized scientific disciplines or communities, for example:
      1. Ontologies in Web Ontology Language (OWL) for food-web networks that are described in Williams 2006 (see Section \@ref(williams2006)).
      2. *Semantic Web for Earth and Environmental Terminology (SWEET)* - Maintained by NASA JPL, use OWL ontologies to improve data integration and access capabilities.
  2. *Framework ontologies* that define general concepts and relationships that others can extend when building domain ontologies, for example:
      1. *Extensible Observation Ontology (OBOE)* -  component of the SEEK project; basic concepts and relationships for describing observational datasets (e.g., field, experimental, simulation and monitoring data). Compatible with EML, and its core concepts (i.e., "Entity", "Characteristic", and "Measurement Standard") can all be extended with domain-specific ontologies.
      2. *Classes for Environmental Data Exchange (CEDEX)* - formal framework for describing monitoring and experimental data from the European consortium of long-term ecological research stations.
      3. *Observations Data Model (ODM)* or the *Open Geospatial Consortium's Observations and Measurements model OGC O&M* - framework intended to facilitate the sharing and reuse of data within particular domains (hydrology projects for the first, geospatial work for the second); free-form extensibility of OBOE is not present.
  3. Other less formal approaches that identify and describe terms and concepts that are relevant to ecological research
    - These commonly exist as thesauri that seek to standardize use of terms (and thus may aid in searching or browsing), but are not as effective as formal ontologies due to informal (and therefore a lack of logical relationships) nature.

The authors provide a nice figure that exemplifies the architectural framework of responsible management and archiving of data:

```{r, echo=FALSE}
knitr::include_graphics(rep("images/madin2008.bmp"))
```

### Reichman, 2011 {#reichman2011}

*"Challenges and opportunities of open data in ecology"* [@reichman2011challenges]

**Key significance:** This piece focuses on how ecology (in some respects) has moved from a site-scale experimental field to one that is capable of large synthesis studies across broad spatial or temporal scales. They discuss issues of heterogeneity in data format and availability, and how to resolve them in order to perform synthetic analyses.

**Key notes: **

Reichman *et al* estimate that just 1% of ecological data that has been produced is available after publication of the associated results.

They note three major technological challenges to "open data" (a la, Michiner, 2006), namely:

  1. Data dispersion - Decentralized existence across a broad range of databases (either local or online)
  2. Heterogeneity - Lack of consistency in collection, formatting and reporting
  3. Provenance - Issues of origin and history (relevant for datasets, as well as data products)
  
*Data dispersion:*
  
  - *DataONE* - A project designed to provide access to a number of central data repositories (e.g., Data Dryad) that facilitates data searching and access (also see *Global Earth Observation System of Systems (GEOSS)*)

*Heterogeneity:*

  - Data heterogeneity unlikely to be resolved due to specific research objectives and logistical constraints. Use of structured metadata such as EML, however, is efficient at aiding access and reporting of heterogeneous data.

*Data provenance:*

  - Need scripted workflows that show quality control, processing and analysis of a dataset, particularly for derived datasets (e.g., meta-analyses or other secondary data objects)
  - Unique identifiers (as DataONE) provides can be key for monitoring data provenance
  
*Social considerations:* Social constraints to formalized sharing and archiving data are likely greater than the technical challenges. A formal reward system for promoting open access data may be key to advancing open data within ecology.

### Wickham, 2014

*"Tidy data"* [@wickham2014tidy]

**Key significance:** A formal description of Hadley's "tidy data" concept. See also this online [article](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html), which is a bit more of a code-based explanation of tidy data.

**Key notes: ** Tidy data has the following fundamental format: each variable is a column, each observation is a row, and each type of observational unit is a table.

*Data tidying* - as organizing datasets such to facilitate analysis

In organizing data, there is a general rule of thumb that it's easier to describe functional relationships between variables (i.e., linear combinations of columns) rather than relationships between observations (rows); AND it's easier to compare across groups of observations (rows) relative to between groups of variables (columns)

There are four basic verbs for data manipulation:

  1. *Filter* - subsetting or removing observations based on a condition
  2. *Transform* - adding or modifying variables
  3. *Aggregate* - collapsing multiple values into a single value
  4. *Sort* - changing the order of observations
  
Tidy data is particularly useful for modeling, and is a large part of why it is designed/promoted.


## Ecosystem stability

### Scheffer, 2001 {#scheffer2001}

*"Catastrophic shifts in ecosystems"* [@scheffer2001catastrophic]

**Key contribution: ** This article focuses on observed large-scale shifts in major ecosystems and their explanations. They provide both theoretical framework as well as overview of results from different ecosystems and highlight different patterns as well as implications for management.

**Key notes: **

*Theoretical framework*

Catastrophic shifts occur abruptly, and thus "early-warning signals" of approaching catastrophic change are difficult to gather.

"Hysteresis" - memory of the system; the path back to a pre-existing state is not simply achieved through reversing the changes that brought about the shifts in states. 

"Resilience" - as defined by Holling is the size of the valley, or basin of attraction around a state, which corresponds to maximum perturbation that can be taken without causing a shift to an alt stable state.

Gradually shifting environmental parameters may not induce a change in state, but may greatly impact the "resilience" of a particular basin of attraction.

*Examples*

Lakes - One of the most common examples is loss of water clarity and vegetation due to eutrophication in shallow lakes. A threshold appears to exist in which below that, nutrients have no effect but above it they induce stark state shifts.

Hysteresis can be exemplified here as reduction in nutrient concentrations is not sufficient to reduce vegetation and return clarity to lake water.

Desertification - Presence of desert vs vegetated landscapes may be conceived of as alt stable states. May occur through time as vegetation promotes more moist conditions and lack of vegetation promotes drier conditions. May have feedbacks with the climate in the Sahel region that produce alternating stable states.

*Emerging patterns*

Strongest cast for extistence of alt stable states is based on combinations of approaches, e.g., observations of repeated shifts, feedback mechanisms that maintain different states, or models showing mechanisms plausibly explain field data.

Contrast in ecosystem states commonly seen by shift in dominance of organisms with different life forms.

State shifts often induced by stochastic events.

Feedbacks that stabilize states involve biological, physical, and chemical mechanisms.

Critically, invisible impacts, for example anthropogenic effects, may not induce state shifts but may greatly impact the resilience of particular alternative stable states.

*Management implications*

Resilience often depends on variables that change slowly over time, and thus may be more easily "managed". Sharp perturbations are less predictable and thus are more difficult to manage. Thus, maintaining resilience is likely the most pragmatic means of managing ecosystems.


### Beisner *et al*, 2003  {#beisner2003}

*"Alternative stable states in ecology"* [@beisner2003alternative]

**Key significance: ** This article provides a nice review of the theory surrounding alternative stable states as they relate to ecology.

**Key notes: **

The idea is credited first to Lewontin, 1969, but two other seminal works are Holling, 1973 and May, 1977.

Theoretical ecologists largely view alternative stable states and the pressures on the resilience of systems in two different conceptual ways:

  1. Different states exist simultaneously under same conditions and that shifts between the two states is due to a sufficiently large perturbation (shifts in variables)
  2. Change in the parameters that determine behavior of state variables and the ways they interact with each other (shifts in parameters)
    - generally occur due to shifts in environmental "drivers" influencing communities
    
*Variable shifts*
    
Perturbations to state variables commonly thought of via two separate "classes" of alternative stable states:

  1. Alternative interior states - involves nonlinearity in the transformation of the state such that multiple stable points exist within the system (local stability does not imply global stability)
  2. Incorporates boundary states (one or more species may be absent) - if systems of equations are linear, only one stable composition is possible with all species represented... but may have other stable points in which some of the species are missing.
    
*Shifts in parameters*

Parameters are defined as "environmental drivers".

Each state considered stable but associated dynamics (local stability, pop fluctuations) are different given that it corresponds to different parameter values.

Given that the "landscape" changes, not all potential alternative stable states exist may be present at all times.

*Resilience* - relates to characteristics of basing that acts to retain the community

*Hysteresis* - the dependence of the state of a system on its history (i.e., memory); "hysteresis is revealed if the return trajectory of the equilibrium point differs from that adopted during its outward journey" <- here the outward journey is influencing the system (i.e., it "remembers" its return) and thus it exhibits a different return trajectory.



### Biggs, 2009 {#riggs2009}

*"Turning back from the brink: Detecting an impending regime shift in time to avert it"* [@biggs2009turning]

**Key contribution: ** This article examines the "timing" of an early warning signal. It is motivated under the premise that EWS may occur as a system approaches a bifurcation point, but it is unclear whether or not effective management action may be taken in time to avert a regime shift. 

They explore:

  1. How close ecosystems can get to ecological thresholds and still avert regime shifts
  2. Which regime shift indicators might give warning before "no point of return"

**Key notes: **

The underlying drivers that reduce resilience in systems are often associated with economic gains.

The findings indicate that if the pressure can be immediately reduced, then action even during beginning stages of regime shift might be effective enough to avert actual shift. For pressures that take much longer to reduce, then action must be taken considerably before the regime shift begins to occur.

EWS such as increases in variance, skewness, kurtosis and AR1 coefficient occur during "policy window" (i.e., the time in which action is still effective) for fast-actions, but only begin to occur after the policy window for slow actions has already closed.

Regime shifts that occur through slow-change of underlying variables occur via bifurcations: small gradual changes can cause abrupt qualitative changes in long-term behavior due to appearance or disappearance of attractors.

Key finding is that within environmental management, there are key windows in which action must be taken to avert shifts. The authors note that this window and the efficacy of different policies over time are rarely taken into consideration.

In managing ecosystems, likely have variables that can be altered both quickly as well as those that change more slowly with time. Thus, may be able to "buy time" in altering the more reactive variables while progress is made on the slower-moving variables.

Their findings show that some of the EWS indicators from Scheffer, 2009 might not be effective from a time-perspective. They use a spectral density ratio, which can be expressed in absolute terms and may be effective in identifying switches in attractors.



### Scheffer, 2009 {#scheffer2009}

*"Early-warning signals for critical transitions"* [@scheffer2009early]

**Key contribution: ** This paper follows up on Scheffer 2001 in that it addresses the issue of actually identifying early warning signals. The key issue here is predicting critical transitions, which is no easy task, may be facilitated by certain generic symptoms that occur as a system approaches a critical point.

**Key notes: ** 

*Bifurcations* 

Catastrophic bifurcations - once a threshold is exceeded, a positive feedback propels system through phase of directional change towards contrasting state.

Other important bifrucations include those that mark transitions from stable equilibrium to cyclic or chaotic attractor.

*Theory*

"Critical slowing down and its symptoms" - recovery rate of the system as it approaches a critical bifurcation can be used as an indicator of how close the system is to said bifurcation point. The *rate* of change is the criteria (approaches 0 with increasing proximity to critical bifurcation), and thus there is not risk of pushing a system into an alt stable state. Key symptoms are:

  1. Slowing down leads to increase in autocorrelation in resulting pattern of fluctuations
  2. Slowing down causes intrinsic rates of change to decrease, thus state of system increasingly resembles previous state (i.e., memory of system; can be seen through lag-1 autocorrelation)
  3. Increased variance in pattern of fluctuations; given lagged response time, the effects of shocks become additive, increasing variance.

Can use presence of natural perturbations to examine recovery rate.

"Skewness and flickering":

  - Skewness - asymmetry of fluctuations; characterized by the state existing in the unstable side of a bifurcation point longer than the stable side (refers to directionality of approach to bifurcation point)
  - Flickering - occurs when stochastic forcing pushes the state back and forth between the basins of attraction; considered early warning because system may permanently shift to alternative stable state

"Cyclic and chaotic systems" - may have transitions from stable systems to oscillatory systems; signaled by critical slowing down characterized by long transient oscillations

"Spatial patterns as warning signals" - may have increased cross-correlation or spatial coherence among units before a critical event (e.g., presence of species in adjacent patch habitats)

*Examples*

Climate is one of most commonly examined as a messy "real-world" system. Analyses indicate that EWS may exist for abrupt shifts in climate (from tropical to ice-age). For example, increases in autocorrelation have been found, as well as flicking prior to Younger Dryas cold period.

### Boettiger, 2012 {#boettiger2012}

*"Quantifying limits to detection of early warning for critical transitions"* [@boettiger2012quantifying]

**Key significance: ** This paper discusses the uncertainty and failures of simple statistical means of EWS in alternative stable state theory. They make the case for a model-based approach, which can help alleviate many of the shortcoming of a purely statistical approach.

**Key notes: **

Summary statistics approach commonly employs increasing trend in summary stats such as variance, autocorrelation, skew, or spectral ratio.

*Shortcomings of summary stats approaches*

There are many hidden assumptions that occur with use of summary statistics. Variance may increase for reasons that do not signal approaching transition.

Approaches based on increasing trends in summary stats assume a changing parameter has brough system closer to a bifurcation. This assumption excludes at least three alternative explanations:

  1. A large perturbation of the system state has moved system to alternative basin of attraction
    - this perturbation is exogenous to system and thus would not be expected to be predicted
  2. Purely noise-induced transition, chance fluctuation that happens to carry system across the boundary
    - studies have shown it's unlikely EWS are able to detect these
  3. System may pass through saddle-node bifurcation, but in rapid or highly nonlinear way making detection of gradual trends impossible.
  
Given that we do not have replicate systems in which to examine EWS or regime shifts, often use shifting window applied to single replicate over time. This is an issue as selection of the window size and whether or how much to overlap are arbitrary decsisions.

Quantitative measures of early warning patterns do not exist.

*Model-based approach*

Read once through, but confused by math... need to go back and read a second time.


### Moffett, 2015 {#moffett2015}

*"Multiple stable states and catastrophic shifts in coastal wetlands: Progress, challenges, and opportunities in validating theory using remote sensing and other methods"* [@moffett2015multiple]

**Key contribution: ** This provides an overview of using remote sensing to examine different alternative stable states in coastal ecosystems. The article is potentially worth a second read, particular the examples section in which they provide some details on systems outside of mangroves.

**Key notes: **

*Applying multiple stable state theory*


"Multiple stable ecosystem states are created by coupled positive and negative feedbacks that link the ecological, hydrological and geomorphological processes in the coastal wetland environment in a self-reinforcing and relatively resilient manner"

*Alt stable states in observation and management*

Several characteristics of alternative stable states make them difficult to observe:

  1. Impossible to uniquely define state of system with single observation points in time, but rather need knowledge of history of the system as well
  2. Practical issue of scale in a system with potential alternative stable states; thus, need observations that span both time and space
  
Alternative stable states should give rise to two important and related phenomena:

  1. Tendency for contrasting states to persist simultaneously along sharp boundary despite gradual change in underlying environmental conditions ("self-organized ecosystem patterning")
    - May be induced by scale-dependent feedbacks
  2. Ability of system to rapidly and wholly switch from one broad ecosystem state to another ("catastrophic ecosystem shift")

*Alternative stable states in mangroves*

"Runaway sedimentation" - One mechanism of forming alternative stable states in mangrove systems; as mangroves expand, they induce sedimentation that may eventually limit their expansion, and thus may reach a new stable state wherein the shoreline is no longer prograding

Notion of alternative stable states may differ depending upon scale:

  - At ecosystem scale may be cosnidered as mangroves vs. other coastal vegetated ecosystems
  - Alternatively may also consider patch occurrence of different mangrove species or zonation
  
## Decision-making

### Fischer, 2009

### Chapin, 2010

### Folke, 2010

### Polasky *et al*, 2011 {#polasky2011}

*"Decision-making under great uncertainty: Environmental management in an era of global change"* [@polasky2011decision]

**Key contribution**: This article discusses the difficulty of environmental management under the pervasive uncertainties that exist due to global change. The authors review several strategies for informed decision-making that may help to alleviate challenges due to uncertainty.

**Key notes: ** 

The authors note that it is important to consider the potential for learning from decisions in addition to the potential impacts that may happen. Decisions that are made should be made such that they will inform iterative management as addtional understanding is achieved.

In the authors view, ideal decision-making approaches should clarify the effect that alternative decisions have on probable desirability of outcomes in terms of stated objectives.

*Decision theory*: 

"Decision theory" requires information about probabilities of various outcomes under alternative management options, as well as the desirability of those outcomes.

In standard decision theory:

  1. Uncertainty is represented by assuming many different states of the system, as well as the various probabilities that each of those outcomes will occur
  2. Outcomes are joint products of action and state, which produces a conditional probability of outcomes given the action
  3. The outcomes are expressed in common metrics, and general best "decision" is the one that maximizes the "utility" of an action
  
Advantage of standard decision theory is that it provides a clear and repeatable framework for generating a course of action that is based on empirical data.

Disadvantage is that the necessary data (i.e., all alternative states) is unlikely to exist.

Incorporating learning into the decision-making process may be one means of mitigating the lack of data: "forward-looking decision makers should take account of how current decisions might influence future conditions, future decisions and the probable impacts of decisions on current and future well-being."

Idea that "decision-making" should be treated as an experiment.

**Alternative methods to use beside decision theory:**

*Thresholds approach* - focus attention on critical boundaries that may lead to persistence of a new alternative state (often times with major consequences)

Common example is the 2 degrees celsius threshold thought to be catastrophic GHG emissions.

Thresholds are rarely ever precisely known, and thus each action incorporates a certain degree of risk with it. Decision-making ultimately needs to decide on what levels of risk are acceptable.

In this context -- those in power are ultimately deciding that the risks currently seen by climate change are "fine".

*Scenario planning* - thinking creatively and systematically about complex futures.

Scenario planning does not incoporate probabilities, but rather models and explores potential alternative outcomes supported by data and simulations. Scenario planning can be helpful in conceptualizing the future.

The key weakness of scenario planning is it is difficult to assess the likelihood of any given scenario occurring.

*Resilience thinking* - focuses on critical thresholds for system performance, as well as the capacities to both adapt and shift to a new system should a former one become untenable.

Employs methods that use multiple perspectives to better-understand and identify uncertainty (similar to scenario planning).

Resilience thinking requires an iterative approach, as assumptions and actions are re-evaluated with increasing availability of information (i.e., "adaptation").

*Decision making for global change*

Standard decision making theories will fail as likelihoods of alternative states are likely to be too subjective and will induce disagreement such that decisions are ultimately not taken. The alternative methods proposed in the study may be one route by which decision-making can be informed, without the necessity of assigning probabilities to particular outcomes or actions.

Decision-making around global change may be thought of as two phases with continuous feedbacks:

  1. Scoping the problem as broadly as possible to expand imaginable states and outcomes
    - "Scenario planning" and "resilience thinking" can be effective here
  2. Actually making decisions given current understanding
    - "Threshold approach" given that science often more robust in identifying vulnerabilities than predicting future
    

## Mangrove thresholds

### Cavanaugh, 2014

### Eslami-Andergoli, 2015

### Ellison, 2015

### Sasmito, 2016
